---
title: "SOLUTIONS: Building a simple linear regression table"
subtitle: "Depression among adolescents in Jimma, Ethiopia"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

These are the completed solutions for the week 7 exercise. Each step first shows the blanked chunk from the exercise (not run), then the filled-in **Solution** chunk that runs. In Part 2 we illustrate one possible choice of two extra predictors (place of residence and Oslo social support); any two of the three remaining predictors are acceptable.

In this exercise we build the **simple linear regression** column of a published table, using the study's own data. We fit the first few rows together, then add more predictors, comparing our numbers to the paper and finding two places where the published table does not match a fresh analysis of the data.

## The study

- Girma et al. (2021), *Depression and its determinants among adolescents in Jimma town, Southwest Ethiopia*, PLOS ONE. A copy is in `paper/depression_paper.pdf`.
- The survey data the authors released is in `data/depression_data.xls`.

## The outcome: the PHQ-9 depression score

The authors measured depression with the **PHQ-9 (Patient Health Questionnaire-9)**, a widely used screening tool. It asks how often, over the last two weeks, a person has been bothered by nine problems, for example *"Little interest or pleasure in doing things"* and *"Feeling down, depressed, or hopeless."* Each item is rated 0 ("Not at all") to 3 ("Nearly every day"). The nine items are summed into a total from 0 to 27 (higher = more symptoms), which we rename to `depression_score`.

# PART 1: Build and check the first rows

## Load packages

Use `p_load()` to load: readxl, tidyverse, here, janitor, broom, gt.

*Exercise chunk (blank):*

```{r load-packages-q, eval=FALSE}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

**Solution:**

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(readxl, tidyverse, here, janitor, broom, gt)
```

We also load the GRAPH Courses fork of the inspectdf package, which is currently unavailable on CRAN.

```{r load-inspectdf}
pacman::p_load_gh("the-graph-courses/inspectdf")
```

## Step 1: Import, clean names, and label every predictor

Import the data and `glimpse()` it to see what we are working with.

```{r data-import}
survey_raw <- read_excel(here("data/depression_data.xls"))
```

*Exercise chunk (blank):*

```{r glimpse-raw-q, eval=FALSE}
________(survey_raw)
```

**Solution:**

```{r glimpse-raw}
glimpse(survey_raw)
```

The raw column names are a little messy. We clean them, then subset to just the columns we will use.

```{r clean-rename}
survey <- survey_raw %>%
  clean_names() %>%
  rename(
    depression_score = scoredep,
    chronological_age = agechron,
    oslo_social_support = osl3,
    sex = sexrespo,
    school_ownership = schoolty,
    residence = residece,
    grade_level = gradelev
  ) %>%
  select(
    depression_score,
    chronological_age, oslo_social_support,
    sex, school_ownership, residence, grade_level
  )

glimpse(survey)
```

**Question**: Based on the study title and the list of variables above, which is the outcome variable in our dataset?

**Answer:** `depression_score` is the outcome. It is the PHQ-9 total the study set out to explain; the other variables are the potential determinants (predictors).

Let's explore the cleaned data with `inspect_cat()` and `inspect_num()`.

```{r eda-raw}
inspect_cat(survey) %>% show_plot() # no categorical variables at the moment
inspect_num(survey) %>% show_plot()
```

Four of these predictors are categorical but stored as numeric codes, so we label them now, ordering the levels according to the paper.

*Exercise chunk (blanks):*

```{r label-q, eval=FALSE}
survey <- survey %>%
  mutate(
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Female", "Male")
    ),
    school_ownership = factor(
      school_ownership,
      levels = c(0, 1),
      labels = ______
    ),
    residence = factor(
      residence,
      levels = c(0, 1),
      labels = ______
    ),
    grade_level = factor(
      grade_level,
      levels = c(1, 2, 3, 4),
      labels = ______
    )
  )
```

**Solution:**

```{r label}
survey <- survey %>%
  mutate(
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Female", "Male")
    ),
    school_ownership = factor(
      school_ownership,
      levels = c(0, 1),
      labels = c("Private", "Public")
    ),
    residence = factor(
      residence,
      levels = c(0, 1),
      labels = c("Urban", "Rural")
    ),
    grade_level = factor(
      grade_level,
      levels = c(1, 2, 3, 4),
      labels = c("Grade 9", "Grade 10", "Grade 11", "Grade 12")
    )
  )
```

Now we can check again: the four labelled variables show up as categorical.

```{r eda-labeled}
inspect_cat(survey) %>% show_plot()
inspect_num(survey) %>% show_plot()
```

**Question**: Which predictors are continuous, binary, or polytomous?

**Answer:**

- Continuous: `chronological_age`, `oslo_social_support`.
- Binary: `sex`, `school_ownership`, `residence`.
- Polytomous (3+ groups): `grade_level` (four grades).

## Step 2: Run one simple linear regression per predictor

*Exercise chunk (blanks):*

```{r models-q, eval=FALSE}
age_model   <- lm(depression_score ~ chronological_age, data = survey)
sex_model   <- lm(___________ ~ ______, data = __________)
grade_model <- lm(___________ ~ ______, data = __________)
```

**Solution:**

```{r models}
age_model   <- lm(depression_score ~ chronological_age, data = survey)
sex_model   <- lm(depression_score ~ sex, data = survey)
grade_model <- lm(depression_score ~ grade_level, data = survey)
```

```{r sex-summary}
summary(sex_model)
```

**Question**: In `summary(sex_model)`, which sex is the reference group? Fill in the blanks in the sentence below to explain the relationship between sex and depression score.

**Answer:** Female is the reference group (the first factor level, folded into the intercept).

> The reference group is **Female**. The `sexMale` coefficient is about **-0.66**, which means that on average, males score about **0.66** points **lower** on the PHQ-9 than females.

(The coefficient keeps its minus sign, -0.66; the "points lower" phrase uses its size, 0.66, plus the direction word "lower".)

## Step 3: Extract the beta and 95% CI with `broom`

```{r tidy-example}
tidy(sex_model, conf.int = TRUE)
tidy(grade_model, conf.int = TRUE)
```

**Question**: What is the 95% CI for the `sexMale` coefficient? Fill in the blanks in the sentence below to interpret it.

**Answer:**

> The 95% CI for the `sexMale` coefficient (the male − female difference) runs from about **-1.35** to **0.03**. Because this interval **includes** 0, the difference **is not** statistically significant at the 0.05 level.

**Question**: The slope estimate for Grade 12 is 1.64. Fill in the blanks in the sentence below to explain the relationship between grade level and depression score.

**Answer:**

> Compared with **Grade 9** (the reference group), students in Grade 12 score about **1.64** points **higher** on the PHQ-9, on average.

### Helper for categorical predictors

Now, to turn these regression results into a paper-style table, we create some helper functions. Most of the formatting is written for you; as before, focus on getting the statistics right and check the values against the model output.

`cat_rows()` handles a categorical predictor: it drops the intercept, strips the variable name off `"sexMale"` to leave `"Male"`, and adds back a **reference row** with β = 0.

```{r helper-cat}
cat_rows <- function(model, var_name, var_label, ref_label) {
  # non-reference rows: drop the intercept, clean the level label
  non_ref <- tidy(model, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(level = str_remove(term, var_name)) %>%
    transmute(variable = var_label, level, beta = estimate, low = conf.low, high = conf.high)

  # the reference row: beta is 0 by definition, CI is blank
  ref <- tibble(variable = var_label, level = ref_label, beta = 0, low = NA_real_, high = NA_real_)

  bind_rows(ref, non_ref)
}
```

```{r helper-cat-demo}
cat_rows(sex_model, var_name = "sex", var_label = "Sex", ref_label = "Female")
```

**Question**: Name one change that `cat_rows()` made compared with `tidy(sex_model, conf.int = TRUE)`.

**Answer:** It added a reference row for Female (β = 0, no CI), which `tidy()` did not show, and it cleaned the level label from `sexMale` to `Male` by stripping the variable name off the front. (It also renamed the columns and dropped the intercept row.)

### Helper for continuous predictors

```{r helper-cont}
cont_rows <- function(model, var_name, var_label, level_label) {
  tidy(model, conf.int = TRUE) %>%
    filter(term == var_name) %>%
    transmute(variable = var_label, level = level_label, beta = estimate, low = conf.low, high = conf.high)
}
```

```{r helper-cont-demo}
cont_rows(age_model, "chronological_age", "Chronological age", "Per 1-year increase")
```

### Build the first three rows

*Exercise chunk (blanks):*

```{r combine-q, eval=FALSE}
results <- bind_rows(
  cont_rows(age_model, "chronological_age", "Chronological age", "Per 1-year increase"),
  cat_rows(sex_model, "______", "Sex", ref_label = "______"),
  cat_rows(grade_model, "grade_level", "Grade level", ref_label = "Grade 9")
)

results
```

**Solution:**

```{r combine}
results <- bind_rows(
  cont_rows(age_model, "chronological_age", "Chronological age", "Per 1-year increase"),
  cat_rows(sex_model, "sex", "Sex", ref_label = "Female"),
  cat_rows(grade_model, "grade_level", "Grade level", ref_label = "Grade 9")
)

results
```

## Step 4: Assemble the table with `gt`

```{r make-table-fn}
make_table <- function(results_df) {
  results_df %>%
    mutate(
      `Beta (95% CI)` = if_else(
        is.na(low),
        "Reference",
        paste0(
          format(round(beta, 2), nsmall = 2), " (",
          format(round(low, 2), nsmall = 2), " to ",
          format(round(high, 2), nsmall = 2), ")"
        )
      )
    ) %>%
    select(variable, level, `Beta (95% CI)`) %>%
    gt(groupname_col = "variable", rowname_col = "level") %>%
    tab_header(
      title = "Depression (PHQ-9) score by participant characteristic",
      subtitle = "Simple (bivariable) linear regression"
    ) %>%
    tab_style(style = cell_text(weight = "bold"), locations = cells_row_groups()) %>%
    tab_options(table.width = pct(100))
}
```

```{r make-table-call}
make_table(results)
```

## Check these three rows against the paper

```{r compare-1}
results %>% filter(!is.na(low)) %>% mutate(across(c(beta, low, high), ~ round(.x, 2)))
```

**Question**: Which of these three predictors matches the paper's numbers?

**Answer:** **Sex** matches the paper exactly (Male β = -0.66, -1.35 to 0.03), but **chronological age** and **grade level** do not: the two errors we mentioned at the start.

For age, the data give β = 0.45 (0.18 to 0.71), but the paper prints 0.59 (0.20 to 0.99). For grade, with Grade 9 as the stated reference group, the data give Grade 10 = 1.34 (0.47 to 2.21) and Grade 12 = 1.64 (0.70 to 2.58), but the paper prints 0.63 and 0.95. (The paper's table also drops the Grade 11 row altogether, even though those 103 students are in the data and in our model.)

If you find errors in published work, it is sometimes helpful to contact the authors and ask them to correct the table. For now, we ignore the errors and continue with the exercise.

# PART 2: Add two predictors of your own

We illustrate with **place of residence** (binary) and **Oslo social support** (continuous). Any two of the three remaining predictors work.

*Exercise chunk (blanks):*

```{r my-models-q, eval=FALSE}
my_model_1 <-
my_model_2 <-
```

**Solution:**

```{r my-models}
my_model_1 <- lm(depression_score ~ residence, data = survey)
my_model_2 <- lm(depression_score ~ oslo_social_support, data = survey)
```

*Exercise chunk (blanks):*

```{r my-rows-q, eval=FALSE}
my_row_1 <- ______
my_row_2 <- ______
```

**Solution:** residence is binary (use `cat_rows`), Oslo social support is continuous (use `cont_rows`).

```{r my-rows}
my_row_1 <- cat_rows(my_model_1, "residence", "Place of residence", ref_label = "Urban")
my_row_2 <- cont_rows(my_model_2, "oslo_social_support", "Oslo social support", "Per 1-point increase")
```

Stack and inspect the two rows before binding into the full table.

```{r check-my-rows}
my_rows <- bind_rows(my_row_1, my_row_2)
my_rows
```

```{r combine-full}
results_full <- bind_rows(results, my_rows)

make_table(results_full)
```

**Question**: For each predictor you added, does its 95% CI include 0? What does this mean about the significance of the predictor?

**Answer:** Neither CI includes 0. Place of residence: Rural β = 1.46 (0.57 to 2.34). Oslo social support: β = -0.62 (-0.73 to -0.50). Because neither interval includes 0, both predictors are statistically significantly associated with the depression score at alpha = 0.05.

**Question**: Do your two added rows match the paper's simple linear regression column?

**Answer:** Yes. Residence (Rural) 1.46 (0.57 to 2.34) matches exactly. Oslo social support -0.62 (-0.73 to -0.50) matches the paper's -0.62 (-0.73 to -0.49) to two decimals; the only difference is the upper bound, where the raw value -0.4987 sits right on the -0.49/-0.50 rounding boundary.

# PART 3: One framework for t-tests, ANOVA, and continuous predictors

Simple linear regression answers the same questions as a t-test and an ANOVA for categorical predictors, while also handling continuous predictors those tests cannot.

## A binary predictor: regression and the t-test

A two-sample t-test compares two group means, asking whether the difference between them could plausibly be zero. Under equal variances, that test and the regression slope test are the same calculation.

```{r ttest-vs-reg}
t.test(depression_score ~ sex, data = survey, var.equal = TRUE)
tidy(sex_model)
```

**Question**: Compare the p-value from `t.test()` with the p-value for `sexMale` in `tidy(sex_model)`. What do you notice?

**Answer:** They are the same, p = 0.062. The t statistics are equal in size (1.87) and differ only in sign, which reflects the order the groups are subtracted. The equal-variance two-sample t-test *is* a simple linear regression with a binary predictor.

**Question**: Compute the two group means and subtract female from male. Does it match the regression coefficient?

```{r check-means}
survey %>%
  group_by(sex) %>%
  summarise(mean_dep = mean(depression_score))
```

**Answer:** Female mean ≈ 7.03, male mean ≈ 6.37. Male minus female ≈ -0.66, which is exactly the `sexMale` coefficient. The slope for a binary predictor is just the difference in group means.

## A polytomous predictor: regression and ANOVA

With more than two groups there is no single slope; `lm()` gives one coefficient per non-reference grade. A one-way ANOVA instead asks one overall question (is there any difference among the group means?), which is the F test at the bottom of `summary(grade_model)`.

```{r anova-vs-reg}
summary(aov(depression_score ~ grade_level, data = survey))
summary(grade_model)
```

**Question**: Compare the p-value from the ANOVA with the p-value at the bottom of `summary(grade_model)`. What do you notice?

**Answer:** They are the same, p = 0.0023. The overall F test at the bottom of the regression summary is the one-way ANOVA.

**Question**: Do the ANOVA F statistic and the F statistic from `summary(grade_model)` match?

**Answer:** Yes, both are F = 4.90 (on 3 and 542 degrees of freedom). A one-way ANOVA is a simple linear regression with a categorical predictor.

## Continuous predictors

The t-test and ANOVA both compare group means, so they only work with categorical predictors; to use them on a continuous variable you would first have to cut it into groups, throwing away information. Regression has no such limit: a continuous predictor goes into the same `lm()` call and comes out as a single slope, as we saw with `chronological_age` in Part 1.

**Question**: The paper's table mixes categorical predictors (sex, residence) with continuous ones (age, social support) in a single column. Why does regression make this easier than using t-tests and ANOVAs?

**Answer:** One `lm()` call and one output format (β with 95% CI) work for every predictor type, so every row of the table comes from the same procedure. Continuous predictors like age or social support fit naturally as slopes, with no need to cut them into groups the way a t-test or ANOVA would require.

# Optional ungraded challenge: ask an LLM for a figure-table

This is optional and ungraded. There is no single official solution; the point is to practice using an LLM for presentation code, while keeping responsibility for checking the numbers.

The published table is clear, but some papers combine tables and graphics in one figure. For example, look at Fig 2 in this PLOS ONE paper: <https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0304487>

The challenge is to use an LLM to turn the existing `results_full` table into a figure-table with:

- a left panel with the variable and level labels
- a column showing the beta and 95% CI as text
- a forest plot showing the simple-regression slopes and 95% confidence intervals
- stars beside rows where the 95% CI excludes 0

Use `results_full`, which you already created. You do not need to rerun the models.

Here is a suggested prompt:

> I have an R data frame called `results_full` with columns `variable`, `level`, `beta`, `low`, and `high`. It contains rows from simple linear regression models of `depression_score`. Reference rows have `low` and `high` set to `NA`.
>
> I have attached a screenshot of the example figure style I want. Please help me make a similar figure-table with a left text panel, a text column showing beta and 95% CI, and a forest plot of `beta` with 95% confidence intervals. Add a star beside rows where the 95% CI excludes 0. Use `ggplot2` and `patchwork`.
>
> Before writing code, inspect this example script and borrow the overall strategy of building separate panels and combining them: <https://github.com/kendavidn/yaounde_serocovpop_shared/blob/v1.0.0/scripts/infection_regn_2.R>

After the LLM gives you code, check at least three values against `results_full`: one reference row, one categorical coefficient, and one continuous coefficient.

# Summary

Using one simple linear regression per predictor plus `broom::tidy(conf.int = TRUE)` and two small row-building helpers, we built the simple-regression column of the paper's table. Sex, residence, and Oslo social support reproduce the paper exactly; recomputing revealed two errors in the paper (the age and grade rows), for which we report the correct values. Part 3 showed that this single regression framework reproduces the t-test and ANOVA from last week, and additionally handles continuous predictors.
