---
title: "SOLUTIONS: Replicating a multiple linear regression table"
subtitle: "COVID-19 psychological distress (K6) and adjusted coefficients"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

These are the completed solutions for the distress-paper version of workshop 8. Each blank exercise chunk is kept, then followed by a runnable solution chunk.

In week 5, we used the COVID-19 psychological distress paper to reproduce part of a descriptive table. This week we return to the same paper and try to reproduce its **multiple linear regression** table.

The outcome is the **Kessler Psychological Distress Scale (K6)** total score. Higher values mean more psychological distress.

## Important caveat

This exercise is about reproducing and interpreting a published regression table. It is **not** evidence that the predictors cause psychological distress.

Several predictors in the table are measured at the same time as distress and are emotionally close to the outcome, especially `emotional_control`, `epidemic_related_dreams`, and `coping_style`. It is plausible that distress influences these responses, that these responses influence distress, or that both are symptoms of a shared underlying state. Because the study is cross-sectional, use cautious language such as **"associated with, adjusting for..."**, not causal language.

# Load packages and data

*Exercise chunk (blank):*

```{r load-packages-q, eval=FALSE}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

**Solution:**

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(readxl, tidyverse, here, janitor, broom, gt)
```

```{r data-import}
survey_raw <- read_excel(here("data/distress_data.xlsx"))
```

*Exercise chunk (blank):*

```{r glimpse-raw-q, eval=FALSE}
________(survey_raw)
```

**Solution:**

```{r glimpse-raw}
glimpse(survey_raw)
```

**Answer:** The raw dataset has **1599 rows and 38 columns**.

# Part 1: Prepare the variables

## Part 1a: Clean names and select the variables

```{r clean-select}
survey <- survey_raw %>%
  clean_names() %>%
  select(
    total_scores_of_k6,
    age,
    marriage,
    history_of_visiting_sea,
    history_of_epidemics_occurring_in_the_community,
    concern_with_media_reports_related_to_the_epidemic,
    changes_over_living_situations,
    emotional_control,
    epidemic_related_dreams,
    coping_style
  )

glimpse(survey)
```

**Question:** Which variable is the outcome?

**Answer:** `total_scores_of_k6` is the outcome.

## Part 1b: Check the coding

```{r value-check}
survey %>%
  summarise(
    across(
      everything(),
      ~ paste(sort(unique(.x)), collapse = ", ")
    )
  ) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "observed_values")
```

**Answer:** The binary variables are `history_of_visiting_sea`, `history_of_epidemics_occurring_in_the_community`, and `coping_style`. After recoding, `married` will also be binary. The numeric or ordinal predictors are `age`, `concern_with_media_reports_related_to_the_epidemic`, `changes_over_living_situations`, `emotional_control`, and `epidemic_related_dreams`.

## Part 1c: Recode marriage

*Exercise chunk (blank):*

```{r recode-q, eval=FALSE}
survey <- survey %>%
  mutate(
    married = if_else(________ == 1, 1, 0)
  )
```

**Solution:**

```{r recode}
survey <- survey %>%
  mutate(
    married = if_else(marriage == 1, 1, 0)
  )
```

```{r marriage-check}
tabyl(survey, marriage, married)
```

**Answer:** The new `married` variable compares married participants (`married = 1`) with everyone else (`married = 0`).

# Part 2: Fit the multiple regression model

## Part 2a: Start with the paper's final model

```{r table4-model}
table4_model <- lm(
  total_scores_of_k6 ~
    married +
    history_of_visiting_sea +
    changes_over_living_situations +
    emotional_control +
    epidemic_related_dreams +
    coping_style,
  data = survey
)

summary(table4_model)
```

**Question:** What is the model R-squared? How would you interpret it?

**Answer:** The R-squared is about **0.862**. About 86.2% of the variation in K6 distress scores is explained by this set of predictors in the fitted sample. This is an in-sample association statement, not a causal statement.

## Part 2b: Interpret adjusted coefficients

**Answer:** The coefficient for `history_of_visiting_sea` is about **1.449**. Holding the other variables constant, people with a history of visiting Wuhan have predicted K6 scores about **1.45 points higher** than those without such a history.

**Answer:** The coefficient for `coping_style` is about **-2.135**. Holding the other variables constant, participants coded as having a positive coping style have predicted K6 scores about **2.14 points lower** than participants coded as having a negative coping style.

**Answer:** The coefficient for `emotional_control` is about **1.995**. Holding the other variables constant, each 1-point increase in reported difficulty with emotional control is associated with about **2.00 points higher** predicted K6 score. We should be cautious because difficulty controlling emotions may be partly a symptom or consequence of psychological distress, not necessarily a prior cause of it.

# Part 3: Add confidence intervals and build a table

## Part 3a: Use `broom::tidy()`

*Exercise chunk (blank):*

```{r tidy-model-q, eval=FALSE}
model_results <- tidy(________, conf.int = TRUE)
model_results
```

**Solution:**

```{r tidy-model}
model_results <- tidy(table4_model, conf.int = TRUE)
model_results
```

## Part 3b: Format the regression table

*Exercise chunk (blank):*

```{r make-table-q, eval=FALSE}
________ %>%
  mutate(
    variable = recode(
      term,
      "(Intercept)" = "Constant",
      "married" = "Marriage",
      "history_of_visiting_sea" = "History of visiting Wuhan",
      "changes_over_living_situations" = "Changes over living situations",
      "emotional_control" = "Emotional control",
      "epidemic_related_dreams" = "Epidemic-related dreams",
      "coping_style" = "Coping style"
    ),
    beta = format(round(estimate, 3), nsmall = 3),
    t = format(round(statistic, 3), nsmall = 3),
    p = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    `95% CI` = paste0(
      format(round(conf.low, 3), nsmall = 3),
      " to ",
      format(round(conf.high, 3), nsmall = 3)
    )
  ) %>%
  select(variable, beta, t, p, `95% CI`) %>%
  gt() %>%
  tab_header(
    title = "Stepwise multiple linear regression for psychological distress",
    subtitle = "Outcome: K6 score"
  )
```

**Solution:**

```{r make-table}
model_results %>%
  mutate(
    variable = recode(
      term,
      "(Intercept)" = "Constant",
      "married" = "Marriage",
      "history_of_visiting_sea" = "History of visiting Wuhan",
      "changes_over_living_situations" = "Changes over living situations",
      "emotional_control" = "Emotional control",
      "epidemic_related_dreams" = "Epidemic-related dreams",
      "coping_style" = "Coping style"
    ),
    beta = format(round(estimate, 3), nsmall = 3),
    t = format(round(statistic, 3), nsmall = 3),
    p = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    `95% CI` = paste0(
      format(round(conf.low, 3), nsmall = 3),
      " to ",
      format(round(conf.high, 3), nsmall = 3)
    )
  ) %>%
  select(variable, beta, t, p, `95% CI`) %>%
  gt() %>%
  tab_header(
    title = "Stepwise multiple linear regression for psychological distress",
    subtitle = "Outcome: K6 score"
  )
```

**Answer:** All six predictors in the Table 4 model are statistically significant at alpha = 0.05. Each p-value is below 0.001.

# Part 4: Compare with the paper

```{r paper-values}
paper_table4 <- tribble(
  ~variable, ~paper_beta, ~paper_t, ~paper_ci,
  "Constant", 3.526, 13.939, "3.030 to 4.022",
  "Marriage", -0.588, -3.682, "-0.901 to -0.275",
  "History of visiting Wuhan", 1.449, 5.265, "0.909 to 1.898",
  "Changes over living situations", 1.040, 9.057, "0.815 to 1.266",
  "Emotional control", 1.995, 21.879, "1.816 to 2.174",
  "Epidemic-related dreams", 1.597, 16.964, "1.412 to 1.782",
  "Coping style", -2.135, -10.575, "-2.531 to -1.739"
)

paper_table4
```

```{r compare-paper}
our_table4 <- model_results %>%
  mutate(
    variable = recode(
      term,
      "(Intercept)" = "Constant",
      "married" = "Marriage",
      "history_of_visiting_sea" = "History of visiting Wuhan",
      "changes_over_living_situations" = "Changes over living situations",
      "emotional_control" = "Emotional control",
      "epidemic_related_dreams" = "Epidemic-related dreams",
      "coping_style" = "Coping style"
    ),
    our_beta = round(estimate, 3),
    our_t = round(statistic, 3),
    our_ci = paste0(
      round(conf.low, 3),
      " to ",
      round(conf.high, 3)
    )
  ) %>%
  select(variable, our_beta, our_t, our_ci)

left_join(paper_table4, our_table4, by = "variable")
```

**Answer:** The estimates and t statistics match the paper after rounding. A few confidence interval endpoints differ slightly in the last decimal, which is likely due to rounding or software conventions.

# Part 5: What did stepwise selection leave out?

```{r larger-model}
larger_model <- lm(
  total_scores_of_k6 ~
    age +
    married +
    history_of_visiting_sea +
    history_of_epidemics_occurring_in_the_community +
    concern_with_media_reports_related_to_the_epidemic +
    changes_over_living_situations +
    emotional_control +
    epidemic_related_dreams +
    coping_style,
  data = survey
)

summary(larger_model)
```

**Answer:** In this larger model, `age`, `history_of_epidemics_occurring_in_the_community`, and `concern_with_media_reports_related_to_the_epidemic` are not statistically significant at alpha = 0.05.

**Answer:** Stepwise selection is a mechanical search through possible models. It can be unstable, depends on the exact sample and candidate variables, and usually makes p-values look cleaner than they really are because the model was selected using the same data.

# Part 6: Practice cautious reporting

**Example answer:** In a multiple linear regression predicting K6 psychological distress score, the final model explained about 86.2% of the variation in K6 scores. Adjusting for the other variables in the model, history of visiting Wuhan was associated with a 1.45-point higher predicted K6 score, while positive coping style was associated with a 2.14-point lower predicted K6 score. Because the data are cross-sectional and several predictors are emotional responses measured at the same time as distress, these results should be interpreted as adjusted associations, not evidence of causation.

# Wrap-up

In this workshop, you practiced fitting a multiple regression from a published table, interpreting adjusted coefficients, reproducing a table with `broom` and `gt`, comparing output against a paper, and explaining why cross-sectional regression tables need cautious interpretation.
