---
title: "EXERCISE: Replicating a multiple linear regression table"
subtitle: "COVID-19 psychological distress (K6) and adjusted coefficients"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In week 5, we used the COVID-19 psychological distress paper to reproduce part of a descriptive table. This week we return to the same paper and try to reproduce its **multiple linear regression** table.

The paper is:

- Wang et al. (2020), *The psychological distress and coping styles in the early stages of the 2019 coronavirus disease (COVID-19) epidemic in the general mainland Chinese population*
- A copy is in `paper/distress_paper.pdf`

The outcome is the **Kessler Psychological Distress Scale (K6)** total score, called `total_scores_of_K6` in the raw data. Higher values mean more psychological distress.

## What we are reproducing

The paper's Table 4 reports a stepwise multiple linear regression for K6 score. The final table includes:

- marriage
- history of visiting Wuhan
- changes over living situations
- emotional control
- epidemic-related dreams
- coping style

## Important caveat

This exercise is about reproducing and interpreting a published regression table. It is **not** evidence that the predictors cause psychological distress.

Several predictors in the table are measured at the same time as distress and are emotionally close to the outcome, especially `emotional_control`, `epidemic_related_dreams`, and `coping_style`. It is plausible that distress influences these responses, that these responses influence distress, or that both are symptoms of a shared underlying state. Because the study is cross-sectional, use cautious language such as **"associated with, adjusting for..."**, not causal language.

# Load packages and data

Use `p_load()` to load `readxl`, `tidyverse`, `here`, `janitor`, `broom`, and `gt`.

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

Import the dataset.

```{r data-import}
survey_raw <- read_excel(here("data/distress_data.xlsx"))
```

Inspect the raw data.

```{r glimpse-raw}
________(survey_raw)
```

**Question:** How many rows and columns are in the raw dataset?

Ans:

# Part 1: Prepare the variables

## Part 1a: Clean names and select the variables

The raw names include spaces and mixed capitalization. We will clean them and keep only the columns needed for Table 4.

```{r clean-select}
survey <- survey_raw %>%
  clean_names() %>%
  select(
    total_scores_of_k6,
    age,
    marriage,
    history_of_visiting_sea,
    history_of_epidemics_occurring_in_the_community,
    concern_with_media_reports_related_to_the_epidemic,
    changes_over_living_situations,
    emotional_control,
    epidemic_related_dreams,
    coping_style
  )

glimpse(survey)
```

**Question:** Which variable is the outcome?

Ans:

## Part 1b: Check the coding

Table 3 in the paper describes how variables were coded for the regression model. Run the code below to see the observed values in our selected data.

```{r value-check}
survey %>%
  summarise(
    across(
      everything(),
      ~ paste(sort(unique(.x)), collapse = ", ")
    )
  ) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "observed_values")
```

**Question:** Which selected variables are binary? Which are numeric or ordinal scores?

Ans:

## Part 1c: Recode marriage

The codebook lists `marriage` as 0 = unmarried and 1 = married, but the data also contain a small number of people coded 2. In week 5, we treated those people as unmarried/non-married. We will use the same practical recode here:

- `married = 1` if `marriage == 1`
- `married = 0` otherwise

```{r recode}
survey <- survey %>%
  mutate(
    married = if_else(________ == 1, 1, 0)
  )
```

Check the recode.

```{r marriage-check}
tabyl(survey, marriage, married)
```

**Question:** What does the new `married` variable compare?

Ans:

# Part 2: Fit the multiple regression model

## Part 2a: Start with the paper's final model

The paper says Table 4 is the final model from stepwise regression. We will not redo the stepwise search. Instead, we will directly fit the final set of predictors shown in the published table.

Fit a model predicting K6 score from the six Table 4 predictors.

```{r table4-model}
table4_model <- lm(
  total_scores_of_k6 ~
    married +
    history_of_visiting_sea +
    changes_over_living_situations +
    emotional_control +
    epidemic_related_dreams +
    coping_style,
  data = survey
)

summary(table4_model)
```

**Question:** What is the model R-squared? How would you interpret it?

Ans:

## Part 2b: Interpret adjusted coefficients

In a multiple regression, each coefficient is interpreted while holding the other predictors constant.

Use `summary(table4_model)` to answer the questions below.

**Question:** What is the coefficient for `history_of_visiting_sea`? Interpret it in one sentence using "holding the other variables constant".

Ans:

**Question:** What is the coefficient for `coping_style`? The coding is 0 = negative coping style and 1 = positive coping style. Interpret the coefficient in one sentence.

Ans:

**Question:** What is the coefficient for `emotional_control`? Why should we be cautious about interpreting this coefficient causally?

Ans:

# Part 3: Add confidence intervals and build a table

## Part 3a: Use `broom::tidy()`

Use `tidy()` to extract the estimates, p-values, and confidence intervals.

```{r tidy-model}
model_results <- tidy(________, conf.int = TRUE)
model_results
```

## Part 3b: Format the regression table

The table below uses `gt` to format the estimates. Fill in the missing data frame name.

```{r make-table}
________ %>%
  mutate(
    variable = recode(
      term,
      "(Intercept)" = "Constant",
      "married" = "Marriage",
      "history_of_visiting_sea" = "History of visiting Wuhan",
      "changes_over_living_situations" = "Changes over living situations",
      "emotional_control" = "Emotional control",
      "epidemic_related_dreams" = "Epidemic-related dreams",
      "coping_style" = "Coping style"
    ),
    beta = format(round(estimate, 3), nsmall = 3),
    t = format(round(statistic, 3), nsmall = 3),
    p = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    `95% CI` = paste0(
      format(round(conf.low, 3), nsmall = 3),
      " to ",
      format(round(conf.high, 3), nsmall = 3)
    )
  ) %>%
  select(variable, beta, t, p, `95% CI`) %>%
  gt() %>%
  tab_header(
    title = "Stepwise multiple linear regression for psychological distress",
    subtitle = "Outcome: K6 score"
  )
```

**Question:** Which variables in this table are statistically significant at alpha = 0.05?

Ans:

# Part 4: Compare with the paper

Create a small table of the values printed in the paper's Table 4.

```{r paper-values}
paper_table4 <- tribble(
  ~variable, ~paper_beta, ~paper_t, ~paper_ci,
  "Constant", 3.526, 13.939, "3.030 to 4.022",
  "Marriage", -0.588, -3.682, "-0.901 to -0.275",
  "History of visiting Wuhan", 1.449, 5.265, "0.909 to 1.898",
  "Changes over living situations", 1.040, 9.057, "0.815 to 1.266",
  "Emotional control", 1.995, 21.879, "1.816 to 2.174",
  "Epidemic-related dreams", 1.597, 16.964, "1.412 to 1.782",
  "Coping style", -2.135, -10.575, "-2.531 to -1.739"
)

paper_table4
```

Now create a rounded version of your model results and join it to the paper values.

```{r compare-paper}
our_table4 <- model_results %>%
  mutate(
    variable = recode(
      term,
      "(Intercept)" = "Constant",
      "married" = "Marriage",
      "history_of_visiting_sea" = "History of visiting Wuhan",
      "changes_over_living_situations" = "Changes over living situations",
      "emotional_control" = "Emotional control",
      "epidemic_related_dreams" = "Epidemic-related dreams",
      "coping_style" = "Coping style"
    ),
    our_beta = round(estimate, 3),
    our_t = round(statistic, 3),
    our_ci = paste0(
      round(conf.low, 3),
      " to ",
      round(conf.high, 3)
    )
  ) %>%
  select(variable, our_beta, our_t, our_ci)

left_join(paper_table4, our_table4, by = "variable")
```

**Question:** Do your estimates match the paper? If not, where are the small differences?

Ans:

# Part 5: What did stepwise selection leave out?

Table 3 says the stepwise procedure considered additional predictors, including age, history of epidemics in the community, and concern with media reports. Fit a larger model with those predictors added.

```{r larger-model}
larger_model <- lm(
  total_scores_of_k6 ~
    age +
    married +
    history_of_visiting_sea +
    history_of_epidemics_occurring_in_the_community +
    concern_with_media_reports_related_to_the_epidemic +
    changes_over_living_situations +
    emotional_control +
    epidemic_related_dreams +
    coping_style,
  data = survey
)

summary(larger_model)
```

**Question:** Which of the added predictors are not statistically significant in the larger model?

Ans:

**Question:** Why should we be cautious about using stepwise variable selection as if it discovers the one true model?

Ans:

# Part 6: Practice cautious reporting

Write a two-to-three sentence result statement for the Table 4 model. Include:

- the outcome
- at least two adjusted coefficients
- one caveat about causality or reverse causation

Your answer:

# Wrap-up

In this workshop, you practiced:

1. fitting a multiple linear regression from a published table;
2. interpreting adjusted coefficients;
3. reproducing a regression table with `broom` and `gt`;
4. comparing your output against a paper;
5. explaining why cross-sectional regression tables need cautious interpretation.
