---
title: "EXERCISE: Replicating a multiple linear regression table"
subtitle: "COVID-19 psychological distress (K6) in mainland China"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Introduction

In Workshop 5 you reproduced **Table 2** from Wang et al. (2020): mean K6 distress scores by participant characteristic, with t-tests and ANOVAs.

In this workshop we move to **Table 4** in the same paper: a **stepwise multiple linear regression** model that adjusts for several predictors at once. This connects directly to the course material on multiple linear regression.

## The study (recap)

- Paper: Wang et al. (2020), *The psychological distress and coping styles in the early stages of the COVID-19 epidemic*, PLOS ONE. See `paper/distress_paper.pdf`.
- Data: `data/distress_data.xlsx` (1,599 participants)
- Outcome: `total_scores_of_K6` (0 to 24, higher = more distress)

## The table we will build

**Table 4** reports adjusted beta coefficients from a multivariate model. The authors' final model includes:

- Marriage (binary)
- History of visiting Wuhan in the past month (binary)
- Changes over living situations (0 to 2)
- Emotional control difficulty (0 to 4)
- Epidemic-related dreams (0 to 4)
- Coping style (negative vs positive)

The paper reports R-squared about **0.862** for this model.

Open `paper/distress_paper.pdf` and find Table 4 before you start.

## A note on interpretation (read this carefully)

We will reproduce the authors' table because it is good practice with real published output. But **reproducing a table is not the same as endorsing every predictor as a causal driver of distress**.

Three cautions:

1. **Cross-sectional data.** Everything was measured at one time point, so we cannot tell what came first.
2. **Symptom overlap.** Items like *difficulty controlling emotion* and *epidemic-related dreams* sound like **distress symptoms themselves**, not just independent risk factors. Putting them on the right-hand side of a regression that predicts K6 distress risks **reverse causation** or **circular reasoning**.
3. **Coping style** may also be shaped by current mood, not only the other way around.

Throughout the exercise, practice describing results as **adjusted associations**, not proven causes.

# PART 1: Load packages and prepare variables

## Load packages

Fill in the blanks to load: `readxl`, `tidyverse`, `here`, `janitor`, `broom`, `gt`.

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

## Import the data

```{r import-raw}
survey_raw <- read_excel(here("data/distress_data.xlsx"))
glimpse(survey_raw)
```

**Checkpoint:** The raw data should have **1,599 rows**.

## Create analysis variables

The paper's Table 3 lists how variables were coded for regression. We need to recreate two derived variables:

- `marriage_bin`: 0 = unmarried, 1 = married
- `coping_style`: 0 = negative, 1 = positive (from the SCSQ questionnaire; see codebook)

The coping-style steps match the paper:

1. Sum the 12 positive-coping SCSQ items and the 8 negative-coping items.
2. Z-score each sum.
3. Subtract the negative z-score from the positive z-score.
4. Code as positive (1) if the result is >= 0, else negative (0).

The positive item numbers are 1, 2, 3, 5, 6, 7, 9, 10, 12, 15, 16, 19. The negative item numbers are 4, 8, 11, 13, 14, 17, 18, 20.

Most of the code below is written out. Fill in the blanks for `marriage_bin` and the `ifelse()` that creates `coping_style`.

```{r prep-vars}
pos_items <- paste0("SCSQ_qes", c(1, 2, 3, 5, 6, 7, 9, 10, 12, 15, 16, 19))
neg_items <- paste0("SCSQ_qes", c(4, 8, 11, 13, 14, 17, 18, 20))

survey <- survey_raw %>%
  mutate(
    scsq_positive = rowSums(across(all_of(pos_items)), na.rm = TRUE),
    scsq_negative = rowSums(across(all_of(neg_items)), na.rm = TRUE),
    coping_tendency = as.numeric(scale(scsq_positive) - scale(scsq_negative)),
    coping_style = ifelse(________ >= 0, 1, 0),
    marriage_bin = ifelse(________ == 1, 1, 0)
  ) %>%
  select(
    total_scores_of_K6,
    marriage_bin,
    history_of_visiting_SEA,
    changes_over_living_situations,
    emotional_control,
    `epidemic-related_dreams`,
    coping_style
  )

glimpse(survey)
```

**Question:** How many people have positive coping style (`coping_style = 1`)? Use `count(coping_style)`.

```{r count-coping, eval=FALSE}
survey %>% count(________)
```

Your answer: _______

**Question:** Why do we keep `emotional_control` and `epidemic-related_dreams` as numeric codes (0 to 4) rather than converting them to factors?

Your answer: _______

# PART 2: Fit the multiple linear regression model

The paper's Table 4 model regresses K6 on six predictors at once. Fill in the formula.

```{r mlr-model, eval=FALSE}
mlr_model <- lm(
  total_scores_of_K6 ~ ________ + ________ + ________ +
    emotional_control + `epidemic-related_dreams` + coping_style,
  data = survey
)

summary(mlr_model)
```

**Question:** What is the adjusted R-squared? Is it close to the paper's 0.862?

Your answer: _______

**Question:** The coefficient on `marriage_bin` is negative. Fill in the interpretation using cautious language.

> Holding the other predictors constant, married participants have K6 scores about ______ points ______ (higher / lower) than unmarried participants, on average.

Your answer: _______

**Question:** The coefficient on `emotional_control` is positive and highly significant. Why might we hesitate to call this a "risk factor" for distress?

Your answer: _______

# PART 3: Extract coefficients with `broom`

To build a publication-style table, use `broom::tidy()` with confidence intervals.

```{r tidy-mlr, eval=FALSE}
mlr_results <- tidy(________, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    variable = case_when(
      term == "marriage_bin" ~ "Marriage (married vs unmarried)",
      term == "history_of_visiting_SEA" ~ "History of visiting Wuhan",
      term == "changes_over_living_situations" ~ "Changes over living situations",
      term == "emotional_control" ~ "Emotional control difficulty",
      term == "`epidemic-related_dreams`" ~ "Epidemic-related dreams",
      term == "coping_style" ~ "Coping style (positive vs negative)",
      TRUE ~ term
    )
  ) %>%
  select(variable, beta = estimate, low = conf.low, high = conf.high, p.value)

mlr_results
```

**Question:** For `coping_style`, does the 95% CI exclude 0? What does that mean at alpha = 0.05?

Your answer: _______

## Build the table with `gt`

```{r make-table, eval=FALSE}
make_mlr_table <- function(results_df) {
  results_df %>%
    mutate(
      `Beta (95% CI)` = paste0(
        format(round(beta, 3), nsmall = 3), " (",
        format(round(low, 3), nsmall = 3), " to ",
        format(round(high, 3), nsmall = 3), ")"
      ),
      `p-value` = if_else(p.value < 0.001, "<0.001", format(round(p.value, 3), nsmall = 3))
    ) %>%
    select(variable, `Beta (95% CI)`, `p-value`) %>%
    gt() %>%
    tab_header(
      title = "Table 4 replication: factors associated with K6 distress",
      subtitle = "Multiple linear regression (adjusted coefficients)"
    ) %>%
    tab_options(table.width = pct(100))
}

make_mlr_table(________)
```

# PART 4: Compare to the published table

Round your coefficients to three decimals and compare to Table 4 in the paper.

```{r compare-paper, eval=FALSE}
mlr_results %>%
  mutate(across(c(beta, low, high), ~ round(.x, 3)))
```

**Question:** Do your coefficients match the paper's betas (for example, marriage about -0.588, emotional control about 1.995, coping style about -2.135)?

Your answer: _______

**Question:** The paper's abstract says age was associated with distress in bivariate tests, but age is **not** in Table 4. What does that suggest about stepwise variable selection?

Your answer: _______

# PART 5: Simple vs multiple regression for one predictor

Pick **history of visiting Wuhan** (`history_of_visiting_SEA`). Fit:

1. A simple regression (one predictor)
2. The full multiple regression (already fitted)

```{r simple-wuhan, eval=FALSE}
wuhan_simple <- lm(total_scores_of_K6 ~ history_of_visiting_SEA, data = survey)
tidy(wuhan_simple, conf.int = TRUE)
tidy(mlr_model, conf.int = TRUE) %>% filter(term == "history_of_visiting_SEA")
```

**Question:** Is the Wuhan coefficient similar in the simple and multiple models, or does it change a lot? What might explain the difference?

Your answer: _______

# PART 6: Thinking critically about the model

**Question:** The model R-squared is very high (about 0.86). Does that mean this is a good **causal** model of distress?

Your answer: _______

**Question:** If you were designing a follow-up study, which two predictors from Table 4 would you treat most cautiously as "exposures", and why?

Your answer: _______

**Question:** Rewrite one sentence from the paper's Discussion in safer language. The original says emotional control difficulty was an "influence factor" of psychological distress. Write a version using "associated with" and mentioning the cross-sectional design.

Your answer: _______

# Wrap-up

You recreated a published **multiple linear regression** table, extracted adjusted coefficients with `broom`, and compared simple vs adjusted estimates.

The statistical skills are real and useful. The **interpretation** skills are just as important: in cross-sectional mental-health data, not every variable on the right-hand side of the regression is a clean causal predictor.
