---
title: "SOLUTIONS: Multiple linear regression with FEV"
subtitle: "Adjusting for confounders in lung function data"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

These are the completed solutions for the FEV multiple linear regression workshop. Each step first shows the blanked chunk from the exercise (not run), then the filled-in **Solution** chunk that runs.

In Workshop 6 you used simple linear regression (one predictor at a time). Here we fit **two or more predictors in the same model** and read each coefficient while holding the others constant.

# PART 1: Load packages and data

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, broom)
```

```{r load-data}
fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint:** `fev` should have **654 rows and 5 columns**.

**Checkpoint answer:** `fev` has **654 rows and 5 columns**.

# PART 2: From one predictor to two

## Part 2a: Two simple models, side by side

*Exercise chunk (blank):*

```{r height-simple-q, eval=FALSE}
height_simple <- lm(________ ~ ________, data = fev)
summary(height_simple)
```

**Solution:**

```{r age-simple}
age_simple <- lm(fev_l ~ age_yrs, data = fev)
summary(age_simple)
```

```{r height-simple}
height_simple <- lm(fev_l ~ height_in, data = fev)
summary(height_simple)
```

**Question:** What is the slope on `height_in` in the simple height model?

**Answer:** About **0.132** litres per inch.

**Question:** Which simple model has higher R-squared?

**Answer:** **Height** (R-squared about 0.754) beats age (about 0.572).

## Part 2b: Fit a two-predictor model

*Exercise chunk (blank):*

```{r age-height-model-q, eval=FALSE}
age_height_model <- lm(fev_l ~ ________ + ________, data = fev)
summary(age_height_model)
```

**Solution:**

```{r age-height-model}
age_height_model <- lm(fev_l ~ age_yrs + height_in, data = fev)
summary(age_height_model)
```

**Question:** Why did the age slope get smaller after adding height?

**Answer:** Much of what looked like an "age effect" in the simple model was really a **height effect**. Older youths are taller, and taller youths have larger lungs. Once height is in the model, age only picks up the extra FEV not already explained by height.

**Question:** Interpret the adjusted age slope.

**Answer:** Holding height constant, each extra year of age is associated with about **0.054** more litres of FEV, on average.

**Question:** Interpret the adjusted height slope.

**Answer:** Holding age constant, each extra inch of height is associated with about **0.110** more litres of FEV, on average.

## Part 2c: Compare coefficients in one table

*Exercise chunk (blank):*

```{r compare-slopes-q, eval=FALSE}
compare_slopes <- bind_rows(
  tidy(age_simple) %>% mutate(model = "Age only"),
  tidy(height_simple) %>% mutate(model = "Height only"),
  tidy(age_height_model) %>% mutate(model = "Age + height")
) %>%
  filter(term != "(Intercept)") %>%
  select(model, term, estimate)

compare_slopes
```

**Solution:**

```{r compare-slopes}
compare_slopes <- bind_rows(
  tidy(age_simple) %>% mutate(model = "Age only"),
  tidy(height_simple) %>% mutate(model = "Height only"),
  tidy(age_height_model) %>% mutate(model = "Age + height")
) %>%
  filter(term != "(Intercept)") %>%
  select(model, term, estimate)

compare_slopes
```

**Question:** What happens to the age slope when height enters the model?

**Answer:** It drops sharply from about **0.222** to about **0.054**, because height absorbs much of the variation that age was capturing on its own.

# PART 3: The smoking confounding story, revisited

## Part 3a: Summarise by smoking status

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    n = n(),
    mean_age = mean(age_yrs),
    mean_height = mean(height_in),
    mean_fev = mean(fev_l),
    .groups = "drop"
  )
```

**Question:** Why might age and height act as confounders?

**Answer:** Age and height influence **both** smoking status (smokers are older) and FEV (older, taller youths have larger lungs). A one-predictor smoking comparison mixes those paths together.

## Part 3b: Raw smoking model

*Exercise chunk (blank):*

```{r smoke-raw-q, eval=FALSE}
smoke_raw <- lm(________ ~ ________, data = fev)
summary(smoke_raw)
```

**Solution:**

```{r smoke-raw}
smoke_raw <- lm(fev_l ~ smoke, data = fev)
summary(smoke_raw)
```

**Question:** What does the raw smoking coefficient mean?

**Answer:** Smokers have mean FEV about **0.71 litres higher** than non-smokers in this unadjusted comparison. That is a difference in group means, not proof that smoking improves lungs.

## Part 3c: Adjusted smoking model

*Exercise chunk (blank):*

```{r smoke-adjusted-q, eval=FALSE}
smoke_adjusted <- lm(fev_l ~ smoke + ________ + ________, data = fev)
summary(smoke_adjusted)
```

**Solution:**

```{r smoke-adjusted}
smoke_adjusted <- lm(fev_l ~ smoke + age_yrs + height_in, data = fev)
summary(smoke_adjusted)
```

**Question:** How does the adjusted coefficient differ from the raw one?

**Answer:** After adjusting for age and height, the smoking coefficient flips to about **-0.11** (smokers lower, not higher). The p-value is about **0.067**, so evidence is suggestive but not strong at alpha = 0.05.

**Question:** Write one safer sentence.

**Answer:** After adjusting for age and height, current smoking is **associated with** slightly lower predicted FEV, but this observational analysis cannot prove causation.

# PART 4: Read the full three-predictor model

*Exercise chunk (blank):*

```{r full-model-q, eval=FALSE}
full_model <- lm(fev_l ~ smoke + age_yrs + height_in, data = fev)
summary(full_model)
```

**Solution:**

```{r full-model}
full_model <- lm(fev_l ~ smoke + age_yrs + height_in, data = fev)
summary(full_model)
```

**Question:** Which slope is largest in absolute value? Does that mean it is most important?

**Answer:** The intercept and height slope look large numerically, but **units differ**. An extra inch of height is common; switching smoking status is a one-time 0/1 contrast. You cannot rank "importance" from raw slopes alone.

**Question:** What is the adjusted R-squared?

**Answer:** About **0.767**.

# PART 5: Prediction with multiple predictors

```{r predict-example}
new_person <- tibble(
  smoke = factor("non_smoker", levels = c("non_smoker", "smoker")),
  age_yrs = 12,
  height_in = 60
)

predict(full_model, newdata = new_person)
```

*Exercise chunk (blank):*

```{r predict-smoker-q, eval=FALSE}
smoker_12_60 <- tibble(
  smoke = factor("smoker", levels = c("non_smoker", "smoker")),
  age_yrs = ________,
  height_in = ________
)

predict(________, newdata = ________)
```

**Solution:**

```{r predict-smoker}
smoker_12_60 <- tibble(
  smoke = factor("smoker", levels = c("non_smoker", "smoker")),
  age_yrs = 12,
  height_in = 60
)

predict(full_model, newdata = smoker_12_60)
```

**Question:** Is predicted FEV higher or lower for the smoker?

**Answer:** **Lower** for the smoker (about 2.54 L vs 2.65 L for the non-smoker at the same age and height).

*Exercise chunk (blank):*

```{r predict-tall-q, eval=FALSE}
tall_non_smoker <- tibble(
  smoke = factor("non_smoker", levels = c("non_smoker", "smoker")),
  age_yrs = ________,
  height_in = ________
)

predict(________, newdata = ________)
```

**Solution:**

```{r predict-tall}
tall_non_smoker <- tibble(
  smoke = factor("non_smoker", levels = c("non_smoker", "smoker")),
  age_yrs = 16,
  height_in = 70
)

predict(full_model, newdata = tall_non_smoker)
```

**Question:** Is the prediction plausible?

**Answer:** About **3.45 litres** for a 16-year-old, 70-inch non-smoker. That is high but within the range of this dataset, so it is more plausible than extrapolating to a 20-inch-tall baby.

# PART 6: R-squared across models

*Exercise chunk (blank):*

```{r r-squared-compare-q, eval=FALSE}
tibble(
  model = c("Age only", "Height only", "Age + height", "Smoke + age + height"),
  r_squared = c(
    summary(________)$r.squared,
    summary(________)$r.squared,
    summary(________)$r.squared,
    summary(________)$r.squared
  )
)
```

**Solution:**

```{r r-squared-compare}
tibble(
  model = c("Age only", "Height only", "Age + height", "Smoke + age + height"),
  r_squared = c(
    summary(age_simple)$r.squared,
    summary(height_simple)$r.squared,
    summary(age_height_model)$r.squared,
    summary(full_model)$r.squared
  )
)
```

**Question:** What happens to R-squared when you add predictors?

**Answer:** R-squared **increases** (or stays the same) as you add predictors. A higher R-squared means more variation explained, but it does not by itself mean the model is causal, correct, or best for prediction on new data.

# Wrap-up

You fitted multi-predictor models, interpreted adjusted coefficients, revisited the smoking confounding example, made predictions with several inputs, and compared R-squared across models.
