---
title: "EXERCISE: Multiple linear regression with FEV"
subtitle: "Adjusting for confounders in lung function data"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In Workshop 6 you used **simple linear regression** (one predictor at a time) to study forced expiratory volume (FEV) in children and adolescents. You saw that a raw smoking comparison was misleading because smokers were older and taller on average.

In this workshop we go further with **multiple linear regression**: fitting **two or more predictors in the same model**. The key idea from the course materials is that each coefficient is read while **holding the other predictors constant**. This is also called **adjusting for** or **controlling for** the other variables.

We will use the same FEV dataset from Workshop 6 (`data/fev_kahn.csv`, 654 youths aged 3 to 19).

The outcome is `fev_l` (litres). The predictors are:

- `age_yrs`: age in years
- `height_in`: height in inches
- `smoke`: current smoking status (`non_smoker` or `smoker`)

# PART 1: Load packages and data

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, broom)
```

```{r load-data}
fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint:** `fev` should have **654 rows and 5 columns**.

# PART 2: From one predictor to two

## Part 2a: Two simple models, side by side

First, refit the two one-predictor models from Workshop 6.

The age model is filled in as an example. Fill in the height model.

```{r age-simple}
age_simple <- lm(fev_l ~ age_yrs, data = fev)
summary(age_simple)
```

```{r height-simple, eval=FALSE}
height_simple <- lm(________ ~ ________, data = fev)
summary(height_simple)
```

**Question:** What is the slope on `height_in` in the simple height model (about how many litres per inch)?

Your answer: _______

**Question:** Which simple model has higher R-squared, age or height?

Your answer: _______

## Part 2b: Fit a two-predictor model

Now fit both predictors together.

```{r age-height-model, eval=FALSE}
age_height_model <- lm(fev_l ~ ________ + ________, data = fev)
summary(age_height_model)
```

**Question:** In the two-predictor model, the slope on `age_yrs` is about **0.054**. In the simple age model it was about **0.222**. Why did the age slope get smaller after adding height?

Your answer: _______

**Question:** Fill in the interpretation.

> Holding height constant, each extra year of age is associated with about ______ more litres of FEV, on average.

Your answer: _______

**Question:** Fill in the interpretation for height.

> Holding age constant, each extra inch of height is associated with about ______ more litres of FEV, on average.

Your answer: _______

## Part 2c: Compare coefficients in one table

`broom::tidy()` is useful for lining up coefficients from different models.

```{r compare-slopes, eval=FALSE}
compare_slopes <- bind_rows(
  tidy(age_simple) %>% mutate(model = "Age only"),
  tidy(height_simple) %>% mutate(model = "Height only"),
  tidy(age_height_model) %>% mutate(model = "Age + height")
) %>%
  filter(term != "(Intercept)") %>%
  select(model, term, estimate)

compare_slopes
```

**Question:** Look at the `age_yrs` row across models. What happens to the age slope when height enters the model?

Your answer: _______

# PART 3: The smoking confounding story, revisited

Workshop 6 showed that smokers had **higher** mean FEV in a raw comparison, but **lower** predicted FEV after adjusting for age and height. Now we will unpack that more carefully.

## Part 3a: Summarise by smoking status

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    n = n(),
    mean_age = mean(age_yrs),
    mean_height = mean(height_in),
    mean_fev = mean(fev_l),
    .groups = "drop"
  )
```

**Question:** Smokers have higher mean FEV. They are also older and taller on average. In one sentence, why might age and height act as confounders here?

Your answer: _______

## Part 3b: Raw smoking model

```{r smoke-raw, eval=FALSE}
smoke_raw <- lm(________ ~ ________, data = fev)
summary(smoke_raw)
```

**Question:** The coefficient on `smokesmoker` in the raw model is about **+0.71**. What does that number mean?

Your answer: _______

## Part 3c: Adjusted smoking model

Now add age and height to the model.

```{r smoke-adjusted, eval=FALSE}
smoke_adjusted <- lm(fev_l ~ smoke + ________ + ________, data = fev)
summary(smoke_adjusted)
```

**Question:** After adjustment, the smoking coefficient is about **-0.11** with p about **0.067**. How does this differ from the raw coefficient? What cautious conclusion would you draw?

Your answer: _______

**Question:** The course materials say we should avoid causal language like "smoking causes lower FEV" in observational data. Write one safer sentence using "associated with" and "adjusting for".

Your answer: _______

# PART 4: Read the full three-predictor model

The adjusted smoking model is a **three-predictor** multiple regression. All three slopes are read while holding the other two predictors constant.

```{r full-model, eval=FALSE}
full_model <- lm(fev_l ~ smoke + age_yrs + height_in, data = fev)
summary(full_model)
```

**Question:** In `summary(full_model)`, which predictor has the largest absolute slope? Does that automatically mean it is the most important predictor in real life? (Hint: think about units.)

Your answer: _______

**Question:** What is the adjusted R-squared for this model?

Your answer: _______

# PART 5: Prediction with multiple predictors

With several predictors, `predict()` needs a row that supplies **every** predictor in the model.

The example below predicts FEV for a 12-year-old non-smoker who is 60 inches tall.

```{r predict-example}
new_person <- tibble(
  smoke = factor("non_smoker", levels = c("non_smoker", "smoker")),
  age_yrs = 12,
  height_in = 60
)

predict(full_model, newdata = new_person)
```

Now try two predictions of your own.

```{r predict-smoker, eval=FALSE}
smoker_12_60 <- tibble(
  smoke = factor("smoker", levels = c("non_smoker", "smoker")),
  age_yrs = ________,
  height_in = ________
)

predict(________, newdata = ________)
```

**Question:** For the same age and height, is predicted FEV higher or lower for the smoker than the non-smoker in the example above?

Your answer: _______

```{r predict-tall, eval=FALSE}
tall_non_smoker <- tibble(
  smoke = factor("non_smoker", levels = c("non_smoker", "smoker")),
  age_yrs = ________,
  height_in = ________
)

predict(________, newdata = ________)
```

**Question:** Predict FEV for a 16-year-old non-smoker who is 70 inches tall. Does the prediction seem plausible for this dataset?

Your answer: _______

# PART 6: R-squared across models

Finally, compare how much variation in FEV each model explains.

```{r r-squared-compare, eval=FALSE}
tibble(
  model = c("Age only", "Height only", "Age + height", "Smoke + age + height"),
  r_squared = c(
    summary(________)$r.squared,
    summary(________)$r.squared,
    summary(________)$r.squared,
    summary(________)$r.squared
  )
)
```

**Question:** What happens to R-squared when you add more predictors? Does a higher R-squared always mean the model is "better" for every purpose?

Your answer: _______

# Wrap-up

In this workshop you:

1. Fitted models with **two and three predictors** using `+` in the formula.
2. Interpreted coefficients as **adjusted** slopes (holding other predictors constant).
3. Saw how a misleading raw association (smoking) can change after adjustment.
4. Used `predict()` and compared **R-squared** across models.

Next up in the course: **interaction terms**, where the slope of one predictor can depend on another.
