---
title: "SOLUTIONS: Replicating a published multiple regression table"
subtitle: "COVID-19 psychological distress (K6): reproducing and critiquing Table 4"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

These are the completed solutions for the week 8 exercise. Each step first shows the blanked chunk from the exercise (not run), then the filled-in **Solution** chunk that runs.

In week 5 we replicated Table 2 of the COVID-19 psychological distress paper: group means, t-tests, and ANOVAs. This week we return to the same paper and data to replicate its **Table 4: the stepwise multiple linear regression** for psychological distress. Along the way we will:

1. Reproduce the paper's regression coefficients exactly, from the raw data.
2. Discover a typo in the published table.
3. Critique the model itself. Reproducing a published analysis and *agreeing with it* are two different things, and this table has real problems worth understanding.

## The study

- Qiu et al.-style survey by Wang et al. (2020), *The psychological distress and coping styles in the early stages of the COVID-19 epidemic in the general population of China*, PLOS ONE. A copy is in `paper/distress_paper.pdf`.
- The survey data (n = 1599) is in `data/distress_data.xlsx`; the codebook is in `data/distress_codebook.xls`.
- The outcome is the **K6 psychological distress score** (`total_scores_of_K6`), the sum of six items each scored 0 to 4, so the total runs 0 to 24. Higher means more distress.

## The table we will build

Table 4 of the paper reports a stepwise multiple linear regression with these results: R² = 0.862, and coefficients for Marriage (β = -0.588), History of visiting Wuhan (β = 1.449), Changes over living situations (β = 1.040), Emotional control (β = 1.995), Epidemic-related dreams (β = 1.597), and Coping style (β = -2.135), all p < 0.001.

Table 3 of the paper tells us how each variable was coded. Three things to notice:

- **Marriage**: 0 = Unmarried, 1 = Married. **Coping style**: 0 = Negative, 1 = Positive. **Wuhan / community history**: 0 = No, 1 = Yes.
- Everything else, including the ordered survey responses (media concern, living-situation changes, emotional control, epidemic-related dreams), was entered at its **"Primary Value"**: the raw 0/1/2/3/4 code, treated as a plain number. This is a modeling choice worth questioning, and we will.
- "Stepwise" means the authors started from a larger set of candidate predictors and let an algorithm drop the weak ones. We will reproduce that too.

# PART 1: Rebuild the analysis dataset

## Load packages

Use `p_load()` to load: readxl, tidyverse, here, janitor, broom, gt.

*Exercise chunk (blank):*

```{r load-packages-q, eval=FALSE}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

**Solution:**

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(readxl, tidyverse, here, janitor, broom, gt)
```

## Step 1: Import and clean names

*Exercise chunk (blank):*

```{r import-q, eval=FALSE}
survey_raw <- read_excel(here("data/distress_data.xlsx")) %>%
  ______()  # clean the messy column names

glimpse(survey_raw)
```

**Solution:**

```{r import}
survey_raw <- read_excel(here("data/distress_data.xlsx")) %>%
  clean_names()

glimpse(survey_raw)
```

**Checkpoint:** 1599 rows, 38 columns, and all names now lowercase with underscores.

## Step 2: Select and recode the model variables

Table 3 of the paper lists the nine candidate predictors. We select them (plus the outcome) and give them shorter names.

One subtlety: the `marriage` column in the raw data has **three** values:

```{r marriage-check}
tabyl(survey_raw, marriage)
```

61 people have code 2 (the codebook shows this is "other", e.g. divorced or widowed). But Table 3 of the paper says marriage was coded "0 = Unmarried, 1 = Married" with no third category. To match the paper's numbers, the 2s must be folded in with the 0s: married (1) versus everyone else (0). The paper never mentions doing this, which is our first small reproducibility lesson: published coding descriptions are often incomplete, and you sometimes have to reverse-engineer what was actually done.

*Exercise chunk (blanks):*

```{r recode-q, eval=FALSE}
survey <- survey_raw %>%
  transmute(
    k6 = total_scores_of_k6,
    age = ______,
    married = if_else(marriage == ______, 1, 0),
    visited_wuhan = history_of_visiting_sea,
    epidemic_in_community = history_of_epidemics_occurring_in_the_community,
    media_concern = concern_with_media_reports_related_to_the_epidemic,
    living_changes = changes_over_living_situations,
    emotional_control = ______,
    epidemic_dreams = epidemic_related_dreams,
    positive_coping = coping_style
  )

glimpse(survey)
```

**Solution:**

```{r recode}
survey <- survey_raw %>%
  transmute(
    k6 = total_scores_of_k6,
    age = age,
    married = if_else(marriage == 1, 1, 0),
    visited_wuhan = history_of_visiting_sea,
    epidemic_in_community = history_of_epidemics_occurring_in_the_community,
    media_concern = concern_with_media_reports_related_to_the_epidemic,
    living_changes = changes_over_living_situations,
    emotional_control = emotional_control,
    epidemic_dreams = epidemic_related_dreams,
    positive_coping = coping_style
  )

glimpse(survey)
```

**Checkpoint:** 1599 rows and 10 columns, all numeric.

**Question**: In week 5 we converted these coded variables to labelled factors before analysis. Why are we deliberately leaving them numeric this time?

**Answer:** Because we are replicating the paper's model, and Table 3 says the authors entered these variables at their "Primary Value", i.e. as plain numbers. If we converted, say, `emotional_control` to a five-level factor, `lm()` would estimate four separate coefficients for it instead of the single slope the paper reports, and our numbers would not match. (Whether treating an ordered response scale as a number is a *good* idea is a separate question; see Part 3.)

# PART 2: From naive model to Table 4

## Step 3: A naive one-predictor model

Before the full model, fit a simple regression of K6 on `visited_wuhan` alone.

*Exercise chunk (blanks):*

```{r naive-q, eval=FALSE}
wuhan_naive <- lm(______ ~ ______, data = survey)
tidy(wuhan_naive, conf.int = TRUE)
```

**Solution:**

```{r naive}
wuhan_naive <- lm(k6 ~ visited_wuhan, data = survey)
tidy(wuhan_naive, conf.int = TRUE)
```

**Question**: What is the coefficient for `visited_wuhan` in this naive model? Sanity-check it against week 5, where we found mean K6 of 18.6 for visitors and 5.0 for non-visitors.

**Answer:** About **13.6** points (13.607). This is exactly the difference in group means, 18.6 - 5.0 ≈ 13.6, as we learned: a simple regression on a binary predictor *is* the two-group mean comparison.

## Step 4: The full multiple regression

Now fit the model with all nine candidate predictors from Table 3.

*Exercise chunk (blanks):*

```{r full-q, eval=FALSE}
full_model <- lm(
  k6 ~ age + married + visited_wuhan + epidemic_in_community +
    media_concern + living_changes + emotional_control +
    epidemic_dreams + positive_coping,
  data = ______
)

summary(______)
```

**Solution:**

```{r full}
full_model <- lm(
  k6 ~ age + married + visited_wuhan + epidemic_in_community +
    media_concern + living_changes + emotional_control +
    epidemic_dreams + positive_coping,
  data = survey
)

summary(full_model)
```

**Question**: The `visited_wuhan` coefficient collapsed from 13.6 in the naive model to about 1.4 here. Interpret the adjusted coefficient in one sentence, and explain in another sentence why it shrank so dramatically.

**Answer:** Holding all the other predictors constant, people who had visited Wuhan score about **1.4 points higher** on the K6 than those who had not. It shrank because visiting Wuhan is strongly associated with the other predictors in the model (visitors also reported far more emotional-control difficulty, epidemic dreams, and negative coping), so most of the raw 13.6-point gap is accounted for by those variables.

**Question**: Which predictors are *not* statistically significant at alpha = 0.05 in the full model?

**Answer:** `age` (p ≈ 0.29), `epidemic_in_community` (p ≈ 0.25), and `media_concern` (p ≈ 0.46).

## Step 5: Stepwise selection

The paper used **stepwise** regression: an algorithm that adds/drops predictors to optimize a fit criterion. R's `step()` function does backward stepwise selection by AIC.

*Exercise chunk (blank):*

```{r step-q, eval=FALSE}
step_model <- step(______, trace = 0)
summary(step_model)
```

**Solution:**

```{r step}
step_model <- step(full_model, trace = 0)
summary(step_model)
```

**Question**: Which predictors did `step()` drop? Does the final model match the list of variables in the paper's Table 4?

**Answer:** It dropped exactly the three non-significant predictors: `age`, `epidemic_in_community`, and `media_concern`. The six survivors (married, visited Wuhan, living changes, emotional control, epidemic dreams, coping style) are precisely the rows of Table 4.

## Step 6: Compare against Table 4, and find the typo

Build a tidy comparison of your coefficients and CIs.

*Exercise chunk (blank):*

```{r tidy-step-q, eval=FALSE}
tidy(step_model, conf.int = ______) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

**Solution:**

```{r tidy-step}
tidy(step_model, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

For reference, the paper's Table 4:

| Variable | β | t | p | 95% CI |
|---|---|---|---|---|
| Constant | 3.526 | 13.939 | <0.001 | 3.030 – 4.022 |
| Marriage | -0.588 | -3.682 | <0.001 | -0.901 – -0.275 |
| History of visiting Wuhan | 1.449 | 5.265 | <0.001 | 0.909 – 1.898 |
| Changes over living situations | 1.040 | 9.057 | <0.001 | 0.815 – 1.266 |
| Emotional control | 1.995 | 21.879 | <0.001 | 1.816 – 2.174 |
| Epidemic-related dreams | 1.597 | 16.964 | <0.001 | 1.412 – 1.782 |
| Coping style | -2.135 | -10.575 | <0.001 | -2.531 – -1.739 |

R² = 0.862, adjusted R² = 0.862.

```{r r-squared}
summary(step_model)$r.squared
summary(step_model)$adj.r.squared
```

**Question**: Go row by row. Do your estimates, t values, and CIs match the paper? Is there any cell that does not match? (Hint: a 95% CI is always symmetric around the estimate. Check each row for that.)

**Answer:** Everything matches to three decimals, with **one exception**: the upper CI bound for history of visiting Wuhan. Our analysis gives 0.909 to **1.989**; the paper prints 0.909 to **1.898**. Since β = 1.449 and the CI must be symmetric around it (1.449 - 0.909 = 0.540, and 1.449 + 0.540 = 1.989), the paper's own numbers are internally inconsistent: 1.898 is 1.989 with the last two digits swapped. Almost certainly a transcription typo. As in week 7, a fresh look at released data caught an error that peer review missed.

**Question**: Interpret the `positive_coping` coefficient (-2.135) in one sentence, remembering it is coded 0 = negative coping, 1 = positive coping.

**Answer:** Holding the other five predictors constant, people with a positive coping style score about **2.1 points lower** on the K6 distress scale than people with a negative coping style, on average.

## Step 7: Assemble a paper-style table with gt

The formatting code is written for you; your job is to check the values against `tidy(step_model)`.

```{r gt-table}
table4_rows <- tidy(step_model, conf.int = TRUE) %>%
  mutate(
    Variable = case_when(
      term == "(Intercept)" ~ "Constant",
      term == "married" ~ "Marriage (married vs other)",
      term == "visited_wuhan" ~ "History of visiting Wuhan (yes vs no)",
      term == "living_changes" ~ "Changes over living situations (per level)",
      term == "emotional_control" ~ "Emotional control difficulty (per level)",
      term == "epidemic_dreams" ~ "Epidemic-related dreams (per level)",
      term == "positive_coping" ~ "Coping style (positive vs negative)"
    ),
    `Beta (95% CI)` = paste0(
      format(round(estimate, 3), nsmall = 3), " (",
      format(round(conf.low, 3), nsmall = 3), " to ",
      format(round(conf.high, 3), nsmall = 3), ")"
    ),
    t = round(statistic, 2),
    p = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))
  ) %>%
  select(Variable, `Beta (95% CI)`, t, p)

table4_rows %>%
  gt() %>%
  tab_header(
    title = "Multiple linear regression for psychological distress (K6), n = 1599",
    subtitle = "Replication of Wang et al. (2020), Table 4 (with corrected Wuhan CI)"
  ) %>%
  tab_options(table.width = pct(100))
```

# PART 3: Should we believe this model?

Reproducing the numbers is not the same as endorsing the analysis. This table has at least three problems worth understanding. This kind of critical reading matters: this model, with its impressive-sounding R² of 0.862, was published and has been cited many times.

## Problem 1: Predictors that are almost the outcome

Look again at the two strongest predictors: **emotional control** ("how difficult has it been to control your emotions?") and **epidemic-related dreams**. Now recall what the K6 asks: how often did you feel *nervous*, *hopeless*, *restless*, *so depressed nothing could cheer you up*, *that everything was an effort*, *worthless*.

```{r symptom-cors}
survey %>%
  summarise(
    cor_emotion_k6 = cor(emotional_control, k6),
    cor_dreams_k6 = cor(epidemic_dreams, k6)
  )
```

**Question**: These two "predictors" correlate with the outcome at about 0.89 and 0.87. Why is it not impressive that they dominate the model? Is this best described as reverse causation, or something even more circular?

**Answer:** Difficulty controlling emotions and recurrent distressing dreams are not upstream *causes* of distress that happen to predict it; they are essentially **symptoms of distress itself**, measured in the same questionnaire at the same time. At best this is reverse causation (being distressed makes emotional control hard and produces bad dreams); more precisely it is near-circularity: the model largely "predicts" distress from other measurements of distress. That is why the R² is a spectacular 0.862, a value you should treat as a red flag, not a triumph, in a cross-sectional survey of psychological states. The model tells us little about what *causes* distress or where to intervene.

Let's quantify how much of the model's fit comes from these two symptom-like variables, by refitting without them.

*Exercise chunk (blank):*

```{r no-symptoms-q, eval=FALSE}
upstream_model <- lm(
  k6 ~ age + married + visited_wuhan + epidemic_in_community +
    media_concern + living_changes + positive_coping,
  data = survey
)

summary(______)$r.squared
```

**Solution:**

```{r no-symptoms}
upstream_model <- lm(
  k6 ~ age + married + visited_wuhan + epidemic_in_community +
    media_concern + living_changes + positive_coping,
  data = survey
)

summary(upstream_model)$r.squared
tidy(upstream_model, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

**Question**: What happens to the R² and to the `visited_wuhan` coefficient when the two symptom-like predictors are removed? Which model do you find more useful for public health purposes?

**Answer:** R² drops from 0.862 to about **0.71**, and the `visited_wuhan` coefficient jumps from 1.4 back up to about **7.8** points. Neither model is "the truth", but the second is arguably more useful for public health: it asks how distress relates to circumstances and exposures (visiting Wuhan, living changes, coping style) without letting distress symptoms explain away other distress symptoms. Note that adjusting for a consequence of the outcome, as Table 4 effectively does, can badly distort the coefficients of the genuine upstream predictors.

## Problem 2: Ordered response scales treated as numbers

Table 3's "Primary Value" coding means, for example, that `emotional_control` (0 = no difficulty ... 4 = extremely difficult) enters as a plain number, so the model assumes moving from "no difficulty" to "less difficult" has exactly the same effect on K6 as moving from "more difficult" to "extremely difficult".

**Question**: What would we have done in week 7 with a variable like this, and what is the trade-off?

**Answer:** In week 7 we would have converted it to a **factor**, giving each level its own coefficient relative to a reference level. That avoids the equal-spacing assumption but costs more parameters (four coefficients instead of one) and no longer yields a single tidy "per level" slope. Treating an ordinal scale as numeric is a common and sometimes defensible shortcut, but it is an assumption the paper never states or checks. (You can check it yourself: `group_by(emotional_control) %>% summarise(mean(k6))` shows the jump from level 3 to 4 is far larger than the earlier jumps.)

## Problem 3: Stepwise selection itself

**Question**: Recall the p-hacking discussion from this week's interaction lesson. What is the concern with running stepwise selection and then reporting the survivors' p-values as if they were ordinary?

**Answer:** The p-values and CIs in Table 4 are computed as though the six-predictor model had been specified in advance. In fact the data were used twice: once to choose the model, then again to test it. Selection makes the surviving coefficients look stronger and their p-values smaller than they honestly are. With n = 1599 and effects this large the distortion here is mild, but the practice, letting an algorithm pick predictors and reporting uncorrected inference, is widely discouraged for exactly this reason. Predictors should ideally be chosen based on the research question and subject-matter knowledge.

# Summary

We reproduced the paper's Table 4 exactly from the raw data, down to R² = 0.862, and found one typo (the Wuhan CI upper bound should be 1.989, not 1.898). We then stepped back from replication to appraisal: the model's headline fit is inflated by two predictors that are effectively symptoms of the outcome (dropping them cuts R² to 0.71 and changes the Wuhan coefficient five-fold), it treats ordered response scales as equally-spaced numbers, and its stepwise-selected p-values are overconfident. Being able to reproduce a published table *and* articulate why you would not have built it that way is exactly the skill this course is aiming for.
