---
title: "EXERCISE: Replicating a published multiple regression table"
subtitle: "COVID-19 psychological distress (K6): reproducing and critiquing Table 4"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In week 5 we replicated Table 2 of the COVID-19 psychological distress paper: group means, t-tests, and ANOVAs. This week we return to the same paper and data to replicate its **Table 4: the stepwise multiple linear regression** for psychological distress. Along the way we will:

1. Reproduce the paper's regression coefficients exactly, from the raw data.
2. Discover a typo in the published table.
3. Critique the model itself. Reproducing a published analysis and *agreeing with it* are two different things, and this table has real problems worth understanding.

## The study

- Wang et al. (2020), *The psychological distress and coping styles in the early stages of the COVID-19 epidemic in the general population of China*, PLOS ONE. A copy is in `paper/distress_paper.pdf`.
- The survey data (n = 1599) is in `data/distress_data.xlsx`; the codebook is in `data/distress_codebook.xls`.
- The outcome is the **K6 psychological distress score** (`total_scores_of_K6`), the sum of six items each scored 0 to 4, so the total runs 0 to 24. Higher means more distress.

## The table we will build

Table 4 of the paper reports a stepwise multiple linear regression with these results: R² = 0.862, and coefficients for Marriage (β = -0.588), History of visiting Wuhan (β = 1.449), Changes over living situations (β = 1.040), Emotional control (β = 1.995), Epidemic-related dreams (β = 1.597), and Coping style (β = -2.135), all p < 0.001.

Table 3 of the paper tells us how each variable was coded. Three things to notice:

- **Marriage**: 0 = Unmarried, 1 = Married. **Coping style**: 0 = Negative, 1 = Positive. **Wuhan / community history**: 0 = No, 1 = Yes.
- Everything else, including the ordered survey responses (media concern, living-situation changes, emotional control, epidemic-related dreams), was entered at its **"Primary Value"**: the raw 0/1/2/3/4 code, treated as a plain number. This is a modeling choice worth questioning, and we will.
- "Stepwise" means the authors started from a larger set of candidate predictors and let an algorithm drop the weak ones. We will reproduce that too.

# PART 1: Rebuild the analysis dataset

## Load packages

Use `p_load()` to load: readxl, tidyverse, here, janitor, broom, gt.

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

## Step 1: Import and clean names

```{r import, eval=FALSE}
survey_raw <- read_excel(here("data/distress_data.xlsx")) %>%
  ______()  # clean the messy column names

glimpse(survey_raw)
```

**Checkpoint:** 1599 rows, 38 columns, and all names now lowercase with underscores.

## Step 2: Select and recode the model variables

Table 3 of the paper lists the nine candidate predictors. We select them (plus the outcome) and give them shorter names.

One subtlety: the `marriage` column in the raw data has **three** values:

```{r marriage-check, eval=FALSE}
tabyl(survey_raw, marriage)
```

61 people have code 2 (the codebook shows this is "other", e.g. divorced or widowed). But Table 3 of the paper says marriage was coded "0 = Unmarried, 1 = Married" with no third category. To match the paper's numbers, the 2s must be folded in with the 0s: married (1) versus everyone else (0). The paper never mentions doing this, which is our first small reproducibility lesson: published coding descriptions are often incomplete, and you sometimes have to reverse-engineer what was actually done.

```{r recode, eval=FALSE}
survey <- survey_raw %>%
  transmute(
    k6 = total_scores_of_k6,
    age = ______,
    married = if_else(marriage == ______, 1, 0),
    visited_wuhan = history_of_visiting_sea,
    epidemic_in_community = history_of_epidemics_occurring_in_the_community,
    media_concern = concern_with_media_reports_related_to_the_epidemic,
    living_changes = changes_over_living_situations,
    emotional_control = ______,
    epidemic_dreams = epidemic_related_dreams,
    positive_coping = coping_style
  )

glimpse(survey)
```

**Checkpoint:** 1599 rows and 10 columns, all numeric.

**Question**: In week 5 we converted these coded variables to labelled factors before analysis. Why are we deliberately leaving them numeric this time?

Your answer:

# PART 2: From naive model to Table 4

## Step 3: A naive one-predictor model

Before the full model, fit a simple regression of K6 on `visited_wuhan` alone.

```{r naive, eval=FALSE}
wuhan_naive <- lm(______ ~ ______, data = survey)
tidy(wuhan_naive, conf.int = TRUE)
```

**Question**: What is the coefficient for `visited_wuhan` in this naive model? Sanity-check it against week 5, where we found mean K6 of 18.6 for visitors and 5.0 for non-visitors.

Your answer:

## Step 4: The full multiple regression

Now fit the model with all nine candidate predictors from Table 3.

```{r full, eval=FALSE}
full_model <- lm(
  k6 ~ age + married + visited_wuhan + epidemic_in_community +
    media_concern + living_changes + emotional_control +
    epidemic_dreams + positive_coping,
  data = ______
)

summary(______)
```

**Question**: The `visited_wuhan` coefficient collapsed from 13.6 in the naive model to about 1.4 here. Interpret the adjusted coefficient in one sentence, and explain in another sentence why it shrank so dramatically.

Your answer:

**Question**: Which predictors are *not* statistically significant at alpha = 0.05 in the full model?

Your answer:

## Step 5: Stepwise selection

The paper used **stepwise** regression: an algorithm that adds/drops predictors to optimize a fit criterion. R's `step()` function does backward stepwise selection by AIC.

```{r step, eval=FALSE}
step_model <- step(______, trace = 0)
summary(step_model)
```

**Question**: Which predictors did `step()` drop? Does the final model match the list of variables in the paper's Table 4?

Your answer:

## Step 6: Compare against Table 4, and find the typo

Build a tidy comparison of your coefficients and CIs.

```{r tidy-step, eval=FALSE}
tidy(step_model, conf.int = ______) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

For reference, the paper's Table 4:

| Variable | β | t | p | 95% CI |
|---|---|---|---|---|
| Constant | 3.526 | 13.939 | <0.001 | 3.030 – 4.022 |
| Marriage | -0.588 | -3.682 | <0.001 | -0.901 – -0.275 |
| History of visiting Wuhan | 1.449 | 5.265 | <0.001 | 0.909 – 1.898 |
| Changes over living situations | 1.040 | 9.057 | <0.001 | 0.815 – 1.266 |
| Emotional control | 1.995 | 21.879 | <0.001 | 1.816 – 2.174 |
| Epidemic-related dreams | 1.597 | 16.964 | <0.001 | 1.412 – 1.782 |
| Coping style | -2.135 | -10.575 | <0.001 | -2.531 – -1.739 |

R² = 0.862, adjusted R² = 0.862.

```{r r-squared, eval=FALSE}
summary(step_model)$r.squared
summary(step_model)$adj.r.squared
```

**Question**: Go row by row. Do your estimates, t values, and CIs match the paper? Is there any cell that does not match? (Hint: a 95% CI is always symmetric around the estimate. Check each row for that.)

Your answer:

**Question**: Interpret the `positive_coping` coefficient (-2.135) in one sentence, remembering it is coded 0 = negative coping, 1 = positive coping.

Your answer:

## Step 7: Assemble a paper-style table with gt

The formatting code is written for you; your job is to check the values against `tidy(step_model)`.

```{r gt-table, eval=FALSE}
table4_rows <- tidy(step_model, conf.int = TRUE) %>%
  mutate(
    Variable = case_when(
      term == "(Intercept)" ~ "Constant",
      term == "married" ~ "Marriage (married vs other)",
      term == "visited_wuhan" ~ "History of visiting Wuhan (yes vs no)",
      term == "living_changes" ~ "Changes over living situations (per level)",
      term == "emotional_control" ~ "Emotional control difficulty (per level)",
      term == "epidemic_dreams" ~ "Epidemic-related dreams (per level)",
      term == "positive_coping" ~ "Coping style (positive vs negative)"
    ),
    `Beta (95% CI)` = paste0(
      format(round(estimate, 3), nsmall = 3), " (",
      format(round(conf.low, 3), nsmall = 3), " to ",
      format(round(conf.high, 3), nsmall = 3), ")"
    ),
    t = round(statistic, 2),
    p = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))
  ) %>%
  select(Variable, `Beta (95% CI)`, t, p)

table4_rows %>%
  gt() %>%
  tab_header(
    title = "Multiple linear regression for psychological distress (K6), n = 1599",
    subtitle = "Replication of Wang et al. (2020), Table 4"
  ) %>%
  tab_options(table.width = pct(100))
```

# PART 3: Should we believe this model?

Reproducing the numbers is not the same as endorsing the analysis. This table has at least three problems worth understanding. This kind of critical reading matters: this model, with its impressive-sounding R² of 0.862, was published and has been cited many times.

## Problem 1: Predictors that are almost the outcome

Look again at the two strongest predictors: **emotional control** ("how difficult has it been to control your emotions?") and **epidemic-related dreams**. Now recall what the K6 asks: how often did you feel *nervous*, *hopeless*, *restless*, *so depressed nothing could cheer you up*, *that everything was an effort*, *worthless*.

```{r symptom-cors, eval=FALSE}
survey %>%
  summarise(
    cor_emotion_k6 = cor(emotional_control, k6),
    cor_dreams_k6 = cor(epidemic_dreams, k6)
  )
```

**Question**: These two "predictors" correlate with the outcome at about 0.89 and 0.87. Why is it not impressive that they dominate the model? Is this best described as reverse causation, or something even more circular?

Your answer:

Let's quantify how much of the model's fit comes from these two symptom-like variables, by refitting without them.

```{r no-symptoms, eval=FALSE}
upstream_model <- lm(
  k6 ~ age + married + visited_wuhan + epidemic_in_community +
    media_concern + living_changes + positive_coping,
  data = survey
)

summary(______)$r.squared
```

**Question**: What happens to the R² and to the `visited_wuhan` coefficient when the two symptom-like predictors are removed? Which model do you find more useful for public health purposes?

Your answer:

## Problem 2: Ordered response scales treated as numbers

Table 3's "Primary Value" coding means, for example, that `emotional_control` (0 = no difficulty ... 4 = extremely difficult) enters as a plain number, so the model assumes moving from "no difficulty" to "less difficult" has exactly the same effect on K6 as moving from "more difficult" to "extremely difficult".

**Question**: What would we have done in week 7 with a variable like this, and what is the trade-off?

Your answer:

## Problem 3: Stepwise selection itself

**Question**: Recall the p-hacking discussion from this week's interaction lesson. What is the concern with running stepwise selection and then reporting the survivors' p-values as if they were ordinary?

Your answer:

# Summary

Write two or three sentences: what did you reproduce, what did you find in the published table, and what is your overall assessment of the model?

Your answer:
