---
title: "Workshop 6: What Regression Is For"
subtitle: "FEV lung function, age, height, and smoking"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this workshop you will use simple linear regression to study lung function in children and adolescents.

The dataset contains 654 youths aged 3 to 19 from East Boston in the 1970s. The outcome is **forced expiratory volume** (`fev_l`), measured in litres. FEV measures how much air a person can forcefully breathe out in one second. Higher values mean better lung function.

The original teaching purpose of this dataset was to study smoking and FEV. The smoking variable records whether the youth was a **current smoker** or **non-smoker** in the dataset documentation.

This workshop is organized around three uses of regression:

1. **Describe** a relationship between two continuous variables.
2. **Predict** the outcome for a new value of the predictor.
3. **Control** for another variable, in a small preview of multiple regression.

# Load packages and data

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, broom)
```

```{r load-data}
fev <- read_csv("/Users/kendavidn/Dropbox/tgc_github_projects/rsr_2026_t2_materials/rsr_week_06_workshop_gpt5_q7b2/fev_single_file/data/fev_kahn.csv")
glimpse(fev)
```

**Checkpoint:** `fev` should have **654 rows and 5 columns**.

The variables we will use are:

- `fev_l`: forced expiratory volume in litres
- `age_yrs`: age in years
- `height_in`: height in inches
- `smoke`: current smoking status, `non_smoker` or `smoker`

# Demo:: Describe a relationship with age

Here we will... do a demo something something just focus on running the code and understanting the outputs..

## 1A. Start with a scatterplot

```{r age-scatter}
ggplot(fev, aes(x = age_yrs, y = fev_l)) +
  geom_point(alpha = 0.45) +
  labs(
    title = "FEV by age",
    x = "Age (years)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

Question: Is there a positive or negative relationship...

## 1B. Correlation

We can use correlation to...
Correlation summarizes the direction and strength of a linear relationship.

Here is a table showing strength....

Now let's run...

```{r age-correlation}
r_age_fev <- cor(fev$age_yrs, fev$fev_l)
r_age_fev
```

The correlation is about **0.756**, a strong positive relationship.

Correlation is useful but lacks some nice features of regression...

question: fill in the blanks in the table below that distinguishes

## 1C. Fit a simple linear regression

Regression gives us a fitted line:

$$\widehat{y} = b_0 + b_1x$$ (for model: avoid latex just use regular text if possible. same below)

```{r age-model}
age_model <- lm(fev_l ~ age_yrs, data = fev)
summary(age_model)
```

The estimate column gives....

The fitted equation is approximately:

$$\widehat{FEV} = 0.432 + 0.222 \times age$$

Interpretation of the slope:

> For each additional 1 year of age, predicted FEV is higher by about 0.222 litres, on average.

The p value tells us....

question: what does the intercept correspond to? an age of how many years...

---

Next let's extract those numbers from the model summary and plot a line with them on top of the ggplot...

code chunk here

we can also do this directly with the geom smooth and lm...

code chunk here

## 1D. R-squared

At the bottom of the summary you will see..

```{r age-glance}

```

The R-squared is about **0.572**.

Interpretation:

> About 57.2% of the variation in FEV is explained by age in this simple linear regression.

In simple linear regression with one predictor, R-squared is equal to the correlation squared:

```{r age-r-squared-from-correlation}
r_age_fev^2
```

R-squared is useful because it connects the relationship to the fitted model and the outcome: it tells us how much of the variation in FEV this one-predictor model explains.

question: if the correlation coefficient is 1, then what percent of variation would be explained by linear model

# Part 2: Fit a better predictor: height

Now it is your turn to write the code and interpret the output. 

Instead of trying to predict lung function with age, we will try to do it with height. 
 FEV better than age.

## 2A. Plot FEV by height

Make a scatterplot with height on the x-axis and FEV on the y-axis. You can inspire yourself based on the code anove..

```{r height-scatter, eval=FALSE}
# YOUR CODE HERE
```

**question.** Does the relationship look positive, negative, or close to zero?

Ans:


## 2B. Correlation

Next, Calculate the correlation between height and FEV using the cor function

```{r height-correlation, eval=FALSE}

```

**Q2b.** Fill in the blanks:

The correlation is about ______. This relationship is ______ (positive / negative) and ______ (weak / moderate / strong).

Ans:

## 2C. Fit the height model

Fit a simple linear regression with `fev_l` as the outcome and `height_in` as the predictor.

```{r height-model, eval=FALSE}
height_model <- lm(_____________________data = fev)
summary(height_model)
```

**question.** Fill in this equation : 

height_fev = 

**Q2d.** The height slope is about **0.132**. Interpret this slope in one sentence.

Ans:

**Q2e.** What is the p-value for `height_in`? At alpha = 0.05, do we reject or fail to reject the null hypothesis that the true slope is 0?

Ans:

## 2D. R-squared and residual standard error

Again, loooking at the...

**Q2f.** The height model has R-squared about **0.754**. Fill in the blanks:

About ______% of the variation in FEV is explained by ______ in this simple linear regression.

Ans:

**Q2g.** Compare the age and height models:

- Age model R-squared: about 0.572
- Height model R-squared: about 0.754

Which predictor gives the better one-variable model for FEV, age or height? How do you know?

Ans:

# Part 3: Predicting with regression or something like that

Another useful something of regression is it lets us ... predict the outcome for a new value of the predictor.

For example, if we want to .,

```{r height-predict, eval=FALSE}
new_youth <- tibble(height_in = 60)
new_youth
```

Now we can...

```{r}
[hey assistant, this part should be filled in cuz they don't know how to do it yet. predict with the age model then it]
predict(__________, newdata = __________)
```

then they predict with the height model for some regular height..



then they predict with the height model for a baby... should probably give something negative and we ask them if this makes sense

Now

# Part 3: controlling...

This final part is a quick preview of a third use of regression: **control**. We have not yet taught you  multiple regression yet. 

The goal is to see why a raw comparison can be misleading in observational data.

The original dataset documentation says `smoke` records current smoking status. First, summarize age, height, and FEV by smoking status.

The dataset records each child's **smoking status** (`smoke`): whether the child or teenager is themselves a smoker. (This was back in the XX's when lots of kids smoked behind their parents backs) Let's try to ask a public-health question: is smoking associated with lung function?

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    n = n(),
    mean_age = mean(age_yrs),
    mean_fev = mean(fev_l),
  )
```

**Q4a.** Which group has higher mean FEV, smokers or non-smokers?

Ans:


The results above may seem surprising but .... let's now seee.

**Q4b.** Which group is older on average?

Ans:

Smokers have higher mean FEV in the raw comparison, but the smokers are much older and taller on average. Older and taller youths naturally have larger lungs. Is there a way we can see ....

This is where multiple regression comes in...

```{r smoke-raw-model}
raw_model <- lm(fev_l ~ smoke, data = fev)
summary(raw_model)
```

The xx line in this output shows the number xxx.. which we interpret as meaning...

The raw smoking model compares mean FEV between smokers and non-smokers. It makes smoking look positively associated with FEV.

But preview a multiple regression model:

```{r smoke-height-model}
controlled_model <- lm(fev_l ~ height_in + smoke, data = fev)
summary(controlled_model)

```

Question... what is the number under xxxx now. this 

**Q4c.** In one or two sentences, explain why the raw smoking comparison was misleading.

Ans:

# Wrap-up

In this workshop, regression helped us do three things:

1. **Describe:** age and height both have positive linear relationships with FEV.
2. **Predict:** the height model can predict FEV for a new height value.
3. **Control:** the smoking example showed why an observational association can change after accounting for another variable.

You will learn multiple regression properly in coming lessons. ...

bla bla...

