---
title: "Workshop 6 Solutions: What Regression Is For"
subtitle: "FEV lung function, age, height, and smoking"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This solutions file is a full worked version of `rsr_week_06_fev_workshop.Rmd`. It keeps the blank student prompts visible, then gives solution code and written answers.

# Load packages and data

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, broom)
```

```{r load-data}
fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint answer:** `fev` has **654 rows and 5 columns**.

# Part 1: Describe a relationship with age

This part was already fully worked in the student file.

```{r age-scatter}
ggplot(fev, aes(x = age_yrs, y = fev_l)) +
  geom_point(alpha = 0.45) +
  labs(
    title = "FEV by age",
    x = "Age (years)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

```{r age-correlation}
r_age_fev <- cor(fev$age_yrs, fev$fev_l)
r_age_fev
```

```{r age-model}
age_model <- lm(fev_l ~ age_yrs, data = fev)
summary(age_model)
tidy(age_model)
glance(age_model)
r_age_fev^2
```

# Part 2: Fit a better predictor: height

Student prompt:

```{r height-scatter-blank, eval=FALSE}
ggplot(fev, aes(x = ____________, y = ____________)) +
  geom_point(alpha = 0.45) +
  labs(
    title = "FEV by height",
    x = "Height (inches)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

Solution:

```{r height-scatter}
ggplot(fev, aes(x = height_in, y = fev_l)) +
  geom_point(alpha = 0.45) +
  labs(
    title = "FEV by height",
    x = "Height (inches)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

**Q2a answer:** The relationship is positive.

```{r height-line}
ggplot(fev, aes(x = height_in, y = fev_l)) +
  geom_point(alpha = 0.45) +
  geom_smooth(method = "lm", se = FALSE, color = "#1f78b4") +
  labs(
    title = "Least-squares line for height and FEV",
    x = "Height (inches)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

Student prompt:

```{r height-correlation-blank, eval=FALSE}
r_height_fev <- cor(fev$__________, fev$__________)
r_height_fev
```

Solution:

```{r height-correlation}
r_height_fev <- cor(fev$height_in, fev$fev_l)
r_height_fev
```

**Q2b answer:** The correlation is about **0.868**. This relationship is **positive** and **strong**.

Student prompt:

```{r height-model-blank, eval=FALSE}
height_model <- lm(__________ ~ __________, data = fev)
summary(height_model)
```

Solution:

```{r height-model}
height_model <- lm(fev_l ~ height_in, data = fev)
summary(height_model)
```

Student prompt:

```{r height-tidy-blank, eval=FALSE}
tidy(__________)
```

Solution:

```{r height-tidy}
tidy(height_model)
```

**Q2c answer:**

$$\widehat{FEV} = -5.43 + 0.132 \times height$$

**Q2d answer:** For each additional 1 inch of height, predicted FEV is higher by about **0.132 litres**, on average.

**Q2e answer:** The p-value for `height_in` is about `1.57e-200`, far below 0.05. We reject the null hypothesis.

Student prompt:

```{r height-glance-blank, eval=FALSE}
glance(__________)
```

Solution:

```{r height-glance}
glance(height_model)
```

**Q2f answer:** About **75.4%** of the variation in FEV is explained by **height** in this simple linear regression.

**Q2g answer:** Height gives the better one-variable model. Its R-squared is about 0.754, higher than the age model's R-squared of about 0.572.

# Part 3: Predict with the height model

Student prompt:

```{r height-predict-blank, eval=FALSE}
new_youth <- tibble(height_in = 60)
predict(__________, newdata = __________)
```

Solution:

```{r height-predict}
new_youth <- tibble(height_in = 60)
predict(height_model, newdata = new_youth)
```

**Q3a answer:** The predicted FEV for a youth who is 60 inches tall is about **2.49 litres**.

Student prompt:

```{r height-augment-blank, eval=FALSE}
fev_with_height_model <- augment(height_model, data = fev)

fev_with_height_model %>%
  select(height_in, fev_l, .fitted, .resid) %>%
  head()
```

Solution:

```{r height-augment}
fev_with_height_model <- augment(height_model, data = fev)

fev_with_height_model %>%
  select(height_in, fev_l, .fitted, .resid) %>%
  head()
```

**Q3b answer:** A positive residual means the observed FEV is above the fitted line.

# Part 4: Smoking, age, and why control can matter

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    n = n(),
    mean_age = mean(age_yrs),
    mean_height = mean(height_in),
    mean_fev = mean(fev_l),
    .groups = "drop"
  )
```

**Q4a answer:** Smokers have higher mean FEV in the raw summary.

**Q4b answer:** Smokers are older on average.

```{r smoke-raw-model}
smoke_raw_model <- lm(fev_l ~ smoke, data = fev)
tidy(smoke_raw_model)
glance(smoke_raw_model)
```

```{r smoke-height-model}
smoke_height_model <- lm(fev_l ~ height_in + smoke, data = fev)
tidy(smoke_height_model)
glance(smoke_height_model)
```

**Q4c answer:** The raw comparison was misleading because smokers in this dataset are much older and taller than non-smokers, and older/taller youths naturally have higher FEV. After controlling for height, the smoking coefficient is close to 0.
