---
title: "DEMO: A four-step recipe for a publication table"
subtitle: "PREVEND systolic blood pressure by participant characteristic"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This demo gives you the recipe you will use in the exercise.

We will ask whether systolic blood pressure differs by sex, age group, education, and statin use in the PREVEND teaching dataset.

The same four steps appear in the exercise:

1. Import and label the data.
2. Calculate group summaries.
3. Run the appropriate tests.
4. Assemble a publication-style table.

# Load packages and data

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, gt)
pacman::p_load(openintro)
```

```{r load-data}
prevend_raw <- openintro::prevend.samp

glimpse(prevend_raw)
```

These are the raw variables we will use as predictors of systolic blood pressure:

```{r predictor-preview}
prevend_raw %>%
  select(Gender, Age, Education, Statin) %>%
  summary()
```

In this teaching dataset, `Statin` is a clean binary variable with no missing values.

# Step 1: Import and label the data

The PREVEND data are already imported. We still need to select the variables, give them readable names, and label the categorical variables.

```{r step-1}
prevend <- prevend_raw %>%
  transmute(
    systolic_bp = SBP,
    age_yrs = Age,
    sex = factor(
      Gender,
      levels = c(0, 1),
      labels = c("Male", "Female")
    ),
    education = factor(
      Education,
      levels = c(0, 1, 2, 3),
      labels = c(
        "Primary school",
        "Lower secondary",
        "Higher secondary",
        "University"
      )
    ),
    statin_use = factor(
      Statin,
      levels = c(0, 1),
      labels = c("No statin", "Statin")
    )
  ) %>%
  mutate(
    age_group = cut(
      age_yrs,
      breaks = c(35, 50, 65, Inf),
      labels = c("36-50", "51-65", "66+")
    )
  )

glimpse(prevend)
```

# Step 2: Descriptive stats for every group

We first reshape the data to long form. This creates one column for the characteristic name and one column for the group label.

```{r step-2-long}
prevend_long <- prevend %>%
  select(
    systolic_bp,
    Sex = sex,
    `Age group` = age_group,
    Education = education,
    `Statin use` = statin_use
  ) %>%
  pivot_longer(
    cols = -systolic_bp,
    names_to = "variable",
    values_to = "group"
  ) %>%
  filter(!is.na(group), !is.na(systolic_bp))

prevend_long
```

Now we calculate the sample size, mean, standard deviation, standard error, and 95% confidence interval for each group.

For example, this code counts the number of people in each group.

```{r step-2-count-example}
prevend_long %>%
  group_by(variable, group) %>%
  summarise(
    n = n()
  )
```

```{r step-2-summary}
group_stats <- prevend_long %>%
  group_by(variable, group) %>%
  summarise(
    n = n(),
    mean = mean(systolic_bp),
    sd = sd(systolic_bp),
    se = sd / sqrt(n),
    ci_lo = mean - qt(0.975, df = n - 1) * se,
    ci_hi = mean + qt(0.975, df = n - 1) * se,
    .groups = "drop"
  ) %>%
  mutate(
    variable = factor(
      variable,
      levels = c("Sex", "Age group", "Education", "Statin use")
    )
  ) %>%
  arrange(variable, group)

group_stats
```

# Step 3: The tests

Use a two-sample t-test when the grouping variable has two groups.

```{r step-3-t-tests}
sex_t <- t.test(systolic_bp ~ sex, data = prevend, var.equal = TRUE)
sex_t
```

Use one-way ANOVA when the grouping variable has three or more groups.

```{r step-3-anova}
age_aov <- aov(systolic_bp ~ age_group, data = prevend)
summary(age_aov)

education_aov <- aov(systolic_bp ~ education, data = prevend)
summary(education_aov)
```

Use the same two-sample t-test for statin use, which has two groups.

```{r step-3-statin}
table(prevend$statin_use)

statin_t <- t.test(
  systolic_bp ~ statin_use,
  data = prevend,
  var.equal = TRUE
)

statin_t
```

# Step 4: Assemble the table

The formatting code is longer than the statistics code. That is normal. Your job is to make sure the formatted table faithfully reports the tests above.

```{r step-4-extract}
age_f <- summary(age_aov)[[1]]$`F value`[1]
age_p <- summary(age_aov)[[1]]$`Pr(>F)`[1]

education_f <- summary(education_aov)[[1]]$`F value`[1]
education_p <- summary(education_aov)[[1]]$`Pr(>F)`[1]
```

```{r step-4-build}
results_table <- group_stats %>%
  mutate(
    `Mean (SD)` = paste0(
      format(round(mean, 1), nsmall = 1),
      " (",
      format(round(sd, 1), nsmall = 1),
      ")"
    ),
    `95% CI` = paste0(
      format(round(ci_lo, 1), nsmall = 1),
      ", ",
      format(round(ci_hi, 1), nsmall = 1)
    ),
    statistic = case_when(
      variable == "Sex" ~ as.numeric(sex_t$statistic),
      variable == "Age group" ~ age_f,
      variable == "Education" ~ education_f,
      variable == "Statin use" ~ as.numeric(statin_t$statistic)
    ),
    p_value = case_when(
      variable == "Sex" ~ sex_t$p.value,
      variable == "Age group" ~ age_p,
      variable == "Education" ~ education_p,
      variable == "Statin use" ~ statin_t$p.value
    ),
    stat_text = as.character(round(statistic, 2)),
    p_text = if_else(p_value < 0.001, "<0.001", as.character(round(p_value, 3)))
  ) %>%
  group_by(variable) %>%
  mutate(row_in_block = row_number()) %>%
  ungroup() %>%
  mutate(
    `t/F` = if_else(row_in_block == 1, stat_text, ""),
    p = if_else(row_in_block == 1, p_text, "")
  ) %>%
  select(variable, group, `Mean (SD)`, `95% CI`, `t/F`, p)
```

```{r step-4-render}
results_table %>%
  gt(groupname_col = "variable", rowname_col = "group") %>%
  tab_header(title = "Systolic blood pressure by participant characteristic") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  tab_options(table.width = pct(100))
```

# Reading the results

Sex has two groups, so the table reports a t statistic. Age group and education have more than two groups, so the table reports F statistics from ANOVA.

A significant ANOVA tells us that at least one group mean differs. It does not tell us which pair of groups differs. For that, we would need a follow-up test such as Tukey's HSD.
