---
title: "SOLUTIONS: Replicating a publication table of group differences"
subtitle: "COVID-19 psychological distress (K6) by characteristic"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

These are the completed solutions for the week 5 exercise.

# PART 1: Replicate the table

## Load packages

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(readxl, tidyverse, gt, gtsummary, here, janitor)

pacman::p_load_current_gh("the-graph-courses/inspectdf")
```

## Step 1: Import and label the data

```{r data-preview}
survey <- read_excel(here("rsr_week_05_workshop/data/distress_data.xlsx"))
```

```{r glimpse-data}
glimpse(survey)
```

**Question**: How many rows and columns does the raw imported dataset have?

Answer: The raw dataset has 1599 rows and 38 columns.

**Question**: What are the variable types? Will we need to convert any variable types?

Answer: The variables are all imported as numeric. We need to convert the categorical variables (gender, age category, marriage, history of visiting Wuhan) to factors before making the table.

```{r missing-check}
summary(survey)
```

**Question**: Do any of the variables in the dataset have missing values?

Answer: No. None of the variables have missing values, so no `NA's` line appears in the summary output.

```{r import}
survey <- survey %>%
  mutate(
    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")),
    age = cut(
      age,
      breaks = c(-Inf, 30, 40, 50, Inf),
      labels = c("18-30", "31-40", "41-50", ">50")
    ),
    marriage = factor(
      if_else(marriage == 1, "married", "unmarried"),
      levels = c("unmarried", "married")
    ),
    history_of_visiting_SEA = factor(
      history_of_visiting_SEA,
      levels = c(0, 1),
      labels = c("No", "Yes")
    )
  ) %>%
  select(total_scores_of_K6, gender, age, marriage, history_of_visiting_SEA)

```

Checkpoint: After selecting variables, the dataset should have 1599 rows and 5 columns.

```{r group-checks}
tabyl(survey, gender)
tabyl(survey, age)
tabyl(survey, marriage)
tabyl(survey, history_of_visiting_SEA)
```

```{r inspect-cat}
inspect_cat(survey) %>% show_plot()
```

**Question**: How many men are in the dataset? How many people had visited Wuhan in the past month?

Answer: There are 531 men. There are 326 people who had visited Wuhan in the past month.

## Step 2: Descriptive stats for every group

```{r reshape}
survey_long <- survey %>%
  pivot_longer(
    cols = -total_scores_of_K6,
    names_to = "variable",
    values_to = "group"
  )

survey_long
```

```{r count-example}
survey_long %>%
  group_by(variable, group) %>%
  summarise(
    n = n()
  )
```

```{r descriptives}
group_stats <- survey_long %>%
  group_by(variable, group) %>%
  summarise(
    n = n(),
    mean = mean(total_scores_of_K6),
    sd = sd(total_scores_of_K6),
    se = sd / sqrt(n),
    ci_95_low_end = mean - qt(0.975, df = n - 1) * se,
    ci_95_high_end = mean + qt(0.975, df = n - 1) * se,
    .groups = "drop"
  ) %>%
  mutate(
    variable = factor(
      variable,
      levels = c("gender", "age", "marriage", "history_of_visiting_SEA"),
      labels = c("Gender", "Age category", "Marriage", "History of visiting Wuhan")
    )
  ) %>%
  arrange(variable, group)

group_stats
```

**Question**: For history of visiting Wuhan, which group has the higher mean distress score?

Answer: The `Yes` group has the higher mean distress score. Its mean is about 18.6, compared with about 5.0 in the `No` group.

## Step 3: The tests

First, the gender t-test.

```{r test-gender}
gender_t <- t.test(total_scores_of_K6 ~ gender, data = survey, var.equal = TRUE)
gender_t
```

**Question**: At alpha = 0.05, is the gender difference statistically significant? Do we reject or fail to reject the null hypothesis?

Answer: No. The p-value is about 0.316, which is greater than 0.05. We fail to reject the null hypothesis.

**Question**: Given this decision, which kind of decision error would we worry about for the gender t-test, Type I or Type II? What would that error mean here?

Answer: Because we fail to reject the null, a Type II error is the concern. It would mean that there is a real gender difference in mean distress in the population, but this study failed to detect it.

Next, the age ANOVA.

```{r test-age}
age_aov <- aov(total_scores_of_K6 ~ age, data = survey)
summary(age_aov)
```

**Question**: For the age ANOVA, is the age difference statistically significant? What does this tell us about the groups?

Answer: Yes. The p-value is less than 0.001. This tells us that at least one age-category mean differs from the others.

**Question**: Given this decision, which kind of decision error would we worry about for the ANOVA, Type I or Type II? What would that error mean here?

Answer: Because we reject the null, a Type I error is the concern. It would mean that we concluded age category is associated with mean distress when, in the population, the age-category means do not really differ.

Finally, the tests for marriage and history of visiting Wuhan.

```{r test-rest}
marriage_t <- t.test(total_scores_of_K6 ~ marriage, data = survey, var.equal = TRUE)

wuhan_t <- t.test(
  total_scores_of_K6 ~ history_of_visiting_SEA,
  data = survey,
  var.equal = TRUE
)

marriage_t
wuhan_t
```

## Step 4: Assemble the table with `gt`

```{r anova-extract}
age_f <- summary(age_aov)[[1]]$`F value`[1]
age_p <- summary(age_aov)[[1]]$`Pr(>F)`[1]

age_f
age_p
```

```{r build-table}
results_table <- group_stats %>%
  mutate(
    `Mean (SD)` = paste0(
      format(round(mean, 1), nsmall = 1),
      " (",
      format(round(sd, 1), nsmall = 1),
      ")"
    ),
    `95% CI` = paste0(
      format(round(ci_95_low_end, 2), nsmall = 2),
      ", ",
      format(round(ci_95_high_end, 2), nsmall = 2)
    ),
    statistic = case_when(
      variable == "Gender" ~ as.numeric(gender_t$statistic),
      variable == "Age category" ~ age_f,
      variable == "Marriage" ~ as.numeric(marriage_t$statistic),
      variable == "History of visiting Wuhan" ~ as.numeric(wuhan_t$statistic)
    ),
    p_value = case_when(
      variable == "Gender" ~ gender_t$p.value,
      variable == "Age category" ~ age_p,
      variable == "Marriage" ~ marriage_t$p.value,
      variable == "History of visiting Wuhan" ~ wuhan_t$p.value
    ),
    stat_text = as.character(round(statistic, 2)),
    p_text = if_else(p_value < 0.001, "<0.001", as.character(round(p_value, 3)))
  ) %>%
  group_by(variable) %>%
  mutate(row_in_block = row_number()) %>%
  ungroup() %>%
  mutate(
    `t/F` = if_else(row_in_block == 1, stat_text, ""),
    p = if_else(row_in_block == 1, p_text, "")
  )
```

```{r render-table}
results_table %>%
  select(variable, group, `Mean (SD)`, `95% CI`, `t/F`, p) %>%
  gt(groupname_col = "variable", rowname_col = "group") %>%
  tab_header(title = "Psychological distress (K6) by characteristic") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  tab_options(table.width = pct(100))
```

**Question**: Because you did not write the gt code yourself (we wrote it for you, and in the future, LLMs might write such formatting code for you), it is a good idea to cross check all values in the table against the raw test output from your `t.test()` or `aov()` functions to make sure it is correct. For now, check the Marriage rows test. Compare the table values to the output from `t.test()`. What did you find? Are there any differences worth highlighting? (e.g. is the p-value reported exactly the same way?)

Answer: The Marriage row in the table shows t = 15.28 and p < 0.001. These match the raw `t.test()` output after rounding. The raw output gives the p-value in scientific notation, about 2.71e-49, while the table reports it as `<0.001`.

**Question**: Which characteristics are statistically significant at alpha = 0.05?

Answer: Age category, Marriage, and History of visiting Wuhan are statistically significant. Gender is not statistically significant.

**Question**: Choose one significant result and report it in plain language as you would in a results section of a paper.

Answer: People who had visited Wuhan in the past month had significantly higher mean K6 distress scores than those who had not (18.6 vs 5.0), t(1597) = -40.9, p < 0.001.

## LLM formatting practice

```{r glimpse-results}
glimpse(results_table)
```

In your prompt, also refer to the current `gt` code from the `render-table` chunk above.

One possible extension:

```{r}
library(gt)
library(gtExtras)

results_table %>%
  # 1. Include the numeric 'mean' column in the selection
  select(variable, group, mean, `Mean (SD)`, `95% CI`, `t/F`, p) %>%
  gt(groupname_col = "variable", rowname_col = "group") %>%
  
  # 2. Add the inline bar chart (scaled = FALSE normalizes to the max value)
  gt_plt_bar_pct(column = mean, fill = "steelblue", scaled = FALSE) %>%
  
  # 3. Label the new bar chart column
  cols_label(mean = "Mean Trend") %>%
  
  tab_header(title = "Psychological distress (K6) by characteristic") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  tab_options(table.width = pct(100))
```

```{r llm-bars, eval=FALSE}
pacman::p_load(gtExtras)

results_table %>%
  mutate(mean_bar = mean) %>%
  select(variable, group, `Mean (SD)`, mean_bar, `95% CI`, `t/F`, p) %>%
  gt(groupname_col = "variable", rowname_col = "group") %>%
  gtExtras::gt_plt_bar(
    column = mean_bar,
    keep_column = TRUE,
    color = "#2F6B9A"
  ) %>%
  cols_label(mean_bar = "Mean") %>%
  tab_header(title = "Psychological distress (K6) by characteristic")
```

## Side note: gtsummary

It is worth knowing that the `gtsummary` package can get us some of the way there quickly. The `tbl_continuous()` function is useful when we have one continuous outcome and can give us the group means and standard deviations, plus p-values. However it is tricky to get the 95% confidence intervals or the t/F statistic column in the same format as the paper. Run the code below to see an example `tbl_continuous()` output.

```{r gtsummary-table}
survey %>%
  tbl_continuous(
    variable = total_scores_of_K6,
    include = c(gender, age, marriage, history_of_visiting_SEA),
    statistic = ~ "{mean} ({sd})",
    label = list(
      gender ~ "Gender",
      age ~ "Age category",
      marriage ~ "Marriage",
      history_of_visiting_SEA ~ "History of visiting Wuhan"
    )
  ) %>%
  add_p()
```

# PART 2: Power and sample size for a follow-up study

Suppose this survey is pilot data for a follow-up randomized study. You want to test whether a stress-management program reduces K6 distress compared with a control group.

```{r pilot-sd}
pilot_sd <- sd(survey$total_scores_of_K6)
round(pilot_sd, 1)
```

## Sample size for 80% power

Suppose a 2-point lower mean K6 score in the intervention group compared with the control group is the smallest difference worth detecting.

```{r power-80}
power_80 <- power.t.test(
  delta = 2,
  sd = pilot_sd,
  sig.level = 0.05,
  power = 0.80,
  type = "two.sample"
)

power_80
```

**Question**: How many participants are needed per group? Round up.

Answer: R returns about 231.8, so we round up to 232 participants per group.

**Question**: If the chance of detecting a 2-point difference is 80% with our sample size, what happens if the actual difference is 4 points? Do we have a higher or lower chance of detecting it?

Answer: We have a higher chance of detecting it. Larger true differences are easier to detect.

## Smaller difference

Now you decide that even a 1-point difference is worth detecting.

```{r power-delta-1}
power_delta_1 <- power.t.test(
  delta = 1,
  sd = pilot_sd,
  sig.level = 0.05,
  power = 0.80,
  type = "two.sample"
)

power_delta_1
```

**Question**: Does the required sample size go up or down compared with the 2-point version? Why?

Answer: It goes up. R returns about 924.4, so we round up to 925 participants per group. A smaller difference is harder to detect, so we need more participants.

# Summary

The table shows no statistically significant gender difference, but age category, marriage, and history of visiting Wuhan are statistically significant. The power section shows that sample size increases when the minimum difference worth detecting gets smaller.
