---
title: "SOLUTIONS: Simple linear regression"
subtitle: "FEV lung function: age then height"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

These are instructor solutions for `fev/rsr_week_06_exercise.Rmd`. Each section shows the **question** (where applicable), **blank code** from the exercise, **completed code**, and **answer**. A consolidated **answer key** is at the end.

In this exercise you will practice **simple linear regression** on lung-function data from youths ([Kahn/Tager, JSE 2005](https://doi.org/10.1080/10691898.2005.11910559)).

- **Outcome (both models):** `fev_l`, forced expiratory volume in one second, in **litres**
- **Dataset:** `data/fev_kahn.csv`, 654 youths
- **Predictors:** `age_yrs` (years) and `height_in` (inches)

There is **no separate demo file**. **Part 1** walks through a complete example (`fev_l ~ age_yrs`) with all code filled in. **Part 2** repeats the same steps for **height**. **Part 3** compares the two models.

# PART 1: Walkthrough: FEV and age (filled in)

Read this section and **run every chunk**. The code is complete; treat this as your syntax reference for Part 2.

## Load packages and data

```{r demo-setup, echo=TRUE, message=FALSE, warning=FALSE}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, janitor)

fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint:** 654 rows, 5 columns. No missing values in `fev_l` or `age_yrs`.

## Plot the relationship

```{r demo-scatter, echo=TRUE}
ggplot(fev, aes(x = age_yrs, y = fev_l)) +
  geom_point(alpha = 0.35, color = "steelblue") +
  labs(
    title = "FEV vs age",
    x = "Age (years)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

Older youths tend to have higher FEV. Lungs grow as children age.

## Correlation

```{r demo-cor, echo=TRUE}
r_age <- cor(fev$fev_l, fev$age_yrs)
round(r_age, 3)
```

**Answer:** $r \approx 0.756$, a strong **positive** linear association.

## Fit the model and read the output

```{r demo-fit, echo=TRUE}
fev_age_model <- lm(fev_l ~ age_yrs, data = fev)
summary(fev_age_model)
```

From the **`age_yrs` row** of the coefficients table:

| Quantity | Approximate value | Meaning |
|----------|-------------------|---------|
| Estimate (slope $b_1$) | 0.222 | Litres of FEV gained per extra year of age |
| `Pr(>\|t\|)` | $< 0.001$ | Test of $H_0: \beta_1 = 0$ |

**Answer:** Each additional year of age is associated with about **0.22 litres** higher FEV, on average.

## $R^2$

```{r demo-r2, echo=TRUE}
summary(fev_age_model)$r.squared
r_age^2
```

**Answer:** $R^2 \approx 0.572$. Age alone explains about **57%** of person-to-person variation in FEV. $0.756^2 \approx 0.572$.

## Fitted line and a prediction

```{r demo-line, echo=TRUE}
ggplot(fev, aes(x = age_yrs, y = fev_l)) +
  geom_point(alpha = 0.35, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
  labs(
    title = "FEV vs age with fitted line",
    x = "Age (years)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

```{r demo-predict, echo=TRUE}
predict(fev_age_model, newdata = data.frame(age_yrs = 10))
```

**Answer:** Predicted FEV at age 10 is about **2.65 litres**.

---

**Before Part 2:** you should know how to go from scatter plot to `cor()`, then `lm()`, then `summary()`, then $R^2$, then `predict()`. Part 2 repeats that recipe with **height** as the predictor.

# PART 2: Your turn: FEV and height

Fill in the blanks (replace `_______`). Use Part 1 as your syntax guide.

## Scatter plot

**Question:** Does FEV tend to increase or decrease with height?

**Blank code:**

```{r scatter-height-blank, eval=FALSE}
ggplot(fev, aes(x = _______, y = _______)) +
  geom_point(alpha = 0.35, color = "steelblue") +
  labs(
    title = "FEV vs height",
    x = "Height (inches)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

**Solution:**

```{r scatter-height}
ggplot(fev, aes(x = height_in, y = fev_l)) +
  geom_point(alpha = 0.35, color = "steelblue") +
  labs(
    title = "FEV vs height",
    x = "Height (inches)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

**Answer:** FEV **increases** with height.

## Correlation

**Question:** What is $r$ for height? Is the association positive or negative?

**Blank code:**

```{r cor-height-blank, eval=FALSE}
r_height <- cor(_______, _______)
round(r_height, 3)
```

**Solution:**

```{r cor-height}
r_height <- cor(fev$fev_l, fev$height_in)
round(r_height, 3)
```

**Answer:** $r \approx 0.868$. The association is **positive**.

## Fit the model

**Question:** What is the slope for `height_in` (Estimate, rounded to 3 decimals)?

**Question:** Interpret the slope in plain language (include units for FEV and height).

**Question:** Is the slope statistically significant at $\alpha = 0.05$?

**Blank code:**

```{r fit-height-blank, eval=FALSE}
fev_height_model <- _______(_______ ~ _______, data = fev)
summary(fev_height_model)
```

**Solution:**

```{r fit-height}
fev_height_model <- lm(fev_l ~ height_in, data = fev)
summary(fev_height_model)
```

**Answer:** Slope $\approx 0.132$. Each additional inch of height is associated with about **0.13 litres** higher FEV, on average. **Yes**, reject $H_0$ at $\alpha = 0.05$ ($p < 0.001$).

## $R^2$

**Question:** What is $R^2$ for the height model?

**Question:** Confirm that $r^2$ from Part 2 matches $R^2$ (allowing for rounding).

**Blank code:**

```{r r2-height-blank, eval=FALSE}
r_squared_height <- summary(_______)$r.squared
round(r_squared_height, 3)
```

**Solution:**

```{r r2-height}
r_squared_height <- summary(fev_height_model)$r.squared
round(r_squared_height, 3)
```

**Answer:** $R^2 \approx 0.754$. Check: $0.868^2 \approx 0.754$.

## Prediction

**Question:** Predicted FEV at 60 inches (litres)?

**Blank code:**

```{r predict-height-blank, eval=FALSE}
predict(
  fev_height_model,
  newdata = data.frame(_______ = 60)
)
```

**Solution:**

```{r predict-height}
predict(
  fev_height_model,
  newdata = data.frame(height_in = 60)
)
```

**Answer:** About **2.49 litres**.

## Fitted line

**Blank code:**

```{r line-height-blank, eval=FALSE}
ggplot(fev, aes(x = height_in, y = fev_l)) +
  geom_point(alpha = 0.35, color = "steelblue") +
  geom_______(method = "lm", se = TRUE, color = "firebrick") +
  labs(
    title = "FEV vs height with fitted line",
    x = "Height (inches)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

**Solution:**

```{r line-height}
ggplot(fev, aes(x = height_in, y = fev_l)) +
  geom_point(alpha = 0.35, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
  labs(
    title = "FEV vs height with fitted line",
    x = "Height (inches)",
    y = "FEV (litres)"
  ) +
  theme_minimal()
```

# PART 3: Compare age and height models

Use your Part 1 walkthrough and your Part 2 answers.

Fill in the code to put both $R^2$ values side by side:

**Blank code:**

```{r compare-r2-blank, eval=FALSE}
tibble(
  predictor = c("age_yrs", "height_in"),
  r_squared = c(
    summary(_______)$r.squared,
    summary(_______)$r.squared
  )
) %>%
  mutate(r_squared = round(r_squared, 3))
```

**Solution:**

```{r compare-r2}
tibble(
  predictor = c("age_yrs", "height_in"),
  r_squared = c(
    summary(fev_age_model)$r.squared,
    summary(fev_height_model)$r.squared
  )
) %>%
  mutate(r_squared = round(r_squared, 3))
```

**Question:** Which predictor has the larger $|r|$, age or height?

**Answer:** **Height** ($|r| \approx 0.87$ vs $|r| \approx 0.76$ for age).

**Question:** Which model has the higher $R^2$?

**Answer:** **Height** ($R^2 \approx 0.75$ vs $R^2 \approx 0.57$ for age).

**Question:** In one or two sentences, which single predictor would you prefer for explaining FEV in this sample, and why?

**Answer:** Prefer **height**. It explains about 75% of FEV variation compared with about 57% for age alone, so predictions from the height model would be tighter.

**Question:** Age and height both increase as children grow. Why might height be a stronger predictor of FEV than age alone?

**Answer:** FEV depends on **lung size**, which tracks body size (height) more directly than age. Two youths the same age can differ in height, and the taller one typically has larger lungs and higher FEV.

# PART 4 OPTIONAL UNGRADED CHALLENGE: Smoking

The dataset includes `smoke` (0 = non-smoker, 1 = smoker). In this sample, smokers tend to be **older**, which can confuse a simple age-only story.

```{r smoke-plot}
ggplot(fev, aes(x = age_yrs, y = fev_l, color = factor(smoke))) +
  geom_point(alpha = 0.4) +
  labs(
    title = "FEV vs age by smoking status",
    x = "Age (years)",
    y = "FEV (litres)",
    color = "Smoker"
  ) +
  theme_minimal()
```

```{r residuals}
par(mfrow = c(1, 2))
plot(fev_age_model, which = 1:2)
par(mfrow = c(1, 1))
```

**Question:** After seeing the smoking plot, why might neither an age-only nor a height-only model tell the whole story about FEV?

**Answer:** **Smoking** may directly affect lung function. Smokers in this sample tend to be older, so part of the age association could be confounded. Neither single-predictor model accounts for smoking (or sex, which is also in the dataset).

---

# Answer key (quick reference)

| Part | Question | Answer |
|------|----------|--------|
| 1 | $r$ (age) | ≈ 0.756 (positive) |
| 1 | Slope (age) | ≈ 0.222 L per year |
| 1 | $R^2$ (age) | ≈ 0.572 |
| 1 | Predict at age 10 | ≈ 2.65 L |
| 2 | Scatter pattern | FEV increases with height |
| 2 | $r$ (height) | ≈ 0.868 (positive) |
| 2 | Slope (height) | ≈ 0.132 L per inch; significant |
| 2 | $R^2$ (height) | ≈ 0.754 |
| 2 | Predict at 60 in | ≈ 2.49 L |
| 3 | Larger $|r|$ | Height |
| 3 | Higher $R^2$ | Height |
| 3 | Preferred predictor | Height (explains more variation) |
| 3 | Why height beats age | Lung size tracks height |
| 4 | Smoking | Confounding; omitted variable |

**Code blanks summary**

| Chunk | Blanks filled with |
|-------|-------------------|
| `scatter-height` | `height_in`, `fev_l` |
| `cor-height` | `fev$fev_l`, `fev$height_in` |
| `fit-height` | `lm`, `fev_l`, `height_in` |
| `r2-height` | `fev_height_model` |
| `predict-height` | `height_in` |
| `line-height` | `smooth` |
| `compare-r2` | `fev_age_model`, `fev_height_model` |
