---
title: "EXERCISE: Multiple linear regression and how to read it"
subtitle: "FEV lung function: age, height, sex, and smoking together"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Two workshops ago, we used the FEV dataset to see the three uses of regression: describe, predict, and control. At the end, we previewed a multiple regression that controlled for age and height. This week, multiple linear regression is the main event. You will build a model up one predictor at a time, and learn to read every row of the output.

The dataset contains 654 youths aged 3 to 19 from East Boston in the 1970s. The outcome is **forced expiratory volume** (`fev_l`), in litres: how much air a person can forcefully breathe out in one second. Higher values mean better lung function.

The variables are:

- `fev_l`: forced expiratory volume in litres (the outcome)
- `age_yrs`: age in years
- `height_in`: height in inches
- `sex`: `female` or `male`
- `smoke`: current smoking status, `non_smoker` or `smoker`

# Load packages and data

Use `p_load()` to load: tidyverse, here, broom.

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

```{r load-data, eval=FALSE}
fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint:** `fev` should have **654 rows and 5 columns**.

# Part 1: One predictor, a recap

In week 6 we found that height was the best single predictor of FEV. Refit that model to remind yourself what its output looks like.

```{r height-model, eval=FALSE}
height_model <- lm(______ ~ ______, data = fev)
summary(height_model)
```

**Question:** Fill in the blanks.

The fitted equation is: predicted FEV = ______ + ______ * height.
The model's R-squared is about ______.

Ans:

# Part 2: Two predictors

## Part 2a: Fitting the model

To fit a multiple linear regression, we simply add predictors on the right of the `~` with `+`. Fit a model with **both height and age** as predictors.

```{r height-age-model, eval=FALSE}
height_age_model <- lm(______ ~ ______ + ______, data = fev)
summary(height_age_model)
```

The fitted equation now has two slopes:

predicted FEV = -4.61 + 0.110 * height + 0.054 * age

## Part 2b: The key change in interpretation

In multiple regression, each coefficient is read **holding the other predictors constant** (also phrased as "adjusting for" or "controlling for" the other predictors).

So the height slope now answers: *comparing two youths of the same age, how much higher is predicted FEV for each extra inch of height?*

**Question:** Interpret the `age_yrs` coefficient (about 0.054) in one sentence, using the phrase "holding height constant".

Ans:

**Question:** In the simple model, the height slope was about 0.132. In this model it dropped to about 0.110. Why did it change? (Hint: compute `cor(fev$height_in, fev$age_yrs)`.)

```{r height-age-cor, eval=FALSE}
cor(fev$height_in, fev$age_yrs)
```

Ans:

**Question:** The R-squared went from 0.754 (height alone) to about 0.766 (height + age). Age was a strong predictor on its own (R-squared 0.572 in week 6). Why did adding it improve the model so little?

Ans:

# Part 3: The full model

## Part 3a: Fit it

Now fit the full model with all four predictors: height, age, sex, and smoking status.

```{r full-model, eval=FALSE}
full_model <- lm(fev_l ~ ______ + ______ + ______ + ______, data = fev)
summary(full_model)
```

## Part 3b: Reading every row

Recall from previous weeks: for a categorical predictor, R picks the first factor level (alphabetically, for character columns) as the **reference group**, and the coefficient row compares the other level with the reference.

**Question:** For the `sexmale` row (coefficient about 0.157): which sex is the reference group, and what does the coefficient mean? Use "holding ... constant" in your answer.

Ans:

**Question:** Fill in the blanks for the `smokesmoker` row.

Holding height, age, and sex constant, smokers have predicted FEV about ______ litres ______ (higher / lower) than non-smokers. The p-value for this coefficient is about ______, so at alpha = 0.05 this difference ______ (is / is not) statistically significant.

Ans:

## Part 3c: Confidence intervals with broom

Use `tidy()` from the broom package to get each coefficient with its 95% confidence interval.

```{r tidy-full, eval=FALSE}
tidy(______, conf.int = ______)
```

**Question:** What is the 95% CI for the `smokesmoker` coefficient? Does it include 0, and how does that relate to the p-value you saw above?

Ans:

## Part 3d: The big picture on smoking

Let's put the smoking story together across models. First, the raw (unadjusted) smoking model:

```{r smoke-raw, eval=FALSE}
raw_smoke_model <- lm(fev_l ~ smoke, data = fev)
tidy(raw_smoke_model, conf.int = TRUE)
```

**Question:** Complete the summary table below in words: what is the smoking coefficient in the raw model, and in the full adjusted model? Explain in two or three sentences what happened and why. (Think back to week 6: who were the smokers in this dataset?)

Ans:

**Question:** A classmate concludes: "the full model proves smoking has no effect on lung function in youths." Give one reason to push back on that conclusion.

Ans:

# Part 4: Prediction with multiple predictors

Prediction works exactly as before, except the new data must supply a value for **every** predictor in the model.

Use the full model to predict FEV for a **12-year-old, 60-inch-tall, female non-smoker**.

```{r predict, eval=FALSE}
new_person <- tibble(
  height_in = ______,
  age_yrs = ______,
  sex = "______",
  smoke = "______"
)
predict(______, newdata = ______)
```

**Question:** What is the predicted FEV? Verify it by hand using the fitted equation (intercept -4.457, height 0.104, age 0.066, sexmale 0.157, smokesmoker -0.087). Remember: for a female non-smoker, the `sexmale` and `smokesmoker` terms contribute 0.

Ans:

**Question:** Without rerunning `predict()`, what would the prediction be for the same person if she were a smoker?

Ans:

# Part 5: Optional challenge, an interaction

In the lessons this week you saw interaction terms, which let the slope of one predictor differ between groups. Fit a model where the **height slope is allowed to differ between smokers and non-smokers**, using the `:` syntax.

```{r interaction, eval=FALSE}
int_model <- lm(fev_l ~ height_in + smoke + ______, data = fev)
summary(int_model)
```

**Question:** Using the coefficient table, what is the height slope for non-smokers, and what is it for smokers? Is there strong evidence that the slopes differ?

Ans:

# Wrap-up

The mechanics of multiple regression are just `+` in the formula. The skill is in the reading:

1. Every coefficient is interpreted **holding the other predictors constant**.
2. When predictors are correlated (like height and age), coefficients shift as you add or remove predictors, and R-squared contributions do not add up.
3. Categorical predictors are read against their reference group, exactly as in simple regression.
4. Adjusted coefficients can tell a completely different story from raw ones.
5. Prediction requires a value for every predictor; changing one predictor by one unit changes the prediction by that coefficient.
