---
title: "EXERCISE: The three purposes of simple linear regression"
subtitle: "Children's lung function (FEV)"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Regression is one of the most useful tools in statistics because it serves **three different purposes**:

1. **Describe** a relationship between two variables (how do they move together?).
2. **Predict** the outcome for a new individual.
3. **Control** for other variables, so we can ask sharper, more causal-style questions.

In this workshop we explore all three using a classic dataset of **654 children and teenagers** (ages 3 to 19). For each child we have their **forced expiratory volume (FEV)**, the air, in litres, they can forcibly blow out in one second. FEV is a standard, concrete measure of lung function. The outcome is `fev_l` throughout.

A note on how to work through this file: the small amount of *new code* is shown worked out for you (computing a correlation, fitting a line, predicting). Your job is mostly to **fill in a few blanks and interpret the results**. Interpretation is where the real skill is.

## Load packages and import the data

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, janitor, broom)
```

```{r import}
fev <- read_csv(here("data/fev_lung_function.csv"))
glimpse(fev)
```

# PURPOSE 1: DESCRIBE the relationship

## 1a. Correlation: one number for the relationship (worked example)

The first thing to do when describing a relationship is to **plot it**. Here is FEV against age:

```{r demo-scatter}
ggplot(fev, aes(x = age_yrs, y = fev_l)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(x = "Age (years)", y = "Lung function, FEV (litres)",
       title = "FEV vs age") +
  theme_minimal()
```

The cloud climbs upward. To put a single number on that, we use the **correlation coefficient `r`**, a number between -1 and +1.

- Near +1: strong positive (both rise together).
- Near -1: strong negative (one rises as the other falls).
- Near 0: no linear relationship.

Here is the correlation between **age** and FEV, worked for you:

```{r demo-cor}
cor(fev$age_yrs, fev$fev_l)
```

An r of 0.76: older children tend to have larger FEV. Useful, but it is just one number. It does not give us the *equation* of the relationship, or let us predict, or tell us the units.

## 1b. Fit a line: the equation of the relationship (worked example)

A linear regression goes further: it gives the line that best describes the relationship. We fit it with `lm()`, always written `outcome ~ predictor`.

```{r demo-fit}
model_age <- lm(fev_l ~ age_yrs, data = fev)
summary(model_age)
```

The fitted line is `predicted FEV = 0.43 + 0.222 * age`. The **slope of 0.222** tells us, in real units, that each additional year of age is associated with about **0.22 litres** more lung capacity. Correlation could not tell us that.

## 1c. R-squared: a better summary of how well the line describes the data

The correlation and the slope each give one piece. **R-squared** ties it together: it is the proportion of the variation in the outcome that the line explains, on a clean 0-to-1 scale.

```{r demo-r2}
glance(model_age)$r.squared
```

An R-squared of 0.57: age explains about **57%** of the child-to-child variation in FEV.

Why is R-squared often more useful than the raw correlation?

- It is on an interpretable **0-1 "proportion explained"** scale, rather than the harder-to-read -1 to +1.
- It is always **positive**, so it summarizes fit without you having to track the sign separately.
- It lets you **compare models** on the same footing, including models with more than one predictor later, where a single correlation no longer applies.

(In simple regression with one predictor, R-squared is exactly the correlation squared: 0.76 squared is 0.57. The two are linked here, but R-squared keeps working when correlation runs out.)

## YOUR TURN: describe FEV using HEIGHT

Now describe the relationship between **height** (`height_in`) and FEV, and see whether height describes lung function better or worse than age does.

Fill in the blanks.

```{r ex-cor}
# correlation between height and FEV
cor(__________, __________)
```

**Question:** What is the correlation between height and FEV? Is it stronger or weaker than the age correlation (0.76)?

Your answer:

```{r ex-fit}
# fit the line: FEV explained by height
model_height <- lm(__________ ~ __________, data = fev)
summary(model_height)
```

**Question:** Write the fitted equation `predicted FEV = b0 + b1 * height_in`. Interpret the slope: how much more FEV per extra inch of height?

Your answer:

**Question:** The intercept is about -5.4 litres (a negative, impossible lung volume). Why is the intercept not a meaningful FEV value here? (Hint: how tall is a child of height 0, and is that near the data?)

Your answer:

```{r ex-r2}
# R-squared for the height model
________(model_height)$r.squared
```

**Question:** What proportion of the variation in FEV does height explain?

Your answer:

**Question (the comparison):** Put the two R-squared values together: age (0.57) versus height. Which variable **describes** FEV better, and by how many percentage points? Which had the correlation closer to 1, and does that match?

Your answer:

# PURPOSE 2: PREDICT a new child's lung function

Once we have a fitted line, we can plug in a predictor value and read off a predicted outcome with `predict()`. Here is a prediction for a 10-year-old, using the age model (worked for you):

```{r demo-predict}
predict(model_age, newdata = data.frame(age_yrs = 10))
```

The model predicts an FEV of about 2.65 litres for a typical 10-year-old.

## YOUR TURN: predict from height

Use your `model_height` to predict the FEV of a child who is **60 inches** tall. Fill in the blank.

```{r ex-predict}
predict(model_height, newdata = data.frame(height_in = ______))
```

**Question:** What FEV does the model predict for a 60-inch child?

Your answer:

**Question:** The children in this study are between 46 and 74 inches tall. Would you trust this model to predict the FEV of a 30-inch toddler? Why or why not? (This is the danger of predicting outside the range of the data, called **extrapolation**.)

Your answer:

# PURPOSE 3: CONTROL for other variables (a peek ahead)

The third purpose is the most powerful, and it is where regression becomes a tool for careful, almost-causal reasoning. We will only preview it here. The full method (multiple linear regression) comes in a later session.

The dataset records each child's **smoking status** (`smoke`): whether the child or teenager is themselves a smoker. Let's ask the obvious public-health question: is smoking associated with lung function?

First, compare average FEV by smoking status (worked for you):

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    n = n(),
    mean_fev = mean(fev_l)
  )
```

**Question:** According to this summary, do smokers have higher or lower average lung function than non-smokers? Does that match what you would expect smoking to do to the lungs?

Your answer:

That result, smokers having **higher** FEV, looks absurd. The catch is that smokers and non-smokers differ in another big way. Compare their average ages:

```{r smoke-age}
fev %>%
  group_by(smoke) %>%
  summarise(
    mean_age = mean(age_yrs)
  )
```

**Question:** Are the smokers older or younger, on average? Combine this with Purpose 1 (older children have bigger lungs). Explain why a simple comparison makes smoking look beneficial.

Your answer:

This is exactly the problem regression's third purpose solves. We can **control for age** by adding it to the model alongside smoking, fitting `fev_l ~ smoke + age_yrs`. This compares smokers and non-smokers *of the same age*. You have not formally learned multiple regression yet, so just run this and look at what happens to the smoking effect:

```{r control-age}
# A peek at MULTIPLE regression: two predictors at once
model_controlled <- lm(fev_l ~ smoke + age_yrs, data = fev)
tidy(model_controlled)
```

**Question:** Look at the `smokesmoker` row. Without controlling for age, smokers' mean FEV was about 0.71 litres *higher*. After we add `age_yrs` to the model, what is the sign of the smoking coefficient now, positive or negative? What does that tell us about smoking and lung function once we compare children of the same age?

Your answer:

**Question (wrap-up):** In your own words, summarize the three purposes of regression you have now seen (**describe**, **predict**, and **control**) in one short sentence each, using the FEV examples.

Your answer:
