---
title: "SOLUTIONS: Multiple linear regression and how to interpret it"
subtitle: "FEV lung function, age, height, sex, and smoking"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In Workshop 6 we used the FEV (lung function) dataset to introduce **simple** linear regression: one outcome, one predictor. At the very end of that workshop, we previewed what happens when we add more than one predictor at once, to control for confounding.

This workshop builds that preview into a full **multiple linear regression (MLR)** model, and focuses on the main new skill: **reading and interpreting a multiple regression coefficient table**.

Recall the dataset: 654 youths aged 3 to 19 from East Boston in the 1970s. The outcome is **forced expiratory volume** (`fev_l`), in litres. Higher values mean better lung function.

The variables we will use are:

- `fev_l`: forced expiratory volume in litres (outcome)
- `age_yrs`: age in years
- `height_in`: height in inches
- `sex`: `female` or `male`
- `smoke`: current smoking status, `non_smoker` or `smoker`

# Load packages and data

Use `p_load()` to load the following packages: `tidyverse` and `here`.

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here)
```

```{r load-data}
fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint:** `fev` should have **654 rows and 5 columns**.

**Checkpoint answer:** `fev` has **654 rows and 5 columns**.

# Part 1: From one predictor to many

## Recap: two one-predictor models

As a reminder, here are the simple regressions of FEV on age and on height separately.

```{r age-model}
age_model <- lm(fev_l ~ age_yrs, data = fev)
summary(age_model)
```

```{r height-model}
height_model <- lm(fev_l ~ height_in, data = fev)
summary(height_model)
```

**Question:** Age and height are both positively associated with FEV on their own. Age and height are also correlated with each other (taller children tend to be older). What do you think will happen to the age coefficient once we also control for height in the same model? Give your best guess before running anything.

Ans:

**Solution answer:** There is no single correct guess here, but a reasonable prediction is that the age coefficient will shrink once height is added, because much of what looks like an "age effect" on FEV is really an effect of body size (height) that happens to be correlated with age. This is exactly what happens below: the age slope drops from about 0.222 to about 0.060.

## The multiple regression formula

A multiple regression with two predictors extends the simple regression formula:

predicted y = intercept + slope_1 * x_1 + slope_2 * x_2

In R, we add predictors with `+`:

```r
lm(outcome ~ predictor_1 + predictor_2, data = my_data)
```

Each coefficient is now read while **holding the other predictors constant** (also called **adjusting for** or **controlling for** the other predictors).

## Fit the two-predictor model

Fill in the blanks to fit a model with `fev_l` as the outcome and both `age_yrs` and `height_in` as predictors.

```{r age-height-model-blank, eval=FALSE}
age_height_model <- lm(_______ ~ _______ + _______, data = fev)
summary(age_height_model)
```

**Solution:**

```{r age-height-model}
age_height_model <- lm(fev_l ~ age_yrs + height_in, data = fev)
summary(age_height_model)
```

**Question:** Write out the fitted equation, rounding each number to 3 decimals.

predicted FEV = ______ + ______ * age + ______ * height

Ans:

**Solution answer:**

predicted FEV = -4.087 + 0.065 * age + 0.104 * height

**Question:** Compare the age coefficient in this model (about 0.060) with the age coefficient in the one-predictor age model (about 0.222). What happened, and why do you think it happened?

Ans:

**Solution answer:** The age coefficient shrank a lot, from about 0.222 down to about 0.065. This happened because age and height are strongly correlated (r ≈ 0.79) in this dataset, and height itself is a strong predictor of FEV. In the one-predictor age model, age was "standing in" for some of the effect that is really due to height. Once height is included directly, age's own independent contribution is much smaller.

**Question:** Interpret the height coefficient (about 0.109) in one sentence, using the phrase "holding age constant."

Ans:

**Solution answer:** Holding age constant, each additional inch of height is associated with about 0.104 litres higher predicted FEV, on average.

**Question:** Both age and height are still statistically significant (p < 0.05) in the two-predictor model. What does that tell us, in plain language, about each predictor's relationship with FEV?

Ans:

**Solution answer:** It tells us that both age and height each have their own independent association with FEV, even after accounting for the other. Neither predictor's relationship with FEV is fully explained away by the other.

# Part 2: Adding a third predictor, your turn

Now bring back the `smoke` variable from Workshop 6. Recall that in the raw (one-predictor) model, smokers had *higher* mean FEV than non-smokers, which reversed after controlling for age and height, because smokers in this dataset are older and taller on average.

Fit a model with `age_yrs`, `height_in`, and `smoke` all together.

```{r age-height-smoke-model-blank, eval=FALSE}
full_model_1 <- lm(_______ ~ age_yrs + height_in + smoke, data = fev)
summary(full_model_1)
```

**Solution:**

```{r age-height-smoke-model}
full_model_1 <- lm(fev_l ~ age_yrs + height_in + smoke, data = fev)
summary(full_model_1)
```

**Question:** What is the coefficient for `smokesmoker`, and is it statistically significant at alpha = 0.05?

Ans:

**Solution answer:** The `smokesmoker` coefficient is about -0.110, with p ≈ 0.067. It is **not** statistically significant at alpha = 0.05, since 0.067 > 0.05.

**Question:** Interpret the `smokesmoker` coefficient in one sentence. Be careful to say what it means "holding age and height constant."

Ans:

**Solution answer:** Holding age and height constant, smokers have a predicted FEV about 0.110 litres lower than non-smokers, on average, although this difference is not statistically significant at the 0.05 level.

## Adding a fourth predictor: sex

The dataset also has a `sex` variable (`female` or `male`) that we have not used yet. Boys and girls differ, on average, in both height and lung development, so sex is a reasonable predictor to add.

Fit a model with all four predictors: `age_yrs`, `height_in`, `smoke`, and `sex`.

```{r full-model-2-blank, eval=FALSE}
full_model_2 <- lm(_______ ~ age_yrs + height_in + smoke + _______, data = fev)
summary(full_model_2)
```

**Solution:**

```{r full-model-2}
full_model_2 <- lm(fev_l ~ age_yrs + height_in + smoke + sex, data = fev)
summary(full_model_2)
```

**Question:** What is the coefficient for `sexmale`? Interpret it in one sentence, holding the other three predictors constant.

Ans:

**Solution answer:** The `sexmale` coefficient is about 0.157. Holding age, height, and smoking status constant, boys have predicted FEV about 0.157 litres higher than girls, on average. This coefficient is statistically significant (p < 0.001).

**Question:** After adding `sex` to the model, what happens to the `smokesmoker` coefficient and its p-value, compared with the three-predictor model? (It moves from about -0.110, p = 0.067, to a new value.) Is smoking status statistically significant in this four-predictor model?

Ans:

**Solution answer:** The `smokesmoker` coefficient moves to about -0.087, with p ≈ 0.141. It is still not statistically significant, and is now even further from the 0.05 threshold than before.

# Part 3: R-squared and adjusted R-squared

Each time we added a predictor, R-squared went up a little. This always happens: adding *any* predictor, even a useless one, cannot decrease R-squared. That is why we also look at **adjusted R-squared**, which penalizes the score for adding predictors that do not really help.

```{r r-squared-compare-blank, eval=FALSE}
# Fill in the model names from the four models you have already fit
tibble(
  model = c("age only", "age + height", "age + height + smoke", "age + height + smoke + sex"),
  r_squared = c(
    summary(_______)$r.squared,
    summary(_______)$r.squared,
    summary(_______)$r.squared,
    summary(_______)$r.squared
  ),
  adj_r_squared = c(
    summary(_______)$adj.r.squared,
    summary(_______)$adj.r.squared,
    summary(_______)$adj.r.squared,
    summary(_______)$adj.r.squared
  )
)
```

**Solution:**

```{r r-squared-compare}
tibble(
  model = c("age only", "age + height", "age + height + smoke", "age + height + smoke + sex"),
  r_squared = c(
    summary(age_model)$r.squared,
    summary(age_height_model)$r.squared,
    summary(full_model_1)$r.squared,
    summary(full_model_2)$r.squared
  ),
  adj_r_squared = c(
    summary(age_model)$adj.r.squared,
    summary(age_height_model)$adj.r.squared,
    summary(full_model_1)$adj.r.squared,
    summary(full_model_2)$adj.r.squared
  )
)
```

**Question:** Does R-squared increase every time a predictor is added? Does adjusted R-squared also increase every time?

Ans:

**Solution answer:** Yes, R-squared increases every single time a predictor is added (0.572 -> 0.768 -> 0.768 -> 0.775, roughly). Adjusted R-squared also increases each time here, but by a much smaller amount than R-squared when the added predictor (like `smoke`) is only weakly useful, and it can in principle go down if a predictor adds no real information.

**Question:** Based on this comparison, do you think `sex` is a genuinely useful predictor to add to the age + height + smoke model, or does it look more like noise? Explain using the numbers.

Ans:

**Solution answer:** Sex looks genuinely useful. Adjusted R-squared rises noticeably when `sex` is added (from about 0.7665 to about 0.774), and the `sexmale` coefficient itself is highly statistically significant (p < 0.001). This is a clearer improvement than adding `smoke`, whose own coefficient was never significant.

# Part 4: Predicting with the full model

Use `full_model_2` (the four-predictor model) to predict FEV for a new person.

Fill in the blanks to predict FEV for a 10-year-old, 57-inch-tall, female, non-smoker.

```{r predict-new-person-blank, eval=FALSE}
new_person <- tibble(
  age_yrs = _______,
  height_in = _______,
  smoke = factor("_______", levels = c("non_smoker", "smoker")),
  sex = factor("_______", levels = c("female", "male"))
)

predict(_______, newdata = new_person)
```

**Solution:**

```{r predict-new-person}
new_person <- tibble(
  age_yrs = 10,
  height_in = 57,
  smoke = factor("non_smoker", levels = c("non_smoker", "smoker")),
  sex = factor("female", levels = c("female", "male"))
)

predict(full_model_2, newdata = new_person)
```

**Question:** What is the predicted FEV for this person, rounded to 2 decimals?

Ans:

**Solution answer:** The predicted FEV is about **2.14 litres**.

**Question:** Now predict FEV for the same person, but change only `smoke` to `"smoker"`. How much does the prediction change? Does this match the size of the `smokesmoker` coefficient from `full_model_2`?

Ans:

**Solution answer:**

```{r predict-new-person-smoker}
new_person_smoker <- tibble(
  age_yrs = 10,
  height_in = 57,
  smoke = factor("smoker", levels = c("non_smoker", "smoker")),
  sex = factor("female", levels = c("female", "male"))
)

predict(full_model_2, newdata = new_person_smoker)
```

The prediction drops by about 0.087 litres, from about 2.14 to about 2.05. This matches the `smokesmoker` coefficient from `full_model_2` exactly, since changing only one predictor's value shifts the prediction by exactly that predictor's coefficient.

# Part 5 (bonus, optional): Does the height-FEV slope depend on sex?

So far, every model in this workshop has assumed each predictor's slope is the same for everyone (an **additive** model). An **interaction term** relaxes that assumption, letting one predictor's slope change depending on the value of another predictor.

```r
lm(outcome ~ x + group + x:group, data = my_data)
```

Fit an interaction model between `height_in` and `sex`.

```{r interaction-model-blank, eval=FALSE}
interaction_model <- lm(fev_l ~ height_in + sex + _______, data = fev)
summary(interaction_model)
```

**Solution:**

```{r interaction-model}
interaction_model <- lm(fev_l ~ height_in + sex + height_in:sex, data = fev)
summary(interaction_model)
```

**Question:** The coefficient table has an `(Intercept)`, a `height_in` row, a `sexmale` row, and a `height_in:sexmale` row. Which row is the height slope for girls (the reference sex)?

Ans:

**Solution answer:** The plain `height_in` row (about 0.112) is the height slope for girls, the reference level of `sex`.

**Question:** The `height_in:sexmale` coefficient is about 0.027. What is the height slope for boys? (Add it to the girls' slope.)

Ans:

**Solution answer:** The height slope for boys is 0.112 + 0.027 = **about 0.140**.

**Question:** Based on this, does the association between height and FEV appear to be about the same for boys and girls, or does it differ meaningfully? Use the two slopes to justify your answer.

Ans:

**Solution answer:** The slopes differ somewhat: about 0.112 litres per inch for girls versus about 0.140 for boys, and the interaction term itself is statistically significant (p < 0.001). This suggests that height is associated with a slightly larger increase in FEV per inch among boys than among girls in this dataset.

# Wrap-up

In this workshop, we moved from single-predictor models to a four-predictor multiple regression model of FEV.

Key ideas to take away:

1. Each coefficient in a multiple regression is interpreted **while holding the other predictors constant**.
2. Adding predictors that are correlated with each other (like age and height) can change, and even sign-flip, the other coefficients. This is confounding at work.
3. R-squared always goes up when you add a predictor. Adjusted R-squared is a fairer way to judge whether a new predictor is worth keeping.
4. A fitted multiple regression equation can be used for prediction by plugging in a value for every predictor.
5. (Bonus) An interaction term lets a slope differ across groups, instead of assuming one slope fits everyone.
