---
title: "SOLUTIONS: Estimating the smoking effect on FEV"
subtitle: "Effect sizes, confounding, and prediction"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

These are the completed solutions for the no-interaction effect-size version of workshop 8.

# Load packages and data

*Exercise chunk (blank):*

```{r load-packages-q, eval=FALSE}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

**Solution:**

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, modelsummary, ggstatsplot, equatiomatic)
pacman::p_load_gh("the-graph-courses/inspectdf")
```

```{r load-data}
fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint answer:** `fev` has **654 rows and 5 columns**.

## Set the reference levels

*Exercise chunk (blank):*

```{r set-factors-q, eval=FALSE}
fev <- fev %>%
  mutate(
    smoke = factor(smoke, levels = c("non_smoker", "smoker")),
    sex = factor(sex, levels = c(________, ________))
  )

levels(fev$smoke)
levels(fev$sex)
```

**Solution:**

```{r set-factors}
fev <- fev %>%
  mutate(
    smoke = factor(smoke, levels = c("non_smoker", "smoker")),
    sex = factor(sex, levels = c("female", "male"))
  )

levels(fev$smoke)
levels(fev$sex)
```

**Answer:** No, it changes nothing here. R's default is to use the alphabetically first level as the reference, and in this dataset the levels we want happen to already be alphabetically first (`"non_smoker"` before `"smoker"`, `"female"` before `"male"`). So the coefficients would be identical without this step.

It is still worth doing for two reasons. First, it makes the baseline an explicit, readable choice rather than an accident of spelling. Second, it is robust: if someone later recoded `"non_smoker"` as `"unexposed"`, or `"smoker"` as `"active_smoker"`, the alphabetical default would silently flip the reference group and reverse the sign of the smoking coefficient, with no error and no warning. Setting `levels` protects against that.

# Part 1: Define the effect and the confounding problem

## Part 1a: The raw smoking difference

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    mean_fev = mean(fev_l),
    mean_height = mean(height_in),
    mean_age = mean(age_yrs),
    .groups = "drop"
  )
```

```{r smoke-plot}
ggbetweenstats(
  data = fev,
  x = smoke,
  y = fev_l
)
```

```{r model-1}
m1_smoke <- lm(fev_l ~ smoke, data = fev)
summary(m1_smoke)
```

```{r model-1-equation}
extract_eq(m1_smoke) %>% preview_eq()
```

**Answer:** Smokers have higher mean FEV in the raw summary and plot.

**Answer:** Smokers are older and taller on average.

**Answer:** The `smokesmoker` coefficient is about **0.711**. In the raw model, smokers have predicted FEV about **0.711 litres higher** than non-smokers, on average.

**Answer:** The raw comparison is misleading because smokers are also older and taller, and older, taller youths naturally have higher FEV.

## Part 1b: Confounders and a simple DAG

Age, height, and sex are all plausible confounders: each may affect both whether a youth smokes and their FEV. We choose them from the DAG, before fitting any models.

# Part 2: Fit a sequence of adjusted models

## Part 2a: Model 2, add age

*Exercise chunk (blank):*

```{r model-2-q, eval=FALSE}
m2_age <- 
summary(m2_age)
```

**Solution:**

```{r model-2}
m2_age <- lm(fev_l ~ smoke + age_yrs, data = fev)
summary(m2_age)
```

**Answer:** After adjusting for age, the smoking coefficient flips from about **+0.711** to about **-0.209**. The positive raw association was largely explained by smokers being older.

## Part 2b: Model 3, add height

*Exercise chunk (blank):*

```{r model-3-q, eval=FALSE}
m3_age_height <- 
summary(m3_age_height)
```

**Solution:**

```{r model-3}
m3_age_height <- lm(fev_l ~ smoke + age_yrs + height_in, data = fev)
summary(m3_age_height)
```

**Answer:** Holding age and height constant, smokers have predicted FEV about **0.110 litres lower** than non-smokers, on average.

## Part 2c: Model 4, add sex

*Exercise chunk (blank):*

```{r model-4-q, eval=FALSE}
m4_age_height_sex <- 
summary(m4_age_height_sex)
```

**Solution:**

```{r model-4}
m4_age_height_sex <- lm(fev_l ~ smoke + age_yrs + height_in + sex, data = fev)
summary(m4_age_height_sex)
```

*Exercise chunk (blank):*

```{r model-4-equation-q, eval=FALSE}
extract_eq(________) %>% preview_eq()
```

**Solution:**

```{r model-4-equation}
extract_eq(m4_age_height_sex) %>% preview_eq()
```

**Answer:** In Model 1 the equation has a single predictor, so the `smoke_smoker` coefficient is just the raw difference in mean FEV between the two groups. In Model 4 the equation has age, height, and sex terms alongside it, so the `smoke_smoker` coefficient now represents the difference between smokers and non-smokers *at fixed values of those other three variables*. The symbol in the equation looks the same; the quantity it estimates does not.

**Answer:** No. The p-value is about **0.14**, so at alpha = 0.05 we fail to reject the null hypothesis for the adjusted smoking coefficient.

# Part 3: Compare models side by side

```{r model-comparison}
model_list <- list(
  "1. Smoking only" = m1_smoke,
  "2. + age" = m2_age,
  "3. + height" = m3_age_height,
  "4. + sex" = m4_age_height_sex
)

modelsummary(
  model_list,
  estimate = "{estimate}",
  statistic = c("[{conf.low}, {conf.high}]", "p = {p.value}"),
  conf_level = 0.95,
  fmt = 3,
  coef_map = c(
    "smokesmoker" = "Smoking (smoker vs non-smoker)",
    "age_yrs" = "Age (years)",
    "height_in" = "Height (inches)",
    "sexmale" = "Sex (male vs female)"
  ),
  gof_map = list(
    list(raw = "nobs", clean = "N (youths)", fmt = 0),
    list(raw = "r.squared", clean = "R-squared", fmt = 3),
    list(raw = "adj.r.squared", clean = "Adjusted R-squared", fmt = 3)
  ),
  title = "Estimated smoking effect on FEV (litres), by adjustment set",
  notes = "Each cell shows the estimate, the 95% confidence interval, and the p-value. Outcome is FEV in litres.",
  output = "gt"
)
```

**Answer:** R-squared increases sharply from about **0.06** in the smoking-only model to about **0.78** in the final model. Adjusted R-squared behaves the same way here, rising from about **0.06** to about **0.77**, because each confounder we add genuinely improves the fit.

**Answer:** In Model 1 the smoking interval is **[0.495, 0.927]**, which lies entirely above zero, matching the very small p-value. In Model 4 it is **[-0.204, 0.029]**, which contains zero, matching p = 0.14. The two ways of reading significance always agree, because they are computed from the same estimate and standard error. Note how much the interval moves once we adjust: it does not merely widen, it shifts across zero to the other side.

# Part 4: Report the result

**Example answer:**

> In the raw comparison, smokers had about **0.71 litres higher** mean FEV than non-smokers. After adjusting for age, height, and sex, smokers had mean FEV about **0.09 litres lower** than non-smokers (95% CI -0.20 to 0.03), so the interval includes zero. Each additional year of age was associated with about **0.07 litres higher** FEV (95% CI 0.05 to 0.08), each additional inch of height with about **0.10 litres higher** FEV (95% CI 0.10 to 0.11), and males had about **0.16 litres higher** FEV than females (95% CI 0.09 to 0.22). The final model explained about **77%** of the variation in FEV (adjusted R-squared). Because this is observational data, we describe these as adjusted associations rather than proof that smoking caused lower lung function.

**Marking notes:**

- Learners should state the adjustment set in the same sentence as the smoking estimate.
- The smoking estimate should be in litres with its 95% CI, not just "not significant".
- An interval that includes zero should be described plainly, not hidden behind p-value language alone.
- Push back on causal wording such as "smoking reduced FEV". Association language is appropriate here.
- Adjusted R-squared should appear somewhere in the paragraph.
- Brief interpretation of the other coefficients is a bonus, not required for full credit.

# Optional challenge: Prediction from the adjusted model

## Challenge part a: Predict with `predict()`

*Exercise chunk (blank):*

```{r predict-nonsmoker-q, eval=FALSE}
new_nonsmoker <- tibble(
  smoke = "__________",
  age_yrs = ______,
  height_in = ______,
  sex = "____"
)

predict(__________, newdata = __________)
```

**Solution:**

```{r predict-nonsmoker}
new_nonsmoker <- tibble(
  smoke = "non_smoker",
  age_yrs = 12,
  height_in = 60,
  sex = "male"
)

predict(m4_age_height_sex, newdata = new_nonsmoker)
```

*Exercise chunk (blank):*

```{r predict-smoker-q, eval=FALSE}
new_smoker <- tibble(
  smoke = "______",
  age_yrs = ______,
  height_in = ______,
  sex = "____"
)

predict(__________, newdata = __________)
```

**Solution:**

```{r predict-smoker}
new_smoker <- tibble(
  smoke = "smoker",
  age_yrs = 12,
  height_in = 60,
  sex = "male"
)

predict(m4_age_height_sex, newdata = new_smoker)
```

**Answer:** The smoker prediction is about **0.087 litres lower** than the non-smoker prediction. That gap matches the smoking coefficient in Model 4 because the two people have the same age, height, and sex.

## Challenge part b: Build the prediction by hand

```{r coefficients}
coef(m4_age_height_sex)
```

*Exercise chunk (blank):*

```{r prediction-by-hand-q, eval=FALSE}
b <- coef(m4_age_height_sex)

b["(Intercept)"] +
  b["height_in"] * ______ +
  b["age_yrs"] * ______ +
  b["sexmale"] * ______ +
  b["smokesmoker"] * ______
```

**Solution:**

```{r prediction-by-hand}
b <- coef(m4_age_height_sex)

b["(Intercept)"] +
  b["height_in"] * 60 +
  b["age_yrs"] * 12 +
  b["sexmale"] * 1 +
  b["smokesmoker"] * 1
```

**Answer:** Yes. It matches the `predict()` result for the 12-year-old, 60-inch-tall male smoker.

# Wrap-up

In this workshop, multiple regression was used as an effect-size estimation tool. The smoking coefficient changed from a misleading positive raw association to a small negative adjusted association after accounting for age, height, and sex. Reported with its confidence interval, that adjusted estimate is -0.087 litres (95% CI -0.204 to 0.029), an interval that includes zero.
