---
title: "SOLUTIONS: Estimating the smoking effect on FEV"
subtitle: "Effect sizes, confounding, and prediction"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

These are the completed solutions for the no-interaction effect-size version of workshop 8.

# Load packages and data

*Exercise chunk (blank):*

```{r load-packages-q, eval=FALSE}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

**Solution:**

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, modelsummary, ggstatsplot)
pacman::p_load_gh("the-graph-courses/inspectdf")
```

```{r load-data}
fev <- read_csv(here("rsr_week_08_workshop/data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint answer:** `fev` has **654 rows and 5 columns**.

# Part 1: Define the effect and the confounding problem

## Part 1a: The raw smoking difference

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    mean_fev = mean(fev_l),
    mean_height = mean(height_in),
    mean_age = mean(age_yrs),
    .groups = "drop"
  )
```

```{r smoke-plot}
ggbetweenstats(
  data = fev,
  x = smoke,
  y = fev_l
)
```

```{r model-1}
m1_smoke <- lm(fev_l ~ smoke, data = fev)
summary(m1_smoke)
```

**Answer:** Smokers have higher mean FEV in the raw summary and plot.

**Answer:** Smokers are older and taller on average.

**Answer:** The `smokesmoker` coefficient is about **0.711**. In the raw model, smokers have predicted FEV about **0.711 litres higher** than non-smokers, on average.

**Answer:** The raw comparison is misleading because smokers are also older and taller, and older, taller youths naturally have higher FEV.

## Part 1b: Confounders, mediators, and a simple DAG

**Answer:**

- **Neighbourhood of residence:** confounder, include. Neighbourhood may be related to smoking rates and to environmental factors that affect lung function.
- **Parental smoking status:** confounder, include. Parental smoking can influence whether a youth smokes and may also affect FEV through shared environment or genetics.
- **Frequency of respiratory infections:** mediator, exclude. Infections can be a consequence of smoking and can lower FEV, so they sit on the causal pathway.

# Part 2: Fit a sequence of adjusted models

## Part 2a: Model 2, add age

*Exercise chunk (blank):*

```{r model-2-q, eval=FALSE}
m2_age <- 
summary(m2_age)
```

**Solution:**

```{r model-2}
m2_age <- lm(fev_l ~ smoke + age_yrs, data = fev)
summary(m2_age)
```

**Answer:** After adjusting for age, the smoking coefficient flips from about **+0.711** to about **-0.209**. The positive raw association was largely explained by smokers being older.

## Part 2b: Model 3, add height

*Exercise chunk (blank):*

```{r model-3-q, eval=FALSE}
m3_age_height <- 
summary(m3_age_height)
```

**Solution:**

```{r model-3}
m3_age_height <- lm(fev_l ~ smoke + age_yrs + height_in, data = fev)
summary(m3_age_height)
```

**Answer:** Holding age and height constant, smokers have predicted FEV about **0.110 litres lower** than non-smokers, on average.

## Part 2c: Model 4, add sex

*Exercise chunk (blank):*

```{r model-4-q, eval=FALSE}
m4_age_height_sex <- 
summary(m4_age_height_sex)
```

**Solution:**

```{r model-4}
m4_age_height_sex <- lm(fev_l ~ smoke + age_yrs + height_in + sex, data = fev)
summary(m4_age_height_sex)
```

**Answer:** Holding age, height, and sex constant, smokers have predicted FEV about **0.087 litres lower** than non-smokers, on average.

**Answer:** No. The p-value is about **0.12**, so at alpha = 0.05 we fail to reject the null hypothesis for the adjusted smoking coefficient.

# Part 3: Compare models side by side

```{r model-comparison}
model_list <- list(
  "1. Smoking only" = m1_smoke,
  "2. + age" = m2_age,
  "3. + height" = m3_age_height,
  "4. + sex" = m4_age_height_sex
)

modelsummary(
  model_list,
  coef_map = c(
    "smokesmoker" = "Smoking (smoker vs non-smoker)",
    "age_yrs" = "Age (years)",
    "height_in" = "Height (inches)",
    "sexmale" = "Sex (male vs female)"
  ),
  gof_map = c("nobs", "r.squared", "adj.r.squared")
)
```

**Answer:** R-squared increases sharply from about **0.06** in the smoking-only model to about **0.78** in the final model. Adjusted R-squared behaves the same way here, rising from about **0.06** to about **0.77**, because each confounder we add genuinely improves the fit.

**Answer:** Model 4. Our goal is to estimate the smoking effect after adjusting for confounders, not to maximise R-squared. Model 4 includes exactly the variables we identified as confounders in Part 1b (age, height, sex), so it is the appropriate main estimate. The smoking coefficient is similar to Model 3, so adding sex changes the estimate only slightly.

# Part 4: Prediction from the adjusted model

## Part 4a: Predict with `predict()`

*Exercise chunk (blank):*

```{r predict-nonsmoker-q, eval=FALSE}
new_nonsmoker <- tibble(
  smoke = "__________",
  age_yrs = ______,
  height_in = ______,
  sex = "____"
)

predict(__________, newdata = __________)
```

**Solution:**

```{r predict-nonsmoker}
new_nonsmoker <- tibble(
  smoke = "non_smoker",
  age_yrs = 12,
  height_in = 60,
  sex = "male"
)

predict(m4_age_height_sex, newdata = new_nonsmoker)
```

*Exercise chunk (blank):*

```{r predict-smoker-q, eval=FALSE}
new_smoker <- tibble(
  smoke = "______",
  age_yrs = ______,
  height_in = ______,
  sex = "____"
)

predict(__________, newdata = __________)
```

**Solution:**

```{r predict-smoker}
new_smoker <- tibble(
  smoke = "smoker",
  age_yrs = 12,
  height_in = 60,
  sex = "male"
)

predict(m4_age_height_sex, newdata = new_smoker)
```

**Answer:** The smoker prediction is about **0.087 litres lower** than the non-smoker prediction. That gap matches the smoking coefficient in Model 4 because the two people have the same age, height, and sex.

## Part 4b: Build the prediction by hand

```{r coefficients}
coef(m4_age_height_sex)
```

*Exercise chunk (blank):*

```{r prediction-by-hand-q, eval=FALSE}
b <- coef(m4_age_height_sex)

b["(Intercept)"] +
  b["height_in"] * ______ +
  b["age_yrs"] * ______ +
  b["sexmale"] * ______ +
  b["smokesmoker"] * ______
```

**Solution:**

```{r prediction-by-hand}
b <- coef(m4_age_height_sex)

b["(Intercept)"] +
  b["height_in"] * 60 +
  b["age_yrs"] * 12 +
  b["sexmale"] * 1 +
  b["smokesmoker"] * 1
```

**Answer:** Yes. It matches the `predict()` result for the 12-year-old, 60-inch-tall male smoker.

# Part 5: Report the result

**Example answer:** In the raw comparison, smokers had about **0.71 litres higher** mean FEV than non-smokers. After adjusting for age, height, and sex, the estimated smoking effect was about **-0.09 litres** (95% CI spans zero; p = 0.12). The final model explained about **77%** of the variation in FEV (adjusted R-squared). Because this is observational data, we describe this as an adjusted association rather than proof that smoking caused lower lung function.

# Wrap-up

In this workshop, multiple regression was used as an effect-size estimation tool. The smoking coefficient changed from a misleading positive raw association to a small negative adjusted association after accounting for age, height, and sex.
