---
title: "DEMO: A four-step recipe for a publication table"
subtitle: "PREVEND systolic blood pressure by participant characteristic"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This demo gives you the recipe you will use in the exercise. We will ask whether systolic blood pressure differs by sex, age group, education, and statin use in the PREVEND teaching dataset.

The same four steps appear in the exercise:

1. Import and label the data.
2. Calculate group summaries.
3. Run the appropriate tests.
4. Assemble a publication-style table.

## Load packages

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, gt, janitor)
pacman::p_load(openintro)
```

## Step 1: Import and label the data

First we import and take a look at the data.

```{r data-preview}
prevend_raw <- openintro::prevend.samp
```

```{r glimpse-data}
glimpse(prevend_raw)
```

Before calculating means and standard deviations, check for missing values. Running `summary()` on the whole dataset reports a count of missing values for each variable on the `NA's` line.

```{r missing-check}
prevend_raw %>%
  select(SBP, Gender, Age, Education, Statin) %>%
  summary()
```

In this teaching dataset, `Statin` is a clean binary variable with no missing values.

We still need to select the variables, give them readable labels, and convert age to groups.

```{r import}
prevend <- prevend_raw %>%
  mutate(
    Gender = factor(
      Gender,
      levels = c(0, 1),
      labels = c("Male", "Female")
    ),
    Age = cut(
      Age,
      breaks = c(35, 50, 65, Inf),
      labels = c("36-50", "51-65", "66+")
    ),
    Education = factor(
      Education,
      levels = c(0, 1, 2, 3),
      labels = c(
        "Primary school",
        "Lower secondary",
        "Higher secondary",
        "University"
      )
    ),
    Statin = factor(
      Statin,
      levels = c(0, 1),
      labels = c("No statin", "Statin")
    )
  ) %>%
  select(SBP, Gender, Age, Education, Statin)

nrow(prevend)
```

Call the `tabyl` function to check the four categorical variables before continuing.

```{r group-checks}
tabyl(prevend, Gender)
tabyl(prevend, Age)
tabyl(prevend, Education)
tabyl(prevend, Statin)
```

## Step 2: Descriptive stats for every group

Now, we will calculate the average systolic blood pressure by each of the four categorical variables, as well as the standard deviation, standard error, and 95% confidence interval.

First, we reshape the data to long form so each row is one person in one group of one characteristic.

```{r reshape}
prevend_long <- prevend %>%
  pivot_longer(
    cols = -SBP,
    names_to = "variable",
    values_to = "group"
  ) %>%
  filter(!is.na(group), !is.na(SBP))

prevend_long
```

Using this, we can easily calculate the summary statistics for every group in one pipeline. For example, below we calculate the number of people in each group.

```{r count-example}
prevend_long %>%
  group_by(variable, group) %>%
  summarise(
    n = n()
  )
```

Now let's do something like that but for all the descriptive statistics we want to calculate:

```{r descriptives}
group_stats <- prevend_long %>%
  group_by(variable, group) %>%
  summarise(
    n = n(),
    mean = mean(SBP),
    sd = sd(SBP),
    se = sd / sqrt(n),
    ci_95_low_end = mean - qt(0.975, df = n - 1) * se,
    ci_95_high_end = mean + qt(0.975, df = n - 1) * se,
    .groups = "drop"
  ) %>%
  mutate(
    variable = factor(
      variable,
      levels = c("Gender", "Age", "Education", "Statin"),
      labels = c("Sex", "Age group", "Education", "Statin use")
    )
  ) %>%
  arrange(variable, group)

group_stats
```

## Step 3: The tests

Next, we want to test for differences in systolic blood pressure between the groups.

We will use a t-test for variables with two groups and a one-way ANOVA for variables with three or more groups.

In these tests, we will use `var.equal = TRUE`, which assumes equal variance between the groups.

```{r tests}
sex_t <- t.test(SBP ~ Gender, data = prevend, var.equal = TRUE)

age_aov <- aov(SBP ~ Age, data = prevend)

education_aov <- aov(SBP ~ Education, data = prevend)

statin_t <- t.test(SBP ~ Statin, data = prevend, var.equal = TRUE)
```

Look at each output.

```{r test-output}
sex_t
summary(age_aov)
summary(education_aov)
statin_t
```

## Step 4: Assemble the table with `gt`

The code below turns your summaries and tests into a publication-style table.

The formatting code is longer than the statistics code. That is normal. Your job is to make sure the formatted table faithfully reports the tests above.

First, we extract the ANOVA F statistics and p-values.

```{r anova-extract}
age_f <- summary(age_aov)[[1]]$`F value`[1]
age_p <- summary(age_aov)[[1]]$`Pr(>F)`[1]

education_f <- summary(education_aov)[[1]]$`F value`[1]
education_p <- summary(education_aov)[[1]]$`Pr(>F)`[1]

age_f
age_p
```

Then, we assemble the table with the `gt` package.

```{r build-table}
results_table <- group_stats %>%
  mutate(
    `Mean (SD)` = paste0(
      format(round(mean, 1), nsmall = 1),
      " (",
      format(round(sd, 1), nsmall = 1),
      ")"
    ),
    `95% CI` = paste0(
      format(round(ci_95_low_end, 2), nsmall = 2),
      ", ",
      format(round(ci_95_high_end, 2), nsmall = 2)
    ),
    statistic = case_when(
      variable == "Sex" ~ as.numeric(sex_t$statistic),
      variable == "Age group" ~ age_f,
      variable == "Education" ~ education_f,
      variable == "Statin use" ~ as.numeric(statin_t$statistic)
    ),
    p_value = case_when(
      variable == "Sex" ~ sex_t$p.value,
      variable == "Age group" ~ age_p,
      variable == "Education" ~ education_p,
      variable == "Statin use" ~ statin_t$p.value
    ),
    stat_text = as.character(round(statistic, 2)),
    p_text = if_else(p_value < 0.001, "<0.001", as.character(round(p_value, 3)))
  ) %>%
  group_by(variable) %>%
  mutate(row_in_block = row_number()) %>%
  ungroup() %>%
  mutate(
    `t/F` = if_else(row_in_block == 1, stat_text, ""),
    p = if_else(row_in_block == 1, p_text, "")
  )
```

Now, we render the table with the `gt` package.

```{r render-table}
results_table %>%
  select(variable, group, `Mean (SD)`, `95% CI`, `t/F`, p) %>%
  gt(groupname_col = "variable", rowname_col = "group") %>%
  tab_header(title = "Systolic blood pressure by participant characteristic") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  tab_options(table.width = pct(100))
```

## Reading the results

Sex has two groups, so the table reports a t statistic. Age group and education have more than two groups, so the table reports F statistics from ANOVA. A significant ANOVA tells us that at least one group mean differs. It does not tell us which pair of groups differs. For that, we would need a follow-up test such as Tukey's HSD.
