---
title: "EXERCISE: Estimating the smoking effect on FEV"
subtitle: "Effect sizes, confounding, interaction, and prediction"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In week 6, we used the FEV data to learn what regression is for: describing relationships, predicting outcomes, and controlling for another variable.

This week we return to the same dataset with a clearer focus on **effect size estimation**. Our main question is:

> What is the estimated effect of smoking on FEV in children and adolescents?

The dataset contains 654 youths aged 3 to 19 from East Boston in the 1970s. The outcome is **forced expiratory volume** (`fev_l`), measured in litres. Higher values mean better lung function.

The variable `smoke` records current smoking status (`non_smoker` or `smoker`). We will treat the smoking coefficient as an **effect size estimate**: the estimated difference in mean FEV between smokers and non-smokers, in litres.

In multiple linear regression, each coefficient is interpreted while holding the other predictors in the model constant. Because this is observational data, the raw smoker versus non-smoker difference can be confounded. We will start with smoking as the only predictor, then add plausible confounders one at a time and track how the estimated smoking effect, R-squared, and adjusted R-squared change.

After the main effect-size models, we will briefly ask whether the smoking association differs by sex.

# Load packages and data

Use `p_load()` to load `tidyverse`, `here`, `broom`, and `gt`.

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

```{r load-data}
fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint:** `fev` should have **654 rows and 5 columns**.

# Part 1: Define the effect and the confounding problem

## Part 1a: The raw smoking difference

Start by summarising the outcome and likely confounders by smoking status.

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    n = n(),
    mean_age = mean(age_yrs),
    mean_height = mean(height_in),
    mean_fev = mean(fev_l),
    .groups = "drop"
  )
```

**Question:** Which group has higher mean FEV in the raw summary?

Ans:

**Question:** Which group is older and taller on average?

Ans:

**Question:** In one sentence, why might the raw comparison be misleading?

Ans:

## Part 1b: Confounders, mediators, and a simple DAG

When we estimate the effect of smoking on FEV, we ideally adjust for variables that **precede** smoking and could confound the association. A **confounder** is related to both the exposure and the outcome.

In this dataset, age, height, and sex are plausible confounders:

- older and taller youths are more likely to smoke
- age, height, and sex are all strongly related to lung size and FEV

One possible DAG is:

```text
age ----------> smoking
  \              |
   \             v
    ----------> FEV

height -------> smoking
  \              |
   \             v
    ----------> FEV

sex ----------> smoking
  \              |
   \             v
    ----------> FEV
```

We generally do **not** want to adjust for a **mediator**, a variable on the causal pathway from smoking to FEV. Adjusting for a mediator can block part of the effect we are trying to estimate.

For example:

```text
smoking -----> airway irritation -----> FEV
```

If `airway_irritation` were measured after smoking started, adjusting for it would partly remove the smoking effect rather than isolate it.

**Question:** In this dataset, why are age, height, and sex reasonable variables to adjust for when estimating the smoking effect?

Ans:

**Question:** Give one example of a variable that would be on the pathway from smoking to FEV and should usually not be adjusted for if our goal is the total effect of smoking.

Ans:

# Part 2: Fit a sequence of effect-size models

We will fit models that all have the same outcome (`fev_l`) and exposure (`smoke`). With each model, we add one adjustment variable.

The smoking coefficient is our estimated smoking effect, in litres.

## Part 2a: Model 1, smoking only

```{r model-1}
m1_smoke <- lm(________ ~ ________, data = fev)
summary(m1_smoke)
```

**Question:** What is the `smokesmoker` coefficient? Interpret it as a raw effect size estimate.

Ans:

## Part 2b: Model 2, add age

Age is a plausible confounder because smokers are older on average, and FEV increases strongly with age.

```{r model-2}
m2_age <- lm(fev_l ~ smoke + ________, data = fev)
summary(m2_age)
```

**Question:** What happened to the smoking coefficient after adjusting for age?

Ans:

## Part 2c: Model 3, add height

Height is also a plausible confounder because smokers are taller on average, and height is strongly related to lung volume.

```{r model-3}
m3_age_height <- lm(fev_l ~ smoke + age_yrs + ________, data = fev)
summary(m3_age_height)
```

**Question:** Interpret the smoking coefficient in this model using the phrase "holding age and height constant."

Ans:

## Part 2d: Model 4, add sex

Finally, add sex. Sex may be related to body size and lung function, and it could also differ between smokers and non-smokers.

```{r model-4}
m4_age_height_sex <- lm(fev_l ~ smoke + age_yrs + height_in + ______, data = fev)
summary(m4_age_height_sex)
```

**Question:** Interpret the smoking coefficient in this model.

Ans:

**Question:** At alpha = 0.05, is the adjusted smoking coefficient statistically significant in Model 4?

Ans:

# Part 3: Track effect sizes and model fit

## Part 3a: Build a comparison table

The helper code below extracts the smoking coefficient and model fit statistics from each model.

```{r model-comparison}
model_list <- list(
  "1. Smoking only" = m1_smoke,
  "2. + age" = m2_age,
  "3. + height" = m3_age_height,
  "4. + sex" = m4_age_height_sex
)

effect_table <- imap_dfr(
  model_list,
  ~ tidy(.x, conf.int = TRUE) %>%
    filter(term == "smokesmoker") %>%
    transmute(
      model = .y,
      smoking_beta = estimate,
      ci_low = conf.low,
      ci_high = conf.high,
      p_value = p.value
    )
)

fit_table <- imap_dfr(
  model_list,
  ~ glance(.x) %>%
    transmute(
      model = .y,
      r_squared = r.squared,
      adjusted_r_squared = adj.r.squared
    )
)

model_summary <- left_join(effect_table, fit_table, by = "model")
model_summary
```

**Question:** Across the four models, how does the estimated smoking effect change?

Ans:

**Question:** What happens to R-squared and adjusted R-squared as we add predictors?

Ans:

**Question:** Why does adjusted R-squared usually increase more slowly than R-squared when we add predictors?

Ans:

## Part 3b: Format the comparison table

```{r formatted-comparison}
model_summary %>%
  mutate(
    `Smoking beta (95% CI)` = paste0(
      format(round(smoking_beta, 3), nsmall = 3),
      " (",
      format(round(ci_low, 3), nsmall = 3),
      " to ",
      format(round(ci_high, 3), nsmall = 3),
      ")"
    ),
    p = if_else(p_value < 0.001, "<0.001", as.character(round(p_value, 3))),
    `R-squared` = round(r_squared, 3),
    `Adjusted R-squared` = round(adjusted_r_squared, 3)
  ) %>%
  select(model, `Smoking beta (95% CI)`, p, `R-squared`, `Adjusted R-squared`) %>%
  gt() %>%
  tab_header(
    title = "Estimated smoking effect on FEV",
    subtitle = "Outcome: FEV in litres"
  )
```

**Question:** Which model would you choose as the main estimate of the smoking effect? Why?

Ans:

# Part 4: Brief interaction check

So far, every model assumed the smoking effect is the same for males and females. That is an **additive** model.

An **interaction** asks whether the smoking association differs across levels of another predictor. For example:

> Is the smoking association with FEV different for males and females?

This model adds a `smoke:sex` interaction while still adjusting for age and height:

```{r interaction-model}
m5_interaction <- lm(
  fev_l ~ smoke * sex + age_yrs + height_in,
  data = fev
)

summary(m5_interaction)
```

The `smoke * sex` shortcut expands to:

```r
smoke + sex + smoke:sex
```

**Question:** What is the p-value for the `smokesmoker:sexmale` interaction term?

Ans:

**Question:** In this interaction model, why is the plain `smokesmoker` coefficient not the overall smoking effect?

Ans:

## Part 4a: Compare predicted smoking gaps by sex

Use the interaction model to compare predicted FEV for smokers and non-smokers among 12-year-old, 60-inch-tall youths, separately for males and females.

```{r interaction-predictions}
interaction_profiles <- expand_grid(
  smoke = c("non_smoker", "smoker"),
  sex = c("female", "male"),
  age_yrs = 12,
  height_in = 60
)

interaction_profiles %>%
  mutate(predicted_fev = predict(________, newdata = interaction_profiles)) %>%
  arrange(sex, smoke)
```

**Question:** Is the predicted smoker versus non-smoker gap the same for males and females?

Ans:

**Question:** Should this interaction model replace Model 4 as your main model, or should it be treated as a quick exploratory check? Explain briefly.

Ans:

# Part 5: Prediction from the adjusted additive model

Return to Model 4, the adjusted additive model, to predict FEV for two otherwise similar 12-year-old, 60-inch-tall males: one non-smoker and one smoker.

## Part 5a: Predict with `predict()`

```{r predict-nonsmoker}
new_nonsmoker <- tibble(
  smoke = "__________",
  age_yrs = ______,
  height_in = ______,
  sex = "____"
)

predict(__________, newdata = __________)
```

```{r predict-smoker}
new_smoker <- tibble(
  smoke = "______",
  age_yrs = ______,
  height_in = ______,
  sex = "____"
)

predict(__________, newdata = __________)
```

**Question:** How far apart are the two predictions? How does that difference relate to the smoking coefficient in Model 4?

Ans:

## Part 5b: Build the prediction by hand

```{r coefficients}
coef(m4_age_height_sex)
```

```{r prediction-by-hand}
b <- coef(m4_age_height_sex)

b["(Intercept)"] +
  b["height_in"] * ______ +
  b["age_yrs"] * ______ +
  b["sexmale"] * ______ +
  b["smokesmoker"] * ______
```

**Question:** Does your hand calculation match `predict()`?

Ans:

# Part 6: Report the result

Write a short results paragraph. Include:

- the raw smoking estimate
- the fully adjusted smoking estimate
- R-squared or adjusted R-squared for the final model
- one sentence about the interaction check
- a cautious interpretation because this is observational data

Your answer:

# Wrap-up

In this workshop, multiple regression was used as an effect-size estimation tool. The smoking coefficient changed from a misleading positive raw association to a small negative adjusted association after accounting for age, height, and sex. The interaction check showed how the estimated smoking association can be allowed to differ by sex, but interactions should usually be chosen from a clear research question rather than searched for automatically.

The most important phrase is: **holding the other variables constant**.
