---
title: "EXERCISE: Estimating the smoking effect on FEV"
subtitle: "Multiple regression, confounding, and adjusted effect sizes"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this workshop, our main question is:

> What is the estimated effect of smoking on FEV in children and adolescents?

The dataset contains 654 youths aged 3 to 19 from East Boston in the 1970s. The outcome is **forced expiratory volume** (`fev_l`), measured in litres. Higher values mean better lung function.

The variable `smoke` records current smoking status. We will treat the smoking coefficient as an **effect size estimate**: the estimated difference in mean FEV between smokers and non-smokers, in litres.

Because this is observational data, the raw smoker versus non-smoker difference can be confounded. We will start with a raw model, then add plausible confounders and track how the estimated smoking effect changes.

# Load packages and data

Use `p_load()` to load `tidyverse`, `here`, `broom`, and `gt`.

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

```{r load-data}
fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint:** `fev` should have **654 rows and 5 columns**.

# Part 1: Define the effect and the confounding problem

## Part 1a: The raw smoking difference

Start by summarising the outcome and likely confounders by smoking status.

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    n = n(),
    mean_age = mean(age_yrs),
    mean_height = mean(height_in),
    mean_fev = mean(fev_l),
    .groups = "drop"
  )
```

**Question:** Which group has higher mean FEV in the raw summary?

Ans:

**Question:** Which group is older and taller on average?

Ans:

## Part 1b: A simple DAG for confounding

A **confounder** is a variable that is related to both the exposure and the outcome. Here, age and height are plausible confounders because:

- older/taller youths may be more likely to smoke in this dataset
- older/taller youths naturally have larger lungs and higher FEV

One possible DAG is:

```text
age ----------> smoking
  \              |
   \             v
    ----------> FEV

height -------> smoking
  \              |
   \             v
    ----------> FEV
```

When estimating the effect of smoking, we usually want to adjust for variables that came **before** smoking or are not caused by smoking. We generally do **not** want to adjust for a mediator, a variable on the pathway from smoking to FEV.

For example:

```text
smoking -----> airway irritation -----> FEV
```

If `airway_irritation` were in the dataset, adjusting for it could remove part of the smoking effect we are trying to estimate.

**Question:** In this dataset, why are age and height reasonable variables to adjust for when estimating the smoking effect?

Ans:

**Question:** Give one example of a variable that would be on the pathway from smoking to FEV and should usually not be adjusted for if our goal is the total effect of smoking.

Ans:

# Part 2: Fit a sequence of effect-size models

We will fit models that all have the same outcome (`fev_l`) and exposure (`smoke`). With each model, we add one or more adjustment variables.

The smoking coefficient is our estimated smoking effect, in litres.

## Part 2a: Model 1, smoking only

```{r model-1}
m1_smoke <- lm(________ ~ ________, data = fev)
summary(m1_smoke)
```

**Question:** What is the `smokesmoker` coefficient? Interpret it as a raw effect size estimate.

Ans:

## Part 2b: Model 2, add age

Age is a plausible confounder because smokers are older on average, and FEV increases strongly with age.

```{r model-2}
m2_age <- lm(fev_l ~ smoke + ________, data = fev)
summary(m2_age)
```

**Question:** What happened to the smoking coefficient after adjusting for age?

Ans:

## Part 2c: Model 3, add height

Height is also a plausible confounder because smokers are taller on average, and height is strongly related to lung volume.

```{r model-3}
m3_age_height <- lm(fev_l ~ smoke + age_yrs + ________, data = fev)
summary(m3_age_height)
```

**Question:** Interpret the smoking coefficient in this model using the phrase "holding age and height constant."

Ans:

## Part 2d: Model 4, add sex

Finally, add sex. Sex may be related to body size and lung function, and it could also differ between smokers and non-smokers.

```{r model-4}
m4_age_height_sex <- lm(fev_l ~ smoke + age_yrs + height_in + ______, data = fev)
summary(m4_age_height_sex)
```

**Question:** Interpret the smoking coefficient in this model.

Ans:

# Part 3: Track effect sizes and model fit

## Part 3a: Build a comparison table

The helper code below extracts the smoking coefficient and model fit statistics from each model.

```{r model-comparison}
model_list <- list(
  "1. Smoking only" = m1_smoke,
  "2. + age" = m2_age,
  "3. + height" = m3_age_height,
  "4. + sex" = m4_age_height_sex
)

effect_table <- imap_dfr(
  model_list,
  ~ tidy(.x, conf.int = TRUE) %>%
    filter(term == "smokesmoker") %>%
    transmute(
      model = .y,
      smoking_beta = estimate,
      ci_low = conf.low,
      ci_high = conf.high,
      p_value = p.value
    )
)

fit_table <- imap_dfr(
  model_list,
  ~ glance(.x) %>%
    transmute(
      model = .y,
      r_squared = r.squared,
      adjusted_r_squared = adj.r.squared
    )
)

model_summary <- left_join(effect_table, fit_table, by = "model")
model_summary
```

**Question:** Across the four models, how does the estimated smoking effect change?

Ans:

**Question:** What happens to R-squared and adjusted R-squared as we add predictors?

Ans:

## Part 3b: Format the comparison table

```{r formatted-comparison}
model_summary %>%
  mutate(
    `Smoking beta (95% CI)` = paste0(
      format(round(smoking_beta, 3), nsmall = 3),
      " (",
      format(round(ci_low, 3), nsmall = 3),
      " to ",
      format(round(ci_high, 3), nsmall = 3),
      ")"
    ),
    p = if_else(p_value < 0.001, "<0.001", as.character(round(p_value, 3))),
    `R-squared` = round(r_squared, 3),
    `Adjusted R-squared` = round(adjusted_r_squared, 3)
  ) %>%
  select(model, `Smoking beta (95% CI)`, p, `R-squared`, `Adjusted R-squared`) %>%
  gt() %>%
  tab_header(
    title = "Estimated smoking effect on FEV",
    subtitle = "Outcome: FEV in litres"
  )
```

**Question:** Which model would you choose as the main estimate of the smoking effect? Why?

Ans:

# Part 4: Prediction from the adjusted model

Use Model 4 to predict FEV for two otherwise similar 12-year-old, 60-inch-tall males: one non-smoker and one smoker.

```{r predict-nonsmoker}
new_nonsmoker <- tibble(
  smoke = "__________",
  age_yrs = ______,
  height_in = ______,
  sex = "____"
)

predict(__________, newdata = __________)
```

```{r predict-smoker}
new_smoker <- tibble(
  smoke = "______",
  age_yrs = ______,
  height_in = ______,
  sex = "____"
)

predict(__________, newdata = __________)
```

**Question:** How far apart are the two predictions? How does that difference relate to the smoking coefficient in Model 4?

Ans:

# Part 5: Report the result

Write a short results paragraph. Include:

- the raw smoking estimate
- the fully adjusted smoking estimate
- R-squared or adjusted R-squared for the final model
- a cautious interpretation because this is observational data

Your answer:

# Wrap-up

In this workshop, multiple regression was used as an effect-size estimation tool. The smoking coefficient changed from a misleading positive raw association to a small negative adjusted association after accounting for age, height, and sex.
