---
title: "EXERCISE: Multiple linear regression with FEV"
subtitle: "Adjusted coefficients, prediction, and confounding"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In week 6, we used the FEV data to learn what regression is for: describing relationships, predicting outcomes, and controlling for another variable.

This week we return to the same dataset and focus properly on **multiple linear regression**. A multiple linear regression has more than one predictor:

```r
lm(outcome ~ predictor_1 + predictor_2, data = my_data)
```

The key interpretation is:

> Each coefficient is interpreted while holding the other predictors in the model constant.

The dataset contains 654 youths aged 3 to 19 from East Boston in the 1970s. The outcome is **forced expiratory volume** (`fev_l`), measured in litres. Higher values mean better lung function.

The variables we will use are:

- `fev_l`: forced expiratory volume in litres
- `age_yrs`: age in years
- `height_in`: height in inches
- `sex`: female or male
- `smoke`: current smoking status, `non_smoker` or `smoker`

# Load packages and data

Use `p_load()` to load `tidyverse`, `here`, `broom`, and `gt`.

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

Import the FEV data and inspect it.

```{r load-data}
fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint:** `fev` should have **654 rows and 5 columns**.

# Part 1: From simple regression to multiple regression

In simple linear regression, we use one predictor. In multiple linear regression, we add more predictors with `+`.

## Part 1a: The simple height model

Fit a simple linear regression predicting FEV from height.

```{r height-model}
height_model <- lm(fev_l ~ height_in, data = fev)
summary(height_model)
```

**Question:** Fill in the blanks.

The height coefficient is about ______. In the simple height model, this means that for each additional 1 inch of height, predicted FEV is higher by about ______ litres, on average.

**Question:** What is the R-squared of the simple height model?

Ans:

```{r height-r2}
summary(height_model)$r.squared
```

## Part 1b: Add age to the model

Now fit a multiple linear regression with both height and age as predictors.

```{r height-age-model}
height_age_model <- lm(________ ~ __________ + ________, data = fev)
summary(height_age_model)
```

**Question:** Fill in the blanks.

The height coefficient is about ______ in the height + age model. It means that, holding ______ constant, each additional 1 inch of height is associated with about ______ litres higher predicted FEV.

The age coefficient is about ______. It means that, holding ______ constant, each additional 1 year of age is associated with about ______ litres higher predicted FEV.

## Part 1c: Compare the height coefficient before and after adjustment

Compare the height coefficient in the simple model with the height coefficient in the height + age model.

```{r compare-height-coef}
tidy(height_model, conf.int = TRUE)
tidy(height_age_model, conf.int = TRUE)
```

**Question:** Did the height coefficient get larger, smaller, or stay about the same after adding age?

Ans:

**Question:** Why can a coefficient change when we add another predictor to the model?

Ans:

# Part 2: The smoking question

In week 6, the raw comparison suggested that smokers had higher mean FEV. That was suspicious because smokers in this dataset were older and taller on average.

Here we will build the comparison more carefully.

## Part 2a: Summarise smokers and non-smokers

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    n = n(),
    mean_age = mean(age_yrs),
    mean_height = mean(height_in),
    mean_fev = mean(fev_l),
    .groups = "drop"
  )
```

**Question:** Which group has higher mean FEV in the raw summary?

Ans:

**Question:** Which group is older and taller on average?

Ans:

## Part 2b: Fit the raw smoking model

Fit a simple model with smoking status as the only predictor.

```{r raw-smoke-model}
raw_smoke_model <- lm(________ ~ ________, data = fev)
summary(raw_smoke_model)
```

**Question:** The `smokesmoker` coefficient compares smokers with non-smokers. What is the coefficient, and what does it mean in the raw model?

Ans:

## Part 2c: Adjust for age and height

Now fit a multiple regression model with smoking status, age, and height.

```{r adjusted-smoke-model}
adjusted_smoke_model <- lm(
  fev_l ~ ________ + ________ + ________,
  data = fev
)

summary(adjusted_smoke_model)
```

**Question:** What is the `smokesmoker` coefficient after adjusting for age and height?

Ans:

**Question:** Interpret the adjusted smoking coefficient in one sentence. Use the phrase "holding age and height constant".

Ans:

**Question:** At alpha = 0.05, is the adjusted smoking coefficient statistically significant?

Ans:

## Part 2d: What changed?

```{r raw-vs-adjusted}
bind_rows(
  tidy(raw_smoke_model, conf.int = TRUE) %>% mutate(model = "Raw smoking model"),
  tidy(adjusted_smoke_model, conf.int = TRUE) %>% mutate(model = "Adjusted for age and height")
) %>%
  filter(term == "smokesmoker") %>%
  select(model, term, estimate, conf.low, conf.high, p.value)
```

**Question:** In one or two sentences, explain why the raw smoking model was misleading.

Ans:

# Part 3: Prediction from a multiple regression equation

Multiple regression can also be used for prediction. The adjusted smoking model has this form:

predicted FEV = intercept + height slope * height + age slope * age + smoking coefficient * smoking

For smoking, R uses an indicator variable:

- non-smoker: `smokesmoker = 0`
- smoker: `smokesmoker = 1`

## Part 3a: Predict with `predict()`

Use the adjusted model to predict FEV for a 12-year-old, 60-inch-tall non-smoker.

```{r predict-nonsmoker}
new_nonsmoker <- tibble(
  height_in = ______,
  age_yrs = ______,
  smoke = "__________"
)

predict(__________, newdata = __________)
```

Now predict FEV for a 12-year-old, 60-inch-tall smoker.

```{r predict-smoker}
new_smoker <- tibble(
  height_in = ______,
  age_yrs = ______,
  smoke = "______"
)

predict(__________, newdata = __________)
```

**Question:** How far apart are these two predictions? How does that difference relate to the smoking coefficient?

Ans:

## Part 3b: Build the prediction by hand

Extract the coefficients.

```{r coefficients}
coef(adjusted_smoke_model)
```

Use the coefficients to calculate the predicted FEV for a 12-year-old, 60-inch-tall smoker by hand.

```{r prediction-by-hand}
b <- coef(adjusted_smoke_model)

b["(Intercept)"] +
  b["height_in"] * ______ +
  b["age_yrs"] * ______ +
  b["smokesmoker"] * ______
```

**Question:** Does your hand calculation match `predict()`?

Ans:

# Part 4: Build a compact regression table

The `broom::tidy()` function turns regression output into a data frame.

```{r tidy-adjusted}
adjusted_results <- tidy(adjusted_smoke_model, conf.int = TRUE)
adjusted_results
```

Now format the coefficient estimates and confidence intervals.

```{r regression-table}
adjusted_results %>%
  mutate(
    `Beta (95% CI)` = paste0(
      format(round(estimate, 3), nsmall = 3),
      " (",
      format(round(conf.low, 3), nsmall = 3),
      " to ",
      format(round(conf.high, 3), nsmall = 3),
      ")"
    ),
    p = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))
  ) %>%
  select(term, `Beta (95% CI)`, p) %>%
  gt() %>%
  tab_header(
    title = "FEV multiple linear regression",
    subtitle = "Outcome: FEV in litres"
  )
```

**Question:** Choose one row from the table and write a one-sentence interpretation.

Ans:

# Part 5: Add one more predictor

Add `sex` to the model, after smoking, age, and height.

```{r sex-model}
sex_model <- lm(
  fev_l ~ smoke + age_yrs + height_in + ______,
  data = fev
)

summary(sex_model)
```

**Question:** What does the `sexmale` coefficient mean? Remember to mention the variables being held constant.

Ans:

**Question:** Did adding sex change the smoking coefficient much?

Ans:

```{r compare-adjusted-models}
bind_rows(
  tidy(adjusted_smoke_model, conf.int = TRUE) %>% mutate(model = "Age + height"),
  tidy(sex_model, conf.int = TRUE) %>% mutate(model = "Age + height + sex")
) %>%
  filter(term == "smokesmoker") %>%
  select(model, estimate, conf.low, conf.high, p.value)
```

# Wrap-up

In this workshop, you practiced:

1. fitting a multiple linear regression with `lm(y ~ x1 + x2 + x3)`;
2. interpreting coefficients as adjusted associations;
3. comparing raw and adjusted coefficients;
4. predicting from a multiple regression model;
5. using cautious language for observational data.

The most important phrase is: **holding the other variables constant**.
