---
title: "SOLUTIONS: Multiple linear regression with FEV"
subtitle: "Adjusted coefficients, prediction, and confounding"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

These are the completed solutions for the FEV version of workshop 8. Each blank exercise chunk is kept, then followed by a runnable solution chunk.

In week 6, we used the FEV data to learn what regression is for: describing relationships, predicting outcomes, and controlling for another variable.

This week we return to the same dataset and focus properly on **multiple linear regression**. A multiple linear regression has more than one predictor:

```r
lm(outcome ~ predictor_1 + predictor_2, data = my_data)
```

The key interpretation is:

> Each coefficient is interpreted while holding the other predictors in the model constant.

The dataset contains 654 youths aged 3 to 19 from East Boston in the 1970s. The outcome is **forced expiratory volume** (`fev_l`), measured in litres. Higher values mean better lung function.

# Load packages and data

*Exercise chunk (blank):*

```{r load-packages-q, eval=FALSE}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(________)
```

**Solution:**

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, here, broom, gt)
```

```{r load-data}
fev <- read_csv(here("data/fev_kahn.csv"))
glimpse(fev)
```

**Checkpoint answer:** `fev` has **654 rows and 5 columns**.

# Part 1: From simple regression to multiple regression

## Part 1a: The simple height model

```{r height-model}
height_model <- lm(fev_l ~ height_in, data = fev)
summary(height_model)
```

**Question:** Fill in the blanks.

**Answer:** The height coefficient is about **0.132**. In the simple height model, this means that for each additional 1 inch of height, predicted FEV is higher by about **0.132 litres**, on average.

```{r height-r2}
summary(height_model)$r.squared
```

**Answer:** The R-squared is about **0.754**, so height alone explains about **75.4%** of the variation in FEV.

## Part 1b: Add age to the model

*Exercise chunk (blank):*

```{r height-age-model-q, eval=FALSE}
height_age_model <- lm(________ ~ __________ + ________, data = fev)
summary(height_age_model)
```

**Solution:**

```{r height-age-model}
height_age_model <- lm(fev_l ~ height_in + age_yrs, data = fev)
summary(height_age_model)
```

**Answer:** The height coefficient is about **0.110** in the height + age model. It means that, holding **age** constant, each additional 1 inch of height is associated with about **0.110 litres** higher predicted FEV.

The age coefficient is about **0.065**. It means that, holding **height** constant, each additional 1 year of age is associated with about **0.065 litres** higher predicted FEV.

## Part 1c: Compare the height coefficient before and after adjustment

```{r compare-height-coef}
tidy(height_model, conf.int = TRUE)
tidy(height_age_model, conf.int = TRUE)
```

**Question:** Did the height coefficient get larger, smaller, or stay about the same after adding age?

**Answer:** It got smaller, from about 0.132 to about 0.110 litres per inch.

**Question:** Why can a coefficient change when we add another predictor to the model?

**Answer:** A coefficient can change when the new predictor is related to both the original predictor and the outcome. Here, age and height are related, and both are related to lung function, so the height-only slope partly mixed height with age.

# Part 2: The smoking question

## Part 2a: Summarise smokers and non-smokers

```{r smoke-summary}
fev %>%
  group_by(smoke) %>%
  summarise(
    n = n(),
    mean_age = mean(age_yrs),
    mean_height = mean(height_in),
    mean_fev = mean(fev_l),
    .groups = "drop"
  )
```

**Question:** Which group has higher mean FEV in the raw summary?

**Answer:** Smokers have higher mean FEV in the raw summary.

**Question:** Which group is older and taller on average?

**Answer:** Smokers are older and taller on average.

## Part 2b: Fit the raw smoking model

*Exercise chunk (blank):*

```{r raw-smoke-model-q, eval=FALSE}
raw_smoke_model <- lm(________ ~ ________, data = fev)
summary(raw_smoke_model)
```

**Solution:**

```{r raw-smoke-model}
raw_smoke_model <- lm(fev_l ~ smoke, data = fev)
summary(raw_smoke_model)
```

**Answer:** The `smokesmoker` coefficient is about **0.711**. In the raw model, smokers have predicted FEV about 0.711 litres higher than non-smokers, on average.

## Part 2c: Adjust for age and height

*Exercise chunk (blank):*

```{r adjusted-smoke-model-q, eval=FALSE}
adjusted_smoke_model <- lm(
  fev_l ~ ________ + ________ + ________,
  data = fev
)

summary(adjusted_smoke_model)
```

**Solution:**

```{r adjusted-smoke-model}
adjusted_smoke_model <- lm(
  fev_l ~ smoke + age_yrs + height_in,
  data = fev
)

summary(adjusted_smoke_model)
```

**Answer:** The adjusted `smokesmoker` coefficient is about **-0.110**.

**Answer:** Holding age and height constant, smokers have predicted FEV about **0.110 litres lower** than non-smokers, on average.

**Answer:** No. The p-value is about **0.067**, so at alpha = 0.05 we fail to reject the null hypothesis for the adjusted smoking coefficient.

## Part 2d: What changed?

```{r raw-vs-adjusted}
bind_rows(
  tidy(raw_smoke_model, conf.int = TRUE) %>% mutate(model = "Raw smoking model"),
  tidy(adjusted_smoke_model, conf.int = TRUE) %>% mutate(model = "Adjusted for age and height")
) %>%
  filter(term == "smokesmoker") %>%
  select(model, term, estimate, conf.low, conf.high, p.value)
```

**Answer:** The raw smoking model was misleading because smokers were much older and taller than non-smokers. Older and taller youths naturally have higher FEV, so the raw comparison mixed smoking status with age and body size.

# Part 3: Prediction from a multiple regression equation

## Part 3a: Predict with `predict()`

*Exercise chunk (blank):*

```{r predict-nonsmoker-q, eval=FALSE}
new_nonsmoker <- tibble(
  height_in = ______,
  age_yrs = ______,
  smoke = "__________"
)

predict(__________, newdata = __________)
```

**Solution:**

```{r predict-nonsmoker}
new_nonsmoker <- tibble(
  height_in = 60,
  age_yrs = 12,
  smoke = "non_smoker"
)

predict(adjusted_smoke_model, newdata = new_nonsmoker)
```

*Exercise chunk (blank):*

```{r predict-smoker-q, eval=FALSE}
new_smoker <- tibble(
  height_in = ______,
  age_yrs = ______,
  smoke = "______"
)

predict(__________, newdata = __________)
```

**Solution:**

```{r predict-smoker}
new_smoker <- tibble(
  height_in = 60,
  age_yrs = 12,
  smoke = "smoker"
)

predict(adjusted_smoke_model, newdata = new_smoker)
```

**Answer:** The smoker prediction is about **0.110 litres lower** than the non-smoker prediction. That gap is exactly the adjusted smoking coefficient, because the two people have the same age and height.

## Part 3b: Build the prediction by hand

```{r coefficients}
coef(adjusted_smoke_model)
```

*Exercise chunk (blank):*

```{r prediction-by-hand-q, eval=FALSE}
b <- coef(adjusted_smoke_model)

b["(Intercept)"] +
  b["height_in"] * ______ +
  b["age_yrs"] * ______ +
  b["smokesmoker"] * ______
```

**Solution:**

```{r prediction-by-hand}
b <- coef(adjusted_smoke_model)

b["(Intercept)"] +
  b["height_in"] * 60 +
  b["age_yrs"] * 12 +
  b["smokesmoker"] * 1
```

**Answer:** Yes. It matches the `predict()` result for the 12-year-old, 60-inch-tall smoker.

# Part 4: Build a compact regression table

```{r tidy-adjusted}
adjusted_results <- tidy(adjusted_smoke_model, conf.int = TRUE)
adjusted_results
```

```{r regression-table}
adjusted_results %>%
  mutate(
    `Beta (95% CI)` = paste0(
      format(round(estimate, 3), nsmall = 3),
      " (",
      format(round(conf.low, 3), nsmall = 3),
      " to ",
      format(round(conf.high, 3), nsmall = 3),
      ")"
    ),
    p = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))
  ) %>%
  select(term, `Beta (95% CI)`, p) %>%
  gt() %>%
  tab_header(
    title = "FEV multiple linear regression",
    subtitle = "Outcome: FEV in litres"
  )
```

**Question:** Choose one row from the table and write a one-sentence interpretation.

**Answer:** Example: Holding smoking status and age constant, each additional inch of height is associated with about **0.109 litres higher** predicted FEV.

# Part 5: Add one more predictor

*Exercise chunk (blank):*

```{r sex-model-q, eval=FALSE}
sex_model <- lm(
  fev_l ~ smoke + age_yrs + height_in + ______,
  data = fev
)

summary(sex_model)
```

**Solution:**

```{r sex-model}
sex_model <- lm(
  fev_l ~ smoke + age_yrs + height_in + sex,
  data = fev
)

summary(sex_model)
```

**Answer:** Holding smoking status, age, and height constant, males have predicted FEV about **0.157 litres higher** than females, on average.

**Answer:** Adding sex changes the smoking coefficient from about **-0.110** to about **-0.087**, so the smoking estimate moves slightly toward zero.

```{r compare-adjusted-models}
bind_rows(
  tidy(adjusted_smoke_model, conf.int = TRUE) %>% mutate(model = "Age + height"),
  tidy(sex_model, conf.int = TRUE) %>% mutate(model = "Age + height + sex")
) %>%
  filter(term == "smokesmoker") %>%
  select(model, estimate, conf.low, conf.high, p.value)
```

# Wrap-up

In this workshop, you practiced fitting multiple regression models, interpreting adjusted coefficients, comparing raw and adjusted coefficients, predicting from a multiple regression model, and using cautious language for observational data.

The most important phrase is: **holding the other variables constant**.
