---
title: "REFERENCE: Building a publication-style regression table from scratch"
subtitle: "Raw numbers, odds ratios and a forest plot in one `gt` table"
author: "GRAPH Courses team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this script is for

This is **not** an exercise. It is a worked reference you can copy from when you need to build a results table like the ones you see in journals: raw numerators and denominators, an observed-percentage bar, the odds ratio with its confidence interval, and a small forest plot, all in one table.

The point of building it from scratch is that you can then adapt every piece. In particular, this example shows how to handle a **continuous** predictor, which most template code gets wrong: the odds ratio comes from the continuous model, but the raw-number rows still have to be shown in bands, because "number of patients at age 23.4" is not a thing.

We use the `birthwt` data from the `MASS` package: 189 births at Baystate Medical Center, Springfield, Massachusetts, in 1986. The outcome is **low birth weight** (under 2500 g).

```{r load-packages}
if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, broom, gt, glue, scales, MASS)
```

Note that `MASS` also has a `select()` function, which masks the `dplyr` one. Below we write `dplyr::select()` where we need it.

# PART 1: Prepare the data

```{r prepare-data}
data(birthwt, package = "MASS")

births <- birthwt %>%
  transmute(
    low_bw = factor(low, levels = c(0, 1), labels = c("Normal", "Low")),
    age_years = age,
    weight_lb = lwt,
    race = factor(race, levels = 1:3, labels = c("White", "Black", "Other")),
    smoking = factor(smoke, levels = c(0, 1), labels = c("No", "Yes")),
    prior_ptl = factor(pmin(ptl, 2), levels = 0:2, labels = c("0", "1", "2+"))
  )

glimpse(births)
```

`ptl` is the count of previous premature labours. We collapse everything above 1 into a `2+` category, because the higher counts are rare.

Our five predictors cover the three shapes you will meet in practice:

- **Continuous:** `age_years`, `weight_lb`
- **Binary:** `smoking`
- **Polytomous:** `race`, `prior_ptl`

```{r outcome-check}
births %>% count(low_bw)
```

# PART 2: Fit the models

## Continuous predictors, rescaled

A one-year change in age or a one-pound change in weight is a very small step, so the odds ratio comes out near 1 and is hard to read. Journals usually report a more meaningful unit: per 5 years, per 10 pounds.

The easiest way to do this is to divide the predictor by the step size before fitting. The coefficient then refers to that step.

```{r rescale}
births <- births %>%
  mutate(age_per_5y = age_years / 5, weight_per_10lb = weight_lb / 10)
```

## Fit one univariable model per predictor

```{r univariable-models}
model_age <- glm(low_bw ~ age_per_5y, family = binomial, data = births)
model_weight <- glm(low_bw ~ weight_per_10lb, family = binomial, data = births)
model_race <- glm(low_bw ~ race, family = binomial, data = births)
model_smoking <- glm(low_bw ~ smoking, family = binomial, data = births)
model_ptl <- glm(low_bw ~ prior_ptl, family = binomial, data = births)
```

## Fit the multivariable model

```{r multivariable-model}
model_multi <- glm(
  low_bw ~ age_per_5y + weight_per_10lb + race + smoking + prior_ptl,
  family = binomial, data = births
)

summary(model_multi)
```

## Collect all the odds ratios in one table

`tidy()` gives us one row per coefficient. We stack the five univariable models, drop the intercepts, and keep the columns we need.

```{r collect-univariable}
univariable_ors <- bind_rows(
  tidy(model_age, conf.int = TRUE, exponentiate = TRUE),
  tidy(model_weight, conf.int = TRUE, exponentiate = TRUE),
  tidy(model_race, conf.int = TRUE, exponentiate = TRUE),
  tidy(model_smoking, conf.int = TRUE, exponentiate = TRUE),
  tidy(model_ptl, conf.int = TRUE, exponentiate = TRUE)
) %>%
  filter(term != "(Intercept)") %>%
  dplyr::select(term, uni_or = estimate, uni_low = conf.low,
                uni_high = conf.high, uni_p = p.value)

univariable_ors
```

```{r collect-multivariable}
multivariable_ors <- tidy(model_multi, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  dplyr::select(term, multi_or = estimate, multi_low = conf.low,
                multi_high = conf.high, multi_p = p.value)

multivariable_ors
```

# PART 3: Build the raw-number rows

This is the part that template code usually skips. Every row of the final table needs a denominator, a numerator and an observed percentage.

## A helper for categorical predictors

For a factor, each level is one row. `term` is built the way R names coefficients, so that we can join to the model output later: the variable name pasted onto the level, for example `smokingYes`.

```{r categorical-rows-helper}
categorical_rows <- function(data, variable) {
  data %>%
    group_by(level = .data[[variable]]) %>%
    summarise(n_total = n(), n_events = sum(low_bw == "Low")) %>%
    mutate(
      variable = variable,
      term = paste0(variable, level),
      percent_events = 100 * n_events / n_total,
      row_type = "estimate"
    )
}

categorical_rows(births, "smoking")
```

The reference level (`smokingNo`) has no matching coefficient, so its odds ratio columns will be empty after the join. That is exactly what we want: the table should show the reference group's raw numbers and the word "Reference".

## A helper for continuous predictors

A continuous predictor produces **two kinds of row**:

1. One **estimate row** carrying the odds ratio per unit step. Its raw-number cell shows the mean and SD rather than a percentage, because there is no single group to count.
2. Several **band rows** showing the observed percentage in each band. These carry no odds ratio; they exist so the reader can see the raw pattern.

```{r continuous-rows-helper}
continuous_rows <- function(data, variable, breaks, labels) {
  estimate_row <- data %>%
    summarise(
      n_total = n(), n_events = sum(low_bw == "Low"),
      mean_value = mean(.data[[variable]]), sd_value = sd(.data[[variable]])
    ) %>%
    mutate(variable = variable, level = "Continuous", row_type = "estimate")

  band_rows <- data %>%
    mutate(band = cut(.data[[variable]], breaks = breaks, labels = labels,
                      include.lowest = TRUE)) %>%
    group_by(level = band) %>%
    summarise(n_total = n(), n_events = sum(low_bw == "Low")) %>%
    mutate(
      variable = variable,
      percent_events = 100 * n_events / n_total,
      row_type = "band"
    )

  bind_rows(estimate_row, band_rows)
}

continuous_rows(births, "age_years", c(14, 19, 24, 29, Inf),
                c("14-19", "20-24", "25-29", "30+"))
```

The estimate row has no `term` yet, because the model coefficient is named after the *rescaled* variable. We attach the right term when we assemble the table.

## Assemble the raw-number rows for all five predictors

```{r all-raw-rows}
raw_rows <- bind_rows(
  continuous_rows(births, "age_years", c(14, 19, 24, 29, Inf),
                  c("14-19", "20-24", "25-29", "30+")) %>%
    mutate(term = if_else(row_type == "estimate", "age_per_5y", NA_character_),
           variable = "Age, years"),
  continuous_rows(births, "weight_lb", c(80, 110, 130, 150, Inf),
                  c("80-110", "111-130", "131-150", "151+")) %>%
    mutate(term = if_else(row_type == "estimate", "weight_per_10lb", NA_character_),
           variable = "Weight at last period, lb"),
  categorical_rows(births, "race") %>% mutate(variable = "Race"),
  categorical_rows(births, "smoking") %>% mutate(variable = "Smoking during pregnancy"),
  categorical_rows(births, "prior_ptl") %>% mutate(variable = "Previous premature labours")
) %>%
  mutate(variable = recode(variable,
    race = "Race",
    smoking = "Smoking during pregnancy",
    prior_ptl = "Previous premature labours"
  ))

raw_rows %>% dplyr::select(variable, level, row_type, n_total, n_events, percent_events)
```

# PART 4: Join the raw numbers to the model results

```{r join-results}
results <- raw_rows %>%
  left_join(univariable_ors, by = "term") %>%
  left_join(multivariable_ors, by = "term") %>%
  mutate(
    variable = factor(variable, levels = c(
      "Age, years", "Weight at last period, lb", "Race",
      "Smoking during pregnancy", "Previous premature labours"
    )),
    is_reference = row_type == "estimate" & is.na(uni_or)
  ) %>%
  arrange(variable)

results %>% dplyr::select(variable, level, row_type, uni_or, multi_or, is_reference)
```

Now every display row exists exactly once, and the three sources of information — raw counts, univariable OR, multivariable OR — sit side by side.

# PART 5: Format the cells

We build the text of each cell before handing anything to `gt`. Doing the formatting in `dplyr` keeps the `gt` call short and makes each decision easy to inspect.

```{r label-column}
results_formatted <- results %>%
  mutate(
    row_label = case_when(
      row_type == "band" ~ paste0("    ", level),
      level == "Continuous" & variable == "Age, years" ~ "Continuous, per 5 years",
      level == "Continuous" ~ "Continuous, per 10 lb",
      TRUE ~ as.character(level)
    )
  )

results_formatted %>% dplyr::select(variable, row_label, row_type)
```

Band rows are indented with non-breaking spaces so that they read as sub-rows of the estimate row above them.

Next, the numeric columns. Continuous estimate rows show mean (SD) instead of a percentage; reference rows show the word "Reference" instead of an odds ratio.

```{r format-cells}
or_text <- function(or, low, high) {
  sprintf("%.2f (%.2f – %.2f)", or, low, high)
}

results_formatted <- results_formatted %>%
  mutate(
    n_events_display = if_else(level == "Continuous", "", as.character(n_events)),
    uni_text = case_when(
      is_reference ~ "1 (reference)",
      row_type == "band" ~ "–",
      TRUE ~ or_text(uni_or, uni_low, uni_high)
    ),
    multi_text = case_when(
      is_reference ~ "1 (reference)",
      row_type == "band" ~ "–",
      TRUE ~ or_text(multi_or, multi_low, multi_high)
    ),
    p_text = if_else(is.na(multi_p), "", pvalue(multi_p, accuracy = 0.001))
  )

results_formatted %>% dplyr::select(row_label, uni_text, multi_text, p_text)
```

## A bar for the observed percentage

There is no need for a plotting package here. A coloured `div` of the right width, written as raw HTML, is enough. We scale the widest bar to 90 pixels.

The continuous estimate rows have no percentage to show, so they get the mean and SD of the predictor instead.

```{r percent-bar}
max_percent <- max(results_formatted$percent_events, na.rm = TRUE)

percent_bar <- function(percent) {
  if (is.na(percent)) return("")
  width <- 90 * percent / max_percent
  glue('<div style="display:flex;align-items:center;gap:6px;">
         <div style="background:#aab6e8;height:13px;width:{round(width)}px;"></div>
         <span style="font-size:12px;">{sprintf("%.1f", percent)}</span>
       </div>')
}

results_formatted <- results_formatted %>%
  mutate(
    percent_html = map_chr(percent_events, percent_bar),
    percent_html = if_else(
      level == "Continuous",
      sprintf("<i style='color:#5d7475;'>mean %.1f (SD %.1f)</i>", mean_value, sd_value),
      percent_html
    )
  )
```

## A forest plot cell for each odds ratio

Each cell is one tiny `ggplot`. The critical detail is that **every cell must use the same x-axis limits**, otherwise the reader compares points that are drawn on different scales. We fix the limits once, on a log scale, and clip any interval that runs past the edge.

The point and line sizes look absurdly large below. That is deliberate: the plot is drawn at normal figure size and then shrunk to the height of a table row, so ordinary sizes disappear.

```{r forest-cell}
or_limits <- c(0.2, 20)

forest_cell <- function(or, low, high) {
  if (is.na(or)) {
    plot_data <- tibble(or = NA_real_, low = NA_real_, high = NA_real_)
  } else {
    plot_data <- tibble(or = or,
                        low = pmax(low, or_limits[1]),
                        high = pmin(high, or_limits[2]))
  }

  ggplot(plot_data, aes(x = or, y = 1)) +
    annotate("rect", xmin = or_limits[1], xmax = or_limits[2],
             ymin = 0.5, ymax = 1.5, fill = "grey96") +
    geom_vline(xintercept = 1, linetype = "dashed",
               colour = "grey45", linewidth = 2.5) +
    geom_segment(aes(x = low, xend = high, y = 1, yend = 1),
                 linewidth = 2.5, na.rm = TRUE) +
    geom_point(size = 14, na.rm = TRUE) +
    scale_x_continuous(transform = "log", limits = or_limits, expand = c(0, 0)) +
    scale_y_continuous(limits = c(0.5, 1.5), expand = c(0, 0)) +
    theme_void()
}

forest_cell(2.02, 1.08, 3.80)
```

`gt::ggplot_image()` turns a plot object into an inline image that a table cell can hold. We build one image per row.

```{r forest-column}
results_formatted <- results_formatted %>%
  mutate(
    uni_plot = pmap_chr(list(uni_or, uni_low, uni_high), function(o, l, h) {
      ggplot_image(forest_cell(o, l, h), height = 26, aspect_ratio = 4.5)
    }),
    multi_plot = pmap_chr(list(multi_or, multi_low, multi_high), function(o, l, h) {
      ggplot_image(forest_cell(o, l, h), height = 26, aspect_ratio = 4.5)
    })
  )
```

# PART 6: Draw the table

Everything is now a string, so the `gt` call only has to arrange and label.

```{r final-table}
final_table <- results_formatted %>%
  dplyr::select(variable, row_label, n_total, n_events_display, percent_html,
                uni_text, uni_plot, multi_text, multi_plot, p_text) %>%
  gt(groupname_col = "variable") %>%
  fmt_markdown(columns = c(percent_html, uni_plot, multi_plot)) %>%
  cols_label(
    row_label = "",
    n_total = "Num births",
    n_events_display = "Num low birth weight",
    percent_html = "% low birth weight",
    uni_text = "Univariable OR (95% CI)",
    uni_plot = "",
    multi_text = "Multivariable OR (95% CI)",
    multi_plot = "",
    p_text = "p-value"
  ) %>%
  cols_align(align = "left", columns = c(row_label, percent_html)) %>%
  cols_align(align = "center", columns = c(n_total, n_events_display,
                                           uni_text, multi_text, p_text)) %>%
  tab_style(style = cell_text(weight = "bold"), locations = cells_row_groups()) %>%
  tab_style(
    style = cell_text(style = "italic", color = "#5d7475"),
    locations = cells_body(columns = row_label,
                           rows = results_formatted$row_type == "band")
  ) %>%
  tab_style(
    style = cell_fill(color = "#eef1fb"),
    locations = cells_body(rows = results_formatted$level == "Continuous")
  ) %>%
  tab_header(
    title = md("**Factors associated with low birth weight**"),
    subtitle = "Baystate Medical Center, 1986 (n = 189)"
  ) %>%
  tab_footnote(
    footnote = "Age and weight were modelled as continuous linear terms. Bands are shown for description only and have no odds ratio.",
    locations = cells_column_labels(columns = row_label)
  ) %>%
  tab_options(table.font.size = px(13), data_row.padding = px(4))

final_table
```

# PART 7: What to take from this

- **Raw numbers and model estimates come from different places.** Build them separately, then join on a key. Here the key is `term`, the name R gives each coefficient.
- **A continuous predictor needs two row types.** One row carries the odds ratio per unit step; the descriptive band rows carry percentages and no odds ratio. Say so in a footnote, so nobody reads the bands as separate estimates.
- **Rescale continuous predictors** to a step that a reader can picture. Per 5 years is easier to hold in mind than per year.
- **Reference levels have no coefficient.** After the join they are `NA`, which is your cue to print "1 (reference)".
- **Fix the forest plot x-axis once** and clip anything that runs off the edge. Per-cell automatic scales silently mislead.
- **Do the formatting in `dplyr`, not in `gt`.** Then you can print the intermediate table and check every value.

Export it when you need to hand it over:

```{r export, eval=FALSE}
gtsave(final_table, "birthwt_regression_table.html")
```

# Reference

Venables WN, Ripley BD (2002). *Modern Applied Statistics with S*, 4th edition. Springer. The `birthwt` data originally come from Hosmer DW, Lemeshow S (1989), *Applied Logistic Regression*.
