---
title: "Example Report on Motorcycle Accident Deaths in Colombia"
subtitle: "Tidyverse and RMarkdown Practice"
author: "GRAPH Courses Team"
output:
  prettydoc::html_pretty:
    theme: architect
---

```{r include = FALSE}
# Set chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
# TASK 1: LOAD THE REQUIRED PACKAGES
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse,
               here,
               janitor, 
               reactable)
```

```{r}
# TASK 2: IMPORT THE DATA
col_accidents <- read_csv(here("data/motorcycle_accidents_colombia.csv"))
#head(col_accidents)
```

```{r}
# TASK 3: CLEAN THE COLUMN NAMES 
col_accidents_clean <- col_accidents %>% 
  clean_names()

#head(col_accidents_clean)
```

```{r}
# TASK 4: RENAME AND TRANSFORM VARIABLES
col_accidents_clean2 <- col_accidents_clean %>%
  rename(position = condition) %>% 
  mutate(gender = str_to_title(gender))
#head(col_accidents_clean2)
```

```{r}
# TASK 5: CREATE A SUBSET
col_accidents_subset <- col_accidents_clean2 %>% 
  select(id, gender, age, age_group, position) %>% 
  filter(position != "unknown")
```

```{r}
# TASK 6: SUMMARIZE DATA
age_sex_count <- col_accidents_subset %>% 
  count(age_group, gender) %>% 
  ungroup()
```

```{r include = F}
# TASK 7: VISUALIZE SUMMARY DATA
death_plot <- 
  ggplot(age_sex_count) +
  aes(x = age_group, y = n, fill = gender) +
  geom_col() +
  #scale_fill_hue(direction = 1) +
  theme_minimal()+
  labs(x= "Age Group", 
       y = "Number of deaths", 
       title = "Number of deaths by Age Groups", 
       fill = "Gender", 
       subtitle = "The graph shows the number of deaths for each age group by gender",
       caption = "Data source: Romero & Porras (2021)") +
  # OPTIONAL: ADD TEXT LABELS
  geom_text(aes(label = n),
            vjust = 1.1,
            size = 3,
            position = "stack",
            color = "white", 
            fontface = "bold")

death_plot
```

```{r eval = F}
# TASK 8: EXPORT YOUR OUTPUTS

write_csv(x = age_sex_count, file = here("outputs/age_sex_count.csv"))

ggsave(plot = death_plot, filename = here("outputs/death_plot.png"))
```

# Introduction

The dataset analysed in this report provides information about **`r nrow(col_accidents_subset)`** deaths caused by motorcycle accidents in Medellín, a Colombian city, from 2012 to 2015. The data is sourced from medical and police certificates and can be accessed [here](https://zenodo.org/record/4836304).

# Age-sex distribution of motorcylce deaths

In this analysis, we looked at the number of deaths per age and sex group.

```{r fig.height=7}
# Print your graph or table here.
death_plot
```

```{r}
# TASK 9: ADD INLINE CODE AND COMMENTARY 

## Each of the 3 statistics in this code chunk will be quoted in the text below.
max_sex_group <- age_sex_count %>% 
  filter(n == max(n)) %>% 
  pull(gender)

max_age_group <- age_sex_count %>% 
  filter(n == max(n)) %>% 
  pull(age_group)

max_n_deaths <- max(age_sex_count$n) 
# OR age_sex_count$n %>% max()
# OR age_sex_count %>% filter(n == max(n)) %>% pull(n)
```

The age-sex group with the most fatal accidents was **`r paste0(max_sex_group, "s")`** aged **`r max_age_group`**, with **`r max_n_deaths`** deaths. 

## Analysis of Driver Position 

```{r}
# TASK 10: SUMMARIZE DRIVER POSITION DATA 

# Create subset for age group and sex with most deaths
driver_position_subset <- col_accidents_subset %>% 
  filter(gender == "Male" & age_group == "20-24") %>% 
  count(position) %>% 
  ungroup()
```

We further analyzed the position of the driver involved in these accidents for the age-sex group with the most fatal accidents. 

The table below shows the breakdown of accidents by driver position:

```{r}
# Paste your table here
reactable::reactable(driver_position_subset,
                     columns = list(
                       position = colDef(name = "Driver Position"),
                       n = colDef(name = "Deaths")
  ))
```
