---
title: "Report on COVID-19 Patient Outcomes in New York City Hospitals"
subtitle: "Tidyverse and RMarkdown Practice"
author: "YOUR NAME HERE" # Remember to add your name!
output:
  prettydoc::html_pretty:
    theme: architect
date: "`r Sys.Date()`"
---

```{r include = FALSE}
# NOTE
# Please check the "rsr_week_01_instructions.Rmd" document as you complete this assignment! We encourage you to have both the instructions document and this assignment document open.
```

```{r include = FALSE}
# Set chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
# TASK 1: LOAD THE REQUIRED PACKAGES
if(!require(pacman)) install.packages("pacman")
"WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 2: IMPORT THE DATA & INSPECT VARIABLES
covid_us <- "WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 3: CLEAN THE COLUMN NAMES 
covid_us_clean <- covid_us %>% 
  "WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 4: RENAME AND TRANSFORM VARIABLES
covid_us_clean2 <- covid_us_clean %>%
  "WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 5: CREATE A SUBSET
covid_us_subset <- "WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 6: SUMMARIZE DATA
race_sex_count <- "WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 7: VISUALIZE SUMMARY DATA
"WRITE_YOUR_CODE_HERE"
```

```{r}
# TASK 8: EXPORT YOUR OUTPUTS
"WRITE_YOUR_CODE_HERE"
```

# Introduction

The dataset analysed in this report contains records of **`r nrow(covid_us_subset)`** COVID-19 deaths among patients admitted to two New York City hospitals (SUNY Downstate and Maimonides Medical Center) during the early waves of the pandemic. The records include patient demographics, comorbidities, and admission details.

# Race-sex distribution of COVID-19 deaths

In this analysis, we looked at the number of deaths per race and sex group.

```{r}
# Print your graph here.

```

```{r}
# TASK 9: ADD INLINE CODE AND COMMENTARY 

## Each of the 3 statistics in this code chunk will be quoted in the text below.

max_sex_group <- race_sex_count %>% 
  filter(n == max(n)) %>% 
  pull(sex)

max_race_group <- race_sex_count %>% 
  "WRITE_YOUR_CODE_HERE"

max_n_deaths <- "WRITE_YOUR_CODE_HERE"
```

The demographic group with the highest number of COVID-19 deaths was `"WRITE_YOUR_INLINE_CODE_HERE"` `r paste0(max_sex_group, "s")`, with `"WRITE_YOUR_INLINE_CODE_HERE"` deaths.

## Demographic Analysis of Expired Male Patients

We further analyzed the racial demographics of male patients who expired from the disease at Maimonides Medical Center.

The table below shows this demographic breakdown:

```{r}
# TASK 10: SUMMARIZE DEMOGRAPHIC DATA 
"WRITE_YOUR_CODE_HERE"
```
